### 2022-06-13 manuscript.Rmd
adding template and outline for manuscript

---
title: "'You Draw It' Implementation of visually fitted trends with `r2d3`"

# to produce blinded version set to 1
blinded: 0

authors: 
- name: Emily A. Robinson 1
  # thanks: The authors gratefully acknowledge ...
  affiliation: Department of Statistics, University of Nebraska-Lincoln

- name: Reka Howard 2
  affiliation: Department of Statistics, University of Nebraska-Lincoln

- name: Susan VanderPlas 3
  affiliation: Department of Statistics, University of Nebraska-Lincoln

keywords:
- graphics
- user interaction
- regression

abstract: |
  How do statistical regression results compare to intuitive, visually fitted results? Fitting lines by eye through a set of points has been explored since the 20th century. Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. In 2015, the New York Times introduced an interactive feature, called 'You Draw It,' where readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality. This research is intended to implement 'You Draw It', adapted from the New York Times, as a way to measure the patterns we see in data. *(116/200 words)*

bibliography: bibliography.bib
output: rticles::asa_article
header-includes:
   - \usepackage[dvipsnames]{xcolor} % colors
   - \newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
   - \newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
   - \newcommand{\rh}[1]{{\textcolor{Green}{#1}}}
   - \usepackage[capitalise]{cleveref}
   - \newcommand\pcref[1]{(\cref{#1})}
---

# Introduction

Our visual system is naturally built to look for structure and identify patterns. For instance, points going down from left to right indicates a negative correlation between the $x$ and $y$ variables. How can we compare our intuitive visual sense of patterns to those determined by statistical methods?

## Testing Statistical Graphics

Graphical tests are useful for studying the perception of statistical graphs. Studies might ask participants to identify differences in graphs, read information off of a chart accurately, use data to make correct real-world decisions, or predict the next few observations. All of these types of tests require different levels of use and manipulation of the information being presented in the chart.

Efforts in the field of statistical graphics have developed graphical testing tools and methods. Through experimentation, graphical testing methods allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation. The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics.

## Fitting Trends by Eye

Initial studies in the 20th century explored the use of fitting lines by eye through a set of points [@finney1951subjective; @mosteller1981eye]. Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.

In 2015, the New York Times introduced an interactive feature, called 'You Draw It' [@aisch2015you], where readers input their own assumptions about various metrics and compare how these assumptions relate to reality. The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics through the use of drawing a line on their computer screen with their mouse.

## Research Objectives

The goal of this research is to implement ‘You Draw It’, adapted from the New York Times feature, as a way to measure the patterns we see in data. Here, we provide technical details of the software development, utilizing interactive graphics in R. We then share results from our study which validates 'You Draw It' as a method for graphical testing and apply an appropriate data analysis method to the participant data.

# Development

## 'You Draw It' Task

Users are shown an interactive scatter-plot \pcref{fig:you-draw-it-example} along with the prompt, "Use your mouse to fill in the trend in the yellow box region." The yellow box region moves along as the user draws their trend-line, providing a visual cue which indicates where the user still needs to complete a trend line. After the entire domain has been visually estimated or predicted, the yellow region disappears, indicating the participant has completed the task. Visit [bit.ly/3zHnlgp](https://bit.ly/3zHnlgp}) for a test applet.

## Source Code

Data Driven Documents (D3), a JavaScript-based graphing framework that facilitates user interaction, is used to create the 'You Draw It' visual. A challenge of working with D3 is the environment necessary to display the graphics and images. The `r2d3` package [@r2d3pkg] provides an efficient integration of D3 visuals into R HTML formats. We integrate the D3 visual source code into an R shiny [@shinypkg] application in order to allow for user interaction and data collection.

We conduct all data simulation and processing in R and output two data sets - *point data* and *line data* - containing ($x$, $y$) coordinates corresponding to either a simulated point or fitted value predicted by a statistical model respectively. Then, the `r2d3` package converts the data sets in R to JSON to be interpreted by the D3.js code. We define functions in D3.js to draw the initial plot and set up drawable points for the user drawn line. Drag events in D3.js are utilized to react to observe and react to user input. Shiny Messages are used to communicate the user interaction between the D3 code and the R environment. The plot is then rendered and updated on user interaction into the R shiny application with the `RenderD3` and `d3Output` functions.

Parameters for aesthetic design choices are defined in a list of options and `r2d3` passes these to the D3.js code. For instance, we can specify the buffer space allowed for the $x$ and $y$ axes to avoid users to anchor their lines to the axes limits.

For `D3.js` source code, visit [bit.ly/3teA3lM](https://bit.ly/3teA3lM).

```{r, you-draw-it-example, echo = F, out.width = "100%", fig.align='center', fig.cap = "'You Draw It' interaction task plot as shown to user. \\textbf{Left:} illustrates what user first sees with the prompt \\textit{`Use your mouse to fill in the trend in the yellow box region.'} \\textbf{Middle:} illustrates what the user sees while completing the task. \\textbf{Right:} illustrates the users finished trend line."}
knitr::include_graphics("images/ydi-stimuli.png")
```

# Application

## Tool Validation

We conducted a study in order to validate 'You Draw It' as a method for graphical testing, comparing results to the less technological method utilized in @mosteller1981eye. Results from our study were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the least-squares regression line. This trend was most prominent when shown data simulated with larger variances. This study reinforces the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.

## Data Analysis

Feedback data from conducted studies are collected and stored in a database for analysis. Within the collected feedback data, we know the simulated data points, the predicted values from the statistical model, and the predicted values from the user drawn line. In our initial studies, a unique data set was simulated independently for each participant. Therefore, we evaluate the accuracy of the user drawn line by observing the deviation, vertical residuals, between the user drawn line and the predicted values from the statistical model. We use a Generalized Additive Mixed Model (GAMM) to model the vertical residuals in order to statistically compare visually fitted trends to actual metrics, simulated data models, or statistical regression results. A benefit of using a GAMM is the estimation of smoothing splines, allowing for flexibility in the residual trend.

# Future Work

In this work, we implemented and validated ‘You Draw It’ as a way to measure the patterns we see in data. We demonstrated the use of generalized additive models to statistically model participant data. While technical details of the development process are presented here, we intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers. Further investigation is necessary to implement this method real data in order to facilitate scientific communication.

### 2022-06-13 manuscript.tex
adding template and outline for manuscript

% !TeX program = pdfLaTeX
\documentclass[12pt]{article}
\usepackage{amsmath}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{textcomp}
\usepackage[hyphens]{url} % not crucial - just used below for the URL
\usepackage{hyperref}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

%% load any required packages here

% tightlist command for lists without linebreak
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage[dvipsnames]{xcolor} % colors
\newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
\newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
\newcommand{\rh}[1]{{\textcolor{Green}{#1}}}
\usepackage[capitalise]{cleveref}
\newcommand\pcref[1]{(\cref{#1})}

\begin{document}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf `You Draw It' Implementation of visually fitted trends with
\texttt{r2d3}}

  \author{
        Emily A. Robinson 1 \\
    Department of Statistics, University of Nebraska-Lincoln\\
     and \\     Reka Howard 2 \\
    Department of Statistics, University of Nebraska-Lincoln\\
     and \\     Susan VanderPlas 3 \\
    Department of Statistics, University of Nebraska-Lincoln\\
      }
  \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf `You Draw It' Implementation of visually fitted trends
with \texttt{r2d3}}
  \end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
How do statistical regression results compare to intuitive, visually
fitted results? Fitting lines by eye through a set of points has been
explored since the 20th century. Common methods of fitting trends by eye
involve maneuvering a string, black thread, or ruler until the fit is
suitable, then drawing the line through the set of points. In 2015, the
New York Times introduced an interactive feature, called `You Draw It,'
where readers are asked to input their own assumptions about various
metrics and compare how these assumptions relate to reality. This
research is intended to implement `You Draw It', adapted from the New
York Times, as a way to measure the patterns we see in data.
\emph{(116/200 words)}
\end{abstract}

\noindent%
{\it Keywords:} graphics, user interaction, regression
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Our visual system is naturally built to look for structure and identify
patterns. For instance, points going down from left to right indicates a
negative correlation between the \(x\) and \(y\) variables. How can we
compare our intuitive visual sense of patterns to those determined by
statistical methods?

\hypertarget{testing-statistical-graphics}{%
\subsection{Testing Statistical
Graphics}\label{testing-statistical-graphics}}

Graphical tests are useful for studying the perception of statistical
graphs. Studies might ask participants to identify differences in
graphs, read information off of a chart accurately, use data to make
correct real-world decisions, or predict the next few observations. All
of these types of tests require different levels of use and manipulation
of the information being presented in the chart.

Efforts in the field of statistical graphics have developed graphical
testing tools and methods. Through experimentation, graphical testing
methods allow researchers to conduct studies geared at understanding
human ability to conduct tasks related to the perception of statistical
charts such as differentiation, prediction, estimation, and
extrapolation. The advancement of graphing software provides the tools
necessary to develop new methods of testing statistical graphics.

\hypertarget{fitting-trends-by-eye}{%
\subsection{Fitting Trends by Eye}\label{fitting-trends-by-eye}}

Initial studies in the 20th century explored the use of fitting lines by
eye through a set of points
\citep{finney1951subjective, mosteller1981eye}. Common methods of
fitting trends by eye involved maneuvering a string, black thread, or
ruler until the fit is suitable, then drawing the line through the set
of points. Recently, \citet{ciccione2021can} conducted a comprehensive
set of studies investigating human ability to detect trends in graphical
representations from a psychophysical approach.

In 2015, the New York Times introduced an interactive feature, called
`You Draw It' \citep{aisch2015you}, where readers input their own
assumptions about various metrics and compare how these assumptions
relate to reality. The New York Times team utilizes Data Driven
Documents (D3) that allows readers to predict these metrics through the
use of drawing a line on their computer screen with their mouse.

\hypertarget{research-objectives}{%
\subsection{Research Objectives}\label{research-objectives}}

The goal of this research is to implement `You Draw It', adapted from
the New York Times feature, as a way to measure the patterns we see in
data. Here, we provide technical details of the software development,
utilizing interactive graphics in R. We then share results from our
study which validates `You Draw It' as a method for graphical testing
and apply an appropriate data analysis method to the participant data.

\hypertarget{development}{%
\section{Development}\label{development}}

\hypertarget{you-draw-it-task}{%
\subsection{`You Draw It' Task}\label{you-draw-it-task}}

Users are shown an interactive scatter-plot
(\cref{fig:you-draw-it-example}) along with the prompt, ``Use your mouse
to fill in the trend in the yellow box region.'' The yellow box region
moves along as the user draws their trend-line, providing a visual cue
which indicates where the user still needs to complete a trend line.
After the entire domain has been visually estimated or predicted, the
yellow region disappears, indicating the participant has completed the
task. Visit \href{https://bit.ly/3zHnlgp\%7D}{bit.ly/3zHnlgp} for a test
applet.

\hypertarget{source-code}{%
\subsection{Source Code}\label{source-code}}

Data Driven Documents (D3), a JavaScript-based graphing framework that
facilitates user interaction, is used to create the `You Draw It'
visual. A challenge of working with D3 is the environment necessary to
display the graphics and images. The \texttt{r2d3} package
\citep{r2d3pkg} provides an efficient integration of D3 visuals into R
HTML formats. We integrate the D3 visual source code into an R shiny
\citep{shinypkg} application in order to allow for user interaction and
data collection.

We conduct all data simulation and processing in R and output two data
sets - \emph{point data} and \emph{line data} - containing (\(x\),
\(y\)) coordinates corresponding to either a simulated point or fitted
value predicted by a statistical model respectively. Then, the
\texttt{r2d3} package converts the data sets in R to JSON to be
interpreted by the D3.js code. We define functions in D3.js to draw the
initial plot and set up drawable points for the user drawn line. Drag
events in D3.js are utilized to react to observe and react to user
input. Shiny Messages are used to communicate the user interaction
between the D3 code and the R environment. The plot is then rendered and
updated on user interaction into the R shiny application with the
\texttt{RenderD3} and \texttt{d3Output} functions.

Parameters for aesthetic design choices are defined in a list of options
and \texttt{r2d3} passes these to the D3.js code. For instance, we can
specify the buffer space allowed for the \(x\) and \(y\) axes to avoid
users to anchor their lines to the axes limits.

For \texttt{D3.js} source code, visit
\href{https://bit.ly/3teA3lM}{bit.ly/3teA3lM}.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{images/ydi-stimuli}

}

\caption{'You Draw It' interaction task plot as shown to user. \textbf{Left:} illustrates what user first sees with the prompt \textit{`Use your mouse to fill in the trend in the yellow box region.'} \textbf{Middle:} illustrates what the user sees while completing the task. \textbf{Right:} illustrates the users finished trend line.}\label{fig:you-draw-it-example}
\end{figure}

\hypertarget{application}{%
\section{Application}\label{application}}

\hypertarget{tool-validation}{%
\subsection{Tool Validation}\label{tool-validation}}

We conducted a study in order to validate `You Draw It' as a method for
graphical testing, comparing results to the less technological method
utilized in \citet{mosteller1981eye}. Results from our study were
consistent with those found in the previous study; when shown points
following a linear trend, participants tended to fit the slope of the
first principal component over the slope of the least-squares regression
line. This trend was most prominent when shown data simulated with
larger variances. This study reinforces the differences between
intuitive visual model fitting and statistical model fitting, providing
information about human perception as it relates to the use of
statistical graphics.

\hypertarget{data-analysis}{%
\subsection{Data Analysis}\label{data-analysis}}

Feedback data from conducted studies are collected and stored in a
database for analysis. Within the collected feedback data, we know the
simulated data points, the predicted values from the statistical model,
and the predicted values from the user drawn line. In our initial
studies, a unique data set was simulated independently for each
participant. Therefore, we evaluate the accuracy of the user drawn line
by observing the deviation, vertical residuals, between the user drawn
line and the predicted values from the statistical model. We use a
Generalized Additive Mixed Model (GAMM) to model the vertical residuals
in order to statistically compare visually fitted trends to actual
metrics, simulated data models, or statistical regression results. A
benefit of using a GAMM is the estimation of smoothing splines, allowing
for flexibility in the residual trend.

\hypertarget{future-work}{%
\section{Future Work}\label{future-work}}

In this work, we implemented and validated `You Draw It' as a way to
measure the patterns we see in data. We demonstrated the use of
generalized additive models to statistically model participant data.
While technical details of the development process are presented here,
we intend to create an R package designed for easy implementation of
`You Draw It' task plots in order to make this tool accessible to other
researchers. Further investigation is necessary to implement this method
real data in order to facilitate scientific communication.

\bibliographystyle{agsm}
\bibliography{bibliography.bib}

\end{document}

### 2022-06-29 2022-sdss-you-draw-it-manuscript.Rmd
initial attempt

---
title: "'You Draw It': Implementation of visually fitted trends with `r2d3`"
bibliography: references.bib
abstract: "How do statistical regression results compare to intuitive, visually fitted results? Fitting lines by eye through a set of points has been explored since the 20th century. Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. In 2015, the New York Times introduced an interactive feature, called 'You Draw It,' where readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality. This research is intended to implement 'You Draw It', adapted from the New York Times, as a way to measure the patterns we see in data."

keywords: 'graphics, user interaction, regression'
output: 
  pdf_document:
    keep_tex: true
    # template: template-blinded.tex
    template: template.tex
header-includes:
   - \usepackage[dvipsnames]{xcolor} % colors
   - \newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
   - \newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
   - \newcommand{\rh}[1]{{\textcolor{Green}{#1}}}
   - \usepackage[capitalise]{cleveref}
   - \newcommand\pcref[1]{(\cref{#1})}

---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = F, 
                      eval = T, 
                      fig.path="./images/", 
                      fig.env="figure*", 
                      fig.align = "center")
library(readr)
library(tidyverse)
library(knitr)
library(bibtex)
```

\section{Introduction}

We all use statistical graphics, but how do we know that the graphics we use are communicating properly?
Through experimentation, graphical testing methods allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation.
All of these types of tests require different levels of use and manipulation of the information being presented in the chart.
Efforts in the field of statistical graphics have developed graphical testing tools and methods, such as the lineup protocol \cite{buja2009statistical}. 
The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics.

\subsection{Measuring Patterns and Trends}

One such aspect of interest is the ability to identify and detect trends in data. 
Our visual system is naturally built to look for structure and identify patterns. 
For instance, points going down from left to right indicates a negative correlation between the $x$ and $y$ variables.

Initial studies in the 20th century explored the use of fitting lines by eye through a set of points \cite{finney1951subjective,mosteller1981eye}. Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points. Recently, Ciccione and Dehaene \shortcite{ciccione2021can} conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.

In 2015, the New York Times introduced an interactive feature, called 'You Draw It' \cite{aisch2015you}, where readers input their own assumptions about various metrics and compare how these assumptions relate to reality. The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics through the use of drawing a line on their computer screen with their mouse.

What remains to be determined is how we can compare our intuitive visual sense of patterns to those determined by statistical methods.

\subsection{Research Objectives}
The goal of this research is to implement ‘You Draw It’, adapted from the New York Times feature, as a way to measure the patterns we see in data. Here, we provide technical details of the software development, utilizing interactive graphics in R. We then share results from our study which validates 'You Draw It' as a method for graphical testing and apply an appropriate data analysis method to the participant data.

\section{Development}

\subsection{`You Draw It' Task}

When completing the graphical task, users are shown an interactive scatter-plot (Figure 1) along with the prompt, "Use your mouse to fill in the trend in the yellow box region." The yellow box region moves along as the user draws their trend-line, providing a visual cue which indicates where the user still needs to complete a trend line. After the entire domain has been visually estimated or predicted, the yellow region disappears, indicating the participant has completed the task. Prior to study participation, example gifs are shown and users are asked to complete practice plots to ease the learning curve of the task. Visit \url{emily-robinson.shinyapps.io/can-you-draw-it} for a test applet.

<!-- # ```{r, you-draw-it-example, echo = F, out.width = "100%", fig.align='center', fig.cap = "'You Draw It' interaction task plot as shown to user. \\textbf{Left:} illustrates what user first sees with the prompt \\textit{`Use your mouse to fill in the trend in the yellow box region.'} \\textbf{Middle:} illustrates what the user sees while completing the task. \\textbf{Right:} illustrates the users finished trend line."} -->
<!-- # knitr::include_graphics("images/ydi-stimuli.png") -->
<!-- # ``` -->

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/ydi-stimuli}}
\caption{You Draw It' task plot as shown to user. \\
\textbf{left:} illustrates what user first sees with the prompt \textit{Use your mouse to fill in the trend in the yellow box region.} \\
\textbf{middle:} illustrates what the user sees while completing the task.\\
\textbf{right:} illustrates the users finished trend line.}
\label{you-draw-it-task-plot}
\end{center}
\end{figure}

\subsection{Code Sketch}

Using Shiny \cite{shinypkg} and JavaScript, we modified the New York Times 'You Draw It' feature for the purpose of testing statistical graphics allowing us to incorporate user interaction, conduct studies online, and store participant responses.

Data Driven Documents (D3) \cite{bostock2011d3}, a JavaScript-based graphing framework that facilitates user interaction, was used to create the 'You Draw It' visual. Major news and research organizations such as the New York Times, FiveThirtyEight, Washington Post, and the pew Research Center use D3 to create and customize graphics. A challenge of working with D3 is the environment necessary to display the graphics and images. The `r2d3` package \cite{r2d3pkg} provides an efficient integration of D3 visuals into R HTML formats. We integrate the D3 visual source code into an R Shiny \cite{shinypkg} application in order to allow for user interaction and data collection.

Figure 2 illustrates the initial setup of the visual stimuli along with the iterative process between the user interaction and plotting. We conducted all data simulation and processing in R and output two data sets - *point data* and *line data* - containing (x, y) coordinates corresponding to either a simulated point or fitted value predicted by a statistical model respectively. Then, the r2d3 package converts the data sets in R to JavaScript Object Notation (JSON) to be interpreted by the `D3.js` code. We define functions in `D3.js` to draw the initial plot and set up drawable points for the user drawn line. Drag events in `D3.js` are utilized to observe and react to user input. Shiny Messages are used to communicate the user interaction between the `D3.js` code and the R environment. The plot is then rendered and updated on user interaction into the R shiny application with the `RenderD3` and `d3Output` functions. Once the user is done drawing the line, we saved the results of the drawn line to a database for analysis.

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/code-sketch-2}}
\caption{Sketch of underlying code for 'You Draw It', illustrating the data simulation conducted in R, the initial setup of the visual stimuli with D3 source code, along with the iterative process between the user interaction and plotting in Shiny. Once the user is done drawing the line, we saved the results of the drawn line to a database for analysis.}
\label{you-draw-it-code-sketch}
\end{center}
\end{figure}

Parameters for aesthetic design choices are defined in a list of options and `r2d3` passes these to the D3.js code. For instance, we can specify the buffer space allowed for the $x$ and $y$ axes to avoid users to anchor their lines to the axes limits. For `D3.js` source code, visit \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.

\subsection{Challenges}

During the development process, there were a few challenges we had to overcome. 
We documented these challenges with GitHub commits and outline places where frustration occurred.
Data Driven documents uses Scalable Vector Graphics (SVG), thus requiring careful transformation between the pixels and plot coordinates to align the simulated data and the user drawn line appropriately.
With the layered framework of SVG's, it was important to place the layers in the right order so certain features would appear where desired. 
For example, we wanted the opacity of the yellow box region to appear under the points while still showing the grid lines.
As previously mentioned, `r2d3` automatically converts the data set to a JSON file to be interpreted by the underlying source code, however we provided two data sets (point data and line data) and converted the two data sets to a JSON file before passing them to the `r2d3` argument so that we could reference both data sets within the `D3.js` code.
One constraint of the development code to date is that it is only built to work with one-to-one functions and does not allow for multiple estimates or predictions for one $x$-value.

\section{Application}

\subsection{Tool Validation}

We conducted a study in order to validate 'You Draw It' as a method for graphical testing, comparing results to the less technological method utilized in Mosteller et al. \shortcite{mosteller1981eye}. Results from our study were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the ordinary least-squares regression line (Figure 3). This trend was most prominent when shown data simulated with larger variances. This study reinforces the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/pca-plot}}
\caption{Comparison between an OLS regression equation which minimizes the vertical distance of points from the line and a regression equation with a slope calculated by the first principal component which minimizes the smallest distance of points from the line.}
\label{pca-plot}
\end{center}
\end{figure}

\subsection{Data Analysis}

Feedback data from conducted studies are collected and stored in a database for analysis (Figure 4). Within the collected feedback data, we know the simulated data points, the predicted values from the statistical model, and the predicted values from the user drawn line. In our initial studies, a unique data set was simulated independently for each participant. Therefore, we evaluate the accuracy of the user drawn line by observing the deviation, vertical residuals, between the user drawn line and the predicted values from the statistical model. We use a Generalized Additive Mixed Model (GAMM) to model the vertical residuals in order to statistically compare visually fitted trends to actual metrics, simulated data models, or statistical regression results. A benefit of using a GAMM is the estimation of smoothing splines, allowing for flexibility in the residual trend.

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-trial-plot}}
\caption{Example of three trend lines showing the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points}
\label{eyefitting-trial-plot}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-lmer-plot}}
\caption{Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a linear mixed model.}
\label{eyefitting-lmer-plot}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-gamm-plot}}
\caption{Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a generalized additive mixed model.}
\label{eyefitting-gamm-plot}
\end{center}
\end{figure}

\begin{figure}[ht]
\begin{center}
\centerline{\includegraphics[width=\columnwidth]{images/exponential-yloess-spaghetti-plot-2-1}}
\caption{Spaghetti plot of results from a study which asked participants to forcast trends of exponentially increasing data. Participants drawn lines on the linear scale are shown in blue and the log scale are shown in orange. Variability in the statistically fitted regression lines occured due to a unique data set being simulated for each individual; the gray band shows the range fitted values from the statistically fitted regression lines.}
\label{exponential-yloess-spaghetti-plot-2-1}
\end{center}
\end{figure}

\section{Future Work}

In this work, we implemented and validated ‘You Draw It’ as a way to measure the patterns we see in data. We demonstrated the use of generalized additive models to statistically model participant data. While technical details of the development process are presented here, we intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers. Further investigation is necessary to implement this method real data in order to facilitate scientific communication.

### 2022-06-29 2022-sdss-you-draw-it-manuscript.tex
initial attempt

--- 2022-06-29 2022-sdss-you-draw-it-manuscript.Rmd
@@
----
-title: "'You Draw It': Implementation of visually fitted trends with `r2d3`"
-bibliography: references.bib
-abstract: "How do statistical regression results compare to intuitive, visually fitted results?
-Fitting lines by eye through a set of points has been explored since the 20th century.
-Common methods of fitting trends by eye involve maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points.
-In 2015, the New York Times introduced an interactive feature, called 'You Draw It,' where readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
-This research is intended to implement 'You Draw It', adapted from the New York Times, as a way to measure the patterns we see in data."
+\documentclass[10pt]{article}
 
-keywords: 'graphics, user interaction, regression'
-output: 
-  pdf_document:
-    keep_tex: true
-    # template: template-blinded.tex
-    template: template.tex
-header-includes:
-   - \usepackage[dvipsnames]{xcolor} % colors
-   - \newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
-   - \newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
-   - \newcommand{\rh}[1]{{\textcolor{Green}{#1}}}
-   - \usepackage[capitalise]{cleveref}
-   - \newcommand\pcref[1]{(\cref{#1})}
+\usepackage{sdss2020} % Uses Times Roman font (either newtx or times package)
+\usepackage{url}
+\usepackage{latexsym}
+\usepackage{amsmath, amsthm, amsfonts}
+\usepackage{algorithm, algorithmic}  
+\usepackage{graphicx}
 
----
+\usepackage[dvipsnames]{xcolor} % colors
+\newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
+\newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
+\newcommand{\rh}[1]{{\textcolor{Green}{#1}}}
+\usepackage[capitalise]{cleveref}
+\newcommand\pcref[1]{(\cref{#1})}
 
-```{r setup, include = FALSE}
-knitr::opts_chunk$set(echo = F, 
-                      eval = T, 
-                      fig.path="./images/", 
-                      fig.env="figure*", 
-                      fig.align = "center")
-library(readr)
-library(tidyverse)
-library(knitr)
-library(bibtex)
-```
+\title{`You Draw It': Implementation of visually fitted trends with
+\texttt{r2d3}}
+
+%\author{by }
+\author{
+  Emily A. Robinson \\
+  \small{University of Nebraska - Lincoln}\\
+  \small{Lincoln, NE} \\
+  \small{\tt emily.robinson@huskers.unl.edu} \\\And
+ Reka Howard \\
+  \small{University of Nebraska - Lincoln}\\
+  \small{Lincoln, NE} \\
+  \small{\tt rekahoward@unl.edu} \\\And
+  Susan VanderPlas \\
+  \small{University of Nebraska - Lincoln}\\
+  \small{Lincoln, NE} \\
+  \small{\tt susan.vanderplas@unl.edu} \\}
+
+\date{}
+
+\begin{document}
+\maketitle
+\begin{abstract}
+How do statistical regression results compare to intuitive, visually
+fitted results? Fitting lines by eye through a set of points has been
+explored since the 20th century. Common methods of fitting trends by eye
+involve maneuvering a string, black thread, or ruler until the fit is
+suitable, then drawing the line through the set of points. In 2015, the
+New York Times introduced an interactive feature, called `You Draw It,'
+where readers are asked to input their own assumptions about various
+metrics and compare how these assumptions relate to reality. This
+research is intended to implement `You Draw It,' adapted from the New
+York Times, as a way to measure the patterns we see in data.
+\end{abstract}
+
+{\bf Keywords:} graphics, user interaction, regression
 
@@
 
-We all use statistical graphics, but how do we know that the graphics we use are communicating properly?
-Through experimentation, graphical testing methods allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation.
-All of these types of tests require different levels of use and manipulation of the information being presented in the chart.
-Efforts in the field of statistical graphics have developed graphical testing tools and methods, such as the lineup protocol \cite{buja2009statistical}. 
-The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics.
+We all use statistical graphics, but how do we know that the graphics we
+use are communicating properly? Through experimentation, graphical
+testing methods allow researchers to conduct studies geared at
+understanding human ability to conduct tasks related to the perception
+of statistical charts such as differentiation, prediction, estimation,
+and extrapolation. All of these types of tests require different levels
+of use and manipulation of the information being presented in the chart.
+Efforts in the field of statistical graphics have developed graphical
+testing tools and methods, such as the lineup protocol
+\cite{buja2009statistical}. The advancement of graphing software
+provides the tools necessary to develop new methods of testing
+statistical graphics.
 
@@
 
-One such aspect of interest is the ability to identify and detect trends in data. 
-Our visual system is naturally built to look for structure and identify patterns. 
-For instance, points going down from left to right indicates a negative correlation between the $x$ and $y$ variables.
+One such aspect of interest is the ability to identify and detect trends
+in data. Our visual system is naturally built to look for structure and
+identify patterns. For instance, points going down from left to right
+indicates a negative correlation between the \(x\) and \(y\) variables.
 
-Initial studies in the 20th century explored the use of fitting lines by eye through a set of points \cite{finney1951subjective,mosteller1981eye}.
-Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points.
-Recently, Ciccione and Dehaene \shortcite{ciccione2021can} conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.
+Initial studies in the 20th century explored the use of fitting lines by
+eye through a set of points
+\cite{finney1951subjective,mosteller1981eye}. Common methods of fitting
+trends by eye involved maneuvering a string, black thread, or ruler
+until the fit is suitable, then drawing the line through the set of
+points. Recently, Ciccione and Dehaene \shortcite{ciccione2021can}
+conducted a comprehensive set of studies investigating human ability to
+detect trends in graphical representations from a psychophysical
+approach.
 
-In 2015, the New York Times introduced an interactive feature, called 'You Draw It' \cite{aisch2015you}, where readers input their own assumptions about various metrics and compare how these assumptions relate to reality.
-The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics through the use of drawing a line on their computer screen with their mouse.
+In 2015, the New York Times introduced an interactive feature, called
+`You Draw It' \cite{aisch2015you}, where readers input their own
+assumptions about various metrics and compare how these assumptions
+relate to reality. The New York Times team utilizes Data Driven
+Documents (D3) that allows readers to predict these metrics through the
+use of drawing a line on their computer screen with their mouse.
 
-What remains to be determined is how we can compare our intuitive visual sense of patterns to those determined by statistical methods.
+What remains to be determined is how we can compare our intuitive visual
+sense of patterns to those determined by statistical methods.
 
 \subsection{Research Objectives}
-The goal of this research is to implement ‘You Draw It’, adapted from the New York Times feature, as a way to measure the patterns we see in data.
-Here, we provide technical details of the software development, utilizing interactive graphics in R.
-We then share results from our study which validates 'You Draw It' as a method for graphical testing and apply an appropriate data analysis method to the participant data.
+
+The goal of this research is to implement `You Draw It,' adapted from
+the New York Times feature, as a way to measure the patterns we see in
+data. Here, we provide technical details of the software development,
+utilizing interactive graphics in R. We then share results from our
+study which validates `You Draw It' as a method for graphical testing
+and apply an appropriate data analysis method to the participant data.
 
@@
 
-When completing the graphical task, users are shown an interactive scatter-plot (Figure 1) along with the prompt, "Use your mouse to fill in the trend in the yellow box region." The yellow box region moves along as the user draws their trend-line, providing a visual cue which indicates where the user still needs to complete a trend line.
-After the entire domain has been visually estimated or predicted, the yellow region disappears, indicating the participant has completed the task.
-Prior to study participation, example gifs are shown and users are asked to complete practice plots to ease the learning curve of the task.
-Visit \url{emily-robinson.shinyapps.io/can-you-draw-it} for a test applet.
-
-<!-- # ```{r, you-draw-it-example, echo = F, out.width = "100%", fig.align='center', fig.cap = "'You Draw It' interaction task plot as shown to user.
-\\textbf{Left:} illustrates what user first sees with the prompt \\textit{`Use your mouse to fill in the trend in the yellow box region.'} \\textbf{Middle:} illustrates what the user sees while completing the task.
-\\textbf{Right:} illustrates the users finished trend line."} -->
-<!-- # knitr::include_graphics("images/ydi-stimuli.png") -->
-<!-- # ``` -->
+When completing the graphical task, users are shown an interactive
+scatter-plot (Figure 1) along with the prompt, ``Use your mouse to fill
+in the trend in the yellow box region.'' The yellow box region moves
+along as the user draws their trend-line, providing a visual cue which
+indicates where the user still needs to complete a trend line. After the
+entire domain has been visually estimated or predicted, the yellow
+region disappears, indicating the participant has completed the task.
+Prior to study participation, example gifs are shown and users are asked
+to complete practice plots to ease the learning curve of the task. Visit
+\url{emily-robinson.shinyapps.io/can-you-draw-it} for a test applet.
 
@@
 
-Using Shiny \cite{shinypkg} and JavaScript, we modified the New York Times 'You Draw It' feature for the purpose of testing statistical graphics allowing us to incorporate user interaction, conduct studies online, and store participant responses.
+Using Shiny \cite{shinypkg} and JavaScript, we modified the New York
+Times `You Draw It' feature for the purpose of testing statistical
+graphics allowing us to incorporate user interaction, conduct studies
+online, and store participant responses.
 
-Data Driven Documents (D3) \cite{bostock2011d3}, a JavaScript-based graphing framework that facilitates user interaction, was used to create the 'You Draw It' visual.
-Major news and research organizations such as the New York Times, FiveThirtyEight, Washington Post, and the pew Research Center use D3 to create and customize graphics.
-A challenge of working with D3 is the environment necessary to display the graphics and images.
-The `r2d3` package \cite{r2d3pkg} provides an efficient integration of D3 visuals into R HTML formats.
-We integrate the D3 visual source code into an R Shiny \cite{shinypkg} application in order to allow for user interaction and data collection.
+Data Driven Documents (D3) \cite{bostock2011d3}, a JavaScript-based
+graphing framework that facilitates user interaction, was used to create
+the `You Draw It' visual. Major news and research organizations such as
+the New York Times, FiveThirtyEight, Washington Post, and the pew
+Research Center use D3 to create and customize graphics. A challenge of
+working with D3 is the environment necessary to display the graphics and
+images. The \texttt{r2d3} package \cite{r2d3pkg} provides an efficient
+integration of D3 visuals into R HTML formats. We integrate the D3
+visual source code into an R Shiny \cite{shinypkg} application in order
+to allow for user interaction and data collection.
 
-Figure 2 illustrates the initial setup of the visual stimuli along with the iterative process between the user interaction and plotting.
-We conducted all data simulation and processing in R and output two data sets - *point data* and *line data* - containing (x, y) coordinates corresponding to either a simulated point or fitted value predicted by a statistical model respectively.
-Then, the r2d3 package converts the data sets in R to JavaScript Object Notation (JSON) to be interpreted by the `D3.js` code.
-We define functions in `D3.js` to draw the initial plot and set up drawable points for the user drawn line.
-Drag events in `D3.js` are utilized to observe and react to user input.
-Shiny Messages are used to communicate the user interaction between the `D3.js` code and the R environment.
-The plot is then rendered and updated on user interaction into the R shiny application with the `RenderD3` and `d3Output` functions.
-Once the user is done drawing the line, we saved the results of the drawn line to a database for analysis.
+Figure 2 illustrates the initial setup of the visual stimuli along with
+the iterative process between the user interaction and plotting. We
+conducted all data simulation and processing in R and output two data
+sets - \emph{point data} and \emph{line data} - containing (x, y)
+coordinates corresponding to either a simulated point or fitted value
+predicted by a statistical model respectively. Then, the r2d3 package
+converts the data sets in R to JavaScript Object Notation (JSON) to be
+interpreted by the \texttt{D3.js} code. We define functions in
+\texttt{D3.js} to draw the initial plot and set up drawable points for
+the user drawn line. Drag events in \texttt{D3.js} are utilized to
+observe and react to user input. Shiny Messages are used to communicate
+the user interaction between the \texttt{D3.js} code and the R
+environment. The plot is then rendered and updated on user interaction
+into the R shiny application with the \texttt{RenderD3} and
+\texttt{d3Output} functions. Once the user is done drawing the line, we
+saved the results of the drawn line to a database for analysis.
 
@@
 
-Parameters for aesthetic design choices are defined in a list of options and `r2d3` passes these to the D3.js code.
-For instance, we can specify the buffer space allowed for the $x$ and $y$ axes to avoid users to anchor their lines to the axes limits.
-For `D3.js` source code, visit \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.
+Parameters for aesthetic design choices are defined in a list of options
+and \texttt{r2d3} passes these to the D3.js code. For instance, we can
+specify the buffer space allowed for the \(x\) and \(y\) axes to avoid
+users to anchor their lines to the axes limits. For \texttt{D3.js}
+source code, visit
+\url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.
 
@@
 
-During the development process, there were a few challenges we had to overcome. 
-We documented these challenges with GitHub commits and outline places where frustration occurred.
-Data Driven documents uses Scalable Vector Graphics (SVG), thus requiring careful transformation between the pixels and plot coordinates to align the simulated data and the user drawn line appropriately.
-With the layered framework of SVG's, it was important to place the layers in the right order so certain features would appear where desired. 
-For example, we wanted the opacity of the yellow box region to appear under the points while still showing the grid lines.
-As previously mentioned, `r2d3` automatically converts the data set to a JSON file to be interpreted by the underlying source code, however we provided two data sets (point data and line data) and converted the two data sets to a JSON file before passing them to the `r2d3` argument so that we could reference both data sets within the `D3.js` code.
-One constraint of the development code to date is that it is only built to work with one-to-one functions and does not allow for multiple estimates or predictions for one $x$-value.
+During the development process, there were a few challenges we had to
+overcome. We documented these challenges with GitHub commits and outline
+places where frustration occurred. Data Driven documents uses Scalable
+Vector Graphics (SVG), thus requiring careful transformation between the
+pixels and plot coordinates to align the simulated data and the user
+drawn line appropriately. With the layered framework of SVG's, it was
+important to place the layers in the right order so certain features
+would appear where desired. For example, we wanted the opacity of the
+yellow box region to appear under the points while still showing the
+grid lines. As previously mentioned, \texttt{r2d3} automatically
+converts the data set to a JSON file to be interpreted by the underlying
+source code, however we provided two data sets (point data and line
+data) and converted the two data sets to a JSON file before passing them
+to the \texttt{r2d3} argument so that we could reference both data sets
+within the \texttt{D3.js} code. One constraint of the development code
+to date is that it is only built to work with one-to-one functions and
+does not allow for multiple estimates or predictions for one
+\(x\)-value.
 
@@
 
-We conducted a study in order to validate 'You Draw It' as a method for graphical testing, comparing results to the less technological method utilized in Mosteller et al.
-\shortcite{mosteller1981eye}.
-Results from our study were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the ordinary least-squares regression line (Figure 3).
-This trend was most prominent when shown data simulated with larger variances.
-This study reinforces the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.
+We conducted a study in order to validate `You Draw It' as a method for
+graphical testing, comparing results to the less technological method
+utilized in Mosteller et al.~\shortcite{mosteller1981eye}. Results from
+our study were consistent with those found in the previous study; when
+shown points following a linear trend, participants tended to fit the
+slope of the first principal component over the slope of the ordinary
+least-squares regression line (Figure 3). This trend was most prominent
+when shown data simulated with larger variances. This study reinforces
+the differences between intuitive visual model fitting and statistical
+model fitting, providing information about human perception as it
+relates to the use of statistical graphics.
 
@@
 
-Feedback data from conducted studies are collected and stored in a database for analysis (Figure 4).
-Within the collected feedback data, we know the simulated data points, the predicted values from the statistical model, and the predicted values from the user drawn line.
-In our initial studies, a unique data set was simulated independently for each participant.
-Therefore, we evaluate the accuracy of the user drawn line by observing the deviation, vertical residuals, between the user drawn line and the predicted values from the statistical model.
-We use a Generalized Additive Mixed Model (GAMM) to model the vertical residuals in order to statistically compare visually fitted trends to actual metrics, simulated data models, or statistical regression results.
-A benefit of using a GAMM is the estimation of smoothing splines, allowing for flexibility in the residual trend.
+Feedback data from conducted studies are collected and stored in a
+database for analysis (Figure 4). Within the collected feedback data, we
+know the simulated data points, the predicted values from the
+statistical model, and the predicted values from the user drawn line. In
+our initial studies, a unique data set was simulated independently for
+each participant. Therefore, we evaluate the accuracy of the user drawn
+line by observing the deviation, vertical residuals, between the user
+drawn line and the predicted values from the statistical model. We use a
+Generalized Additive Mixed Model (GAMM) to model the vertical residuals
+in order to statistically compare visually fitted trends to actual
+metrics, simulated data models, or statistical regression results. A
+benefit of using a GAMM is the estimation of smoothing splines, allowing
+for flexibility in the residual trend.
 
@@
 
-In this work, we implemented and validated ‘You Draw It’ as a way to measure the patterns we see in data.
-We demonstrated the use of generalized additive models to statistically model participant data.
-While technical details of the development process are presented here, we intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers.
-Further investigation is necessary to implement this method real data in order to facilitate scientific communication.
+In this work, we implemented and validated `You Draw It' as a way to
+measure the patterns we see in data. We demonstrated the use of
+generalized additive models to statistically model participant data.
+While technical details of the development process are presented here,
+we intend to create an R package designed for easy implementation of
+`You Draw It' task plots in order to make this tool accessible to other
+researchers. Further investigation is necessary to implement this method
+real data in order to facilitate scientific communication.
+
+\bibliographystyle{sdss2020}
+\bibliography{references.bib}
+% \bibliography{references}
+
+\end{document}

### 2022-07-01 2022-sdss-you-draw-it-manuscript.Rmd
rough rough draft

@@
 Prior to study participation, example gifs are shown and users are asked to complete practice plots to ease the learning curve of the task.
-Visit \url{emily-robinson.shinyapps.io/can-you-draw-it} for a test applet.
+^[Visit \url{emily-robinson.shinyapps.io/can-you-draw-it} for a test applet.]
 
@@
 For instance, we can specify the buffer space allowed for the $x$ and $y$ axes to avoid users to anchor their lines to the axes limits.
-For `D3.js` source code, visit \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.
+^[For `D3.js` source code, visit \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.]
 
@@
 This trend was most prominent when shown data simulated with larger variances.
-This study reinforces the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.
+This study reinforced the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.
 
@@
 
-Feedback data from conducted studies are collected and stored in a database for analysis (Figure 4).
-Within the collected feedback data, we know the simulated data points, the predicted values from the statistical model, and the predicted values from the user drawn line.
-In our initial studies, a unique data set was simulated independently for each participant.
-Therefore, we evaluate the accuracy of the user drawn line by observing the deviation, vertical residuals, between the user drawn line and the predicted values from the statistical model.
-We use a Generalized Additive Mixed Model (GAMM) to model the vertical residuals in order to statistically compare visually fitted trends to actual metrics, simulated data models, or statistical regression results.
-A benefit of using a GAMM is the estimation of smoothing splines, allowing for flexibility in the residual trend.
-
 \begin{figure}[ht]
@@
 \centerline{\includegraphics[width=\columnwidth]{images/eyefitting-trial-plot}}
-\caption{Example of three trend lines showing the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points}
+\caption{Example of validation feedback data where three trend lines show the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points}
 \label{eyefitting-trial-plot}
@@
 
+Feedback data from conducted studies were collected and stored in a database for analysis (Figure 4).
+Within the collected feedback data, we knew the simulated data points, the predicted values from the statistical model, and the predicted values from the user drawn line.
+In our initial studies, a unique data set was simulated independently for each participant.
+Therefore, we evaluate the accuracy of the user drawn line by observing the deviation, vertical residuals, between the user drawn line and the predicted values from the statistical model.
+Due to participant variation, we use mixed models to evaluate the vertical residuals in order to statistically compare visually fitted trends to actual metrics, simulated data models, or statistical regression results.
+In our validation study, we first analyzed the residual trend using a linear mixed model (LMM), thus constraining the fit to a linear trend (Figure 5).
+In addition, we fit a Generalized Additive Mixed Model (GAMM) to allow for flexibility in the residual trend and to generalized the method to nonlinear models (Figure 6).
+
 \begin{figure}[ht]
@@
 \centerline{\includegraphics[width=\columnwidth]{images/eyefitting-lmer-plot}}
-\caption{Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a linear mixed model.}
+\caption{Validation study estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a linear mixed model.}
 \label{eyefitting-lmer-plot}
@@
 \centerline{\includegraphics[width=\columnwidth]{images/eyefitting-gamm-plot}}
-\caption{Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a generalized additive mixed model.}
+\caption{Validation study estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a generalized additive mixed model.}
 \label{eyefitting-gamm-plot}
@@
 \end{figure}
+
+After validating the 'You Draw It' method, we conducted a study utilizing the new method to evaluate participants ability to make forecasts for exponentially increasing data on log and linear scales.
+Along with analyzing the feedback data with the GAMM method for flexibility due to nonlinear data, we used spaghetti plots to visualize participants forecasts compared to the nonlinear least squares statistical model and make comparisons between two chart design features (Figure 6).
 
@@
 
-In this work, we implemented and validated ‘You Draw It’ as a way to measure the patterns we see in data.
-We demonstrated the use of generalized additive models to statistically model participant data.
+In this work, we implemented and validated ‘You Draw It’ as a new method to measure the patterns we see in data.
+We demonstrated the use of generalized additive models to statistically model participant feedback data and generalized the analysis method to nonlinear trends.
 While technical details of the development process are presented here, we intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers.
-Further investigation is necessary to implement this method real data in order to facilitate scientific communication.
+Further investigation is necessary to implement this method with real data in order to facilitate scientific communication.

### 2022-07-01 2022-sdss-you-draw-it-manuscript.tex
rough rough draft

@@
 Prior to study participation, example gifs are shown and users are asked
-to complete practice plots to ease the learning curve of the task. Visit
-\url{emily-robinson.shinyapps.io/can-you-draw-it} for a test applet.
+to complete practice plots to ease the learning curve of the
+task.\footnote{Visit \url{emily-robinson.shinyapps.io/can-you-draw-it}
+  for a test applet.}
 
@@
 specify the buffer space allowed for the \(x\) and \(y\) axes to avoid
-users to anchor their lines to the axes limits. For \texttt{D3.js}
-source code, visit
-\url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.
+users to anchor their lines to the axes limits.\footnote{For
+  \texttt{D3.js} source code, visit
+  \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.}
 
@@
 least-squares regression line (Figure 3). This trend was most prominent
-when shown data simulated with larger variances. This study reinforces
+when shown data simulated with larger variances. This study reinforced
 the differences between intuitive visual model fitting and statistical
@@
 
-Feedback data from conducted studies are collected and stored in a
+\begin{figure}[ht]
+\begin{center}
+\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-trial-plot}}
+\caption{Example of validation feedback data where three trend lines show the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points}
+\label{eyefitting-trial-plot}
+\end{center}
+\end{figure}
+
+Feedback data from conducted studies were collected and stored in a
 database for analysis (Figure 4). Within the collected feedback data, we
-know the simulated data points, the predicted values from the
+knew the simulated data points, the predicted values from the
 statistical model, and the predicted values from the user drawn line. In
@@
 line by observing the deviation, vertical residuals, between the user
-drawn line and the predicted values from the statistical model. We use a
-Generalized Additive Mixed Model (GAMM) to model the vertical residuals
-in order to statistically compare visually fitted trends to actual
-metrics, simulated data models, or statistical regression results. A
-benefit of using a GAMM is the estimation of smoothing splines, allowing
-for flexibility in the residual trend.
-
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-trial-plot}}
-\caption{Example of three trend lines showing the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points}
-\label{eyefitting-trial-plot}
-\end{center}
-\end{figure}
+drawn line and the predicted values from the statistical model. Due to
+participant variation, we use mixed models to evaluate the vertical
+residuals in order to statistically compare visually fitted trends to
+actual metrics, simulated data models, or statistical regression
+results. In our validation study, we first analyzed the residual trend
+using a linear mixed model (LMM), thus constraining the fit to a linear
+trend (Figure 5). In addition, we fit a Generalized Additive Mixed Model
+(GAMM) to allow for flexibility in the residual trend and to generalized
+the method to nonlinear models (Figure 6).
 
@@
 \end{figure}
+
+After validating the `You Draw It' method, we conducted a study
+utilizing the new method to evaluate participants ability to make
+forecasts for exponentially increasing data on log and linear scales.
+Along with analyzing the feedback data with the GAMM method for
+flexibility due to nonlinear data, we used spaghetti plots to visualize
+participants forecasts compared to the nonlinear least squares
+statistical model and make comparisons between two chart design features
+(Figure 6).
 
@@
 
-In this work, we implemented and validated `You Draw It' as a way to
-measure the patterns we see in data. We demonstrated the use of
-generalized additive models to statistically model participant data.
-While technical details of the development process are presented here,
-we intend to create an R package designed for easy implementation of
-`You Draw It' task plots in order to make this tool accessible to other
+In this work, we implemented and validated `You Draw It' as a new method
+to measure the patterns we see in data. We demonstrated the use of
+generalized additive models to statistically model participant feedback
+data and generalized the analysis method to nonlinear trends. While
+technical details of the development process are presented here, we
+intend to create an R package designed for easy implementation of `You
+Draw It' task plots in order to make this tool accessible to other
 researchers. Further investigation is necessary to implement this method
-real data in order to facilitate scientific communication.
+with real data in order to facilitate scientific communication.
 

### 2022-07-05 2022-sdss-you-draw-it-manuscript.Rmd
Add references and example of mental operations

@@
 library(bibtex)
+theme_set(theme_bw())
+
+library(readxl)
+library(lubridate)
 ```
@@
 
-We all use statistical graphics, but how do we know that the graphics we use are communicating properly?
-Through experimentation, graphical testing methods allow researchers to conduct studies geared at understanding human ability to conduct tasks related to the perception of statistical charts such as differentiation, prediction, estimation, and extrapolation.
-All of these types of tests require different levels of use and manipulation of the information being presented in the chart.
-Efforts in the field of statistical graphics have developed graphical testing tools and methods, such as the lineup protocol \cite{buja2009statistical}. 
-The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics.
+We all use statistical graphics, but how do we know that the graphics we use are communicating \svp{effectively}?
+Through experimentation, graphical testing methods allow researchers to conduct studies \svp{designed to understand how we perceive graphics and perform graphical tasks} such as differentiation, prediction, estimation, and extrapolation.
+\svp{Each of these levels of interaction with a graph require a different method of engagement with and use of the information presented in a chart.}
+<!-- All of these types of tests require different levels of use and manipulation of the information being presented in the chart. -->
+<!-- Efforts in the field of statistical graphics have developed graphical testing tools and methods, such as the lineup protocol \cite{buja2009statistical}.  -->
+<!-- The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics. -->
+\svp{In this paper, we describe the adaptation of an old tool for graphical testing and evaluation, eye-fitting, for use in modern web-applications suitable for testing statistical graphics. 
+We present an empirical evaluation of this testing method for linear regression, and briefly discuss an extension of this method to non-linear applications.}
 
 \subsection{Measuring Patterns and Trends}
+\svp{One of the most common charts created is a scatter plot of points over time; these charts show up regularly in news articles and in scientific publications alike.
+These charts rely on our }ability to identify and detect trends in data. 
+Our visual system is naturally built to look for structure and identify patterns, \svp{including patterns and trends over time}. 
+\svp{As shown in \cref{fig:gas-prices}, a viewer engaging with a plot of weekly average gas prices over time will likely perform several cognitive operations.
+First, they scan the plot and assesses points to determine if there are any outliers or otherwise remarkable points.
+Then, they may fit a rough mental smooth/trend line to the points to summarize the useful information and remove variability.
+Finally, the viewer may pull in additional contextual information from long-term memory, seeking to explain variation in the mental 'trend' with supplemental information, such as COVID lock downs which reduced gasoline demand and the beginning of the war in Ukraine, which wreaked havoc on the supply and demand for global energy sources.}
 
-One such aspect of interest is the ability to identify and detect trends in data. 
-Our visual system is naturally built to look for structure and identify patterns. 
-For instance, points going down from left to right indicates a negative correlation between the $x$ and $y$ variables.
+```{r gas-prices, echo = F, message = F, warning = F}
+#| fig-cap: "Weekly average gas prices in the United States, 2019-June 2022.
+Additional mental operations a viewer might perform while looking at the plot are annotated in grey.
+\\cite{usenergyinformationadministrationWeeklyAllGrades2022}."
+#| out-width: "\\columnwidth"
+#| label: gas-prices
+#| fig-res: 300
+#| fig-width: 8
+#| fig-height: 4
+#| fig-env: figure
+
+gas <- read_xls("EMM_EPM0_PTE_NUS_DPGw.xls", skip = 3, sheet = 2) %>%
+  set_names(c("Date", "Price"))
+
+annotations <- tibble(
+  xmin = as.POSIXct(ymd(c("2020-03-15", "2022-02-26"))),
+  xmax = as.POSIXct(ymd(c("2020-05-30", "2022-06-29"))),
+  ymin = -Inf,
+  ymax = Inf,
+  label = c("COVID lockdowns", "War in Ukraine")
+)
+ggplot(gas, aes(x = Date, y = Price)) + 
+  geom_rect(data = annotations, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), alpha= 0.2, fill = "grey", inherit.aes = F) + 
+  geom_text(data = annotations, aes(x = xmin, y = 1, label = label), color = "grey") + 
+  ylab("Price per Gallon ($)") + 
+  ggtitle("Weekly Average Gas Prices in the United States, 2019 - 2022", subtitle = "US Energy Information Administration") + 
+  geom_point() + 
+  scale_x_datetime(limits = as.POSIXct(ymd("2019-01-01", "2022-06-29"))) + 
+  geom_line(se = F, color = "grey", method = "loess", span = .1, alpha = .5, stat = "smooth", size = 1.5)
+```
 
 Initial studies in the 20th century explored the use of fitting lines by eye through a set of points \cite{finney1951subjective,mosteller1981eye}.
-Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit is suitable, then drawing the line through the set of points.
-Recently, Ciccione and Dehaene \shortcite{ciccione2021can} conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.
+Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit was suitable, then drawing the line through the set of points.
+Recently, Ciccione and Dehaene \shortcite{ciccione2021can} conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.\svp{XXX and found what?
+Describe the results XXX}
 
-In 2015, the New York Times introduced an interactive feature, called 'You Draw It' \cite{aisch2015you}, where readers input their own assumptions about various metrics and compare how these assumptions relate to reality.
-The New York Times team utilizes Data Driven Documents (D3) that allows readers to predict these metrics through the use of drawing a line on their computer screen with their mouse.
+\svp{While psychologists and statisticians have been using eye-fitting techniques to assess our innate perceptual statistical modeling abilities, news organizations have used similar techniques in order to draw readers in and demonstrate the difference between readers' assumptions and reality.} In 2015, the New York Times introduced an interactive feature, called 'You Draw It' \cite{aisch2015you}, where readers input their own assumptions about various \svp{metrics of news interest} and compare these assumptions to reality.
+The New York Times team used Data Driven Documents (D3), \svp{a JavaScript library} that allows readers to \svp{interact with a chart directly by} drawing a line on their computer screen with a mouse.
+\svp{Despite the somewhat different purpose behind this feature, the D3 driven method used by the NYT is wonderfully intuitive, and does not require the assumption of linearity, making it much more adaptable to testing how viewers perceive and predict when presented with non-linear data.}
 
-What remains to be determined is how we can compare our intuitive visual sense of patterns to those determined by statistical methods.
-
-\subsection{Research Objectives}
-The goal of this research is to implement ‘You Draw It’, adapted from the New York Times feature, as a way to measure the patterns we see in data.
-Here, we provide technical details of the software development, utilizing interactive graphics in R.
-We then share results from our study which validates 'You Draw It' as a method for graphical testing and apply an appropriate data analysis method to the participant data.
+\svp{We set out to implement} ‘You Draw It’, adapted from the New York Times feature, as a way to measure the patterns we see in data. 
+Here, we provide technical details of the software development, utilizing interactive graphics in R\cite{r-software}. 
+We then share results from our study which validates 'You Draw It' as a method for graphical testing \svp{on linear trends} and apply an appropriate data analysis method to the participant data.
+\svp{We also briefly demonstrate the use of the 'You Draw It' method and analysis on nonlinear data.}
 

### 2022-07-05 2022-sdss-you-draw-it-manuscript.tex
Add references and example of mental operations

@@
 metrics and compare how these assumptions relate to reality. This
-research is intended to implement `You Draw It,' adapted from the New
+research is intended to implement `You Draw It', adapted from the New
 York Times, as a way to measure the patterns we see in data.
@@
 We all use statistical graphics, but how do we know that the graphics we
-use are communicating properly? Through experimentation, graphical
-testing methods allow researchers to conduct studies geared at
-understanding human ability to conduct tasks related to the perception
-of statistical charts such as differentiation, prediction, estimation,
-and extrapolation. All of these types of tests require different levels
-of use and manipulation of the information being presented in the chart.
-Efforts in the field of statistical graphics have developed graphical
-testing tools and methods, such as the lineup protocol
-\cite{buja2009statistical}. The advancement of graphing software
-provides the tools necessary to develop new methods of testing
-statistical graphics.
+use are communicating {\textcolor{RedOrange}{effectively}}? Through
+experimentation, graphical testing methods allow researchers to conduct
+studies
+{\textcolor{RedOrange}{designed to understand how we perceive graphics and perform graphical tasks}}
+such as differentiation, prediction, estimation, and extrapolation.
+{\textcolor{RedOrange}{Each of these levels of interaction with a graph require a different method of engagement with and use of the information presented in a chart.}}
+{\textcolor{RedOrange}{In this paper, we describe the adaptation of an old tool for graphical testing and evaluation, eye-fitting, for use in modern web-applications suitable for testing statistical graphics.
+We present an empirical evaluation of this testing method for linear regression, and briefly discuss an extension of this method to non-linear applications.}}
 
@@
 
-One such aspect of interest is the ability to identify and detect trends
-in data. Our visual system is naturally built to look for structure and
-identify patterns. For instance, points going down from left to right
-indicates a negative correlation between the \(x\) and \(y\) variables.
+{\textcolor{RedOrange}{One of the most common charts created is a scatter plot of points over time; these charts show up regularly in news articles and in scientific publications alike.
+These charts rely on our }}ability to identify and detect trends in
+data. Our visual system is naturally built to look for structure and
+identify patterns,
+{\textcolor{RedOrange}{including patterns and trends over time}}.
+{\textcolor{RedOrange}{As shown in \cref{fig:gas-prices}, a viewer engaging with a plot of weekly average gas prices over time will likely perform several cognitive operations.
+First, they scan the plot and assesses points to determine if there are any outliers or otherwise remarkable points.
+Then, they may fit a rough mental smooth/trend line to the points to summarize the useful information and remove variability.
+Finally, the viewer may pull in additional contextual information from long-term memory, seeking to explain variation in the mental 'trend' with supplemental information, such as COVID lock downs which reduced gasoline demand and the beginning of the war in Ukraine, which wreaked havoc on the supply and demand for global energy sources.}}
+
+\begin{figure}
+
+{\centering \includegraphics[width=\columnwidth]{./images/gas-prices-1}
+
+}
+
+\caption{Weekly average gas prices in the United States, 2019-June 2022.
+Additional mental operations a viewer might perform while looking at the plot are annotated in grey.
+\cite{usenergyinformationadministrationWeeklyAllGrades2022}.}\label{fig:gas-prices}
+\end{figure}
 
@@
 trends by eye involved maneuvering a string, black thread, or ruler
-until the fit is suitable, then drawing the line through the set of
+until the fit was suitable, then drawing the line through the set of
 points. Recently, Ciccione and Dehaene \shortcite{ciccione2021can}
@@
 detect trends in graphical representations from a psychophysical
-approach.
+approach.{\textcolor{RedOrange}{XXX and found what? Describe the results XXX}}
 
+{\textcolor{RedOrange}{While psychologists and statisticians have been using eye-fitting techniques to assess our innate perceptual statistical modeling abilities, news organizations have used similar techniques in order to draw readers in and demonstrate the difference between readers' assumptions and reality.}}
 In 2015, the New York Times introduced an interactive feature, called
 `You Draw It' \cite{aisch2015you}, where readers input their own
-assumptions about various metrics and compare how these assumptions
-relate to reality. The New York Times team utilizes Data Driven
-Documents (D3) that allows readers to predict these metrics through the
-use of drawing a line on their computer screen with their mouse.
+assumptions about various
+{\textcolor{RedOrange}{metrics of news interest}} and compare these
+assumptions to reality. The New York Times team used Data Driven
+Documents (D3), {\textcolor{RedOrange}{a JavaScript library}} that
+allows readers to
+{\textcolor{RedOrange}{interact with a chart directly by}} drawing a
+line on their computer screen with a mouse.
+{\textcolor{RedOrange}{Despite the somewhat different purpose behind this feature, the D3 driven method used by the NYT is wonderfully intuitive, and does not require the assumption of linearity, making it much more adaptable to testing how viewers perceive and predict when presented with non-linear data.}}
 
-What remains to be determined is how we can compare our intuitive visual
-sense of patterns to those determined by statistical methods.
-
-\subsection{Research Objectives}
-
-The goal of this research is to implement `You Draw It,' adapted from
-the New York Times feature, as a way to measure the patterns we see in
-data. Here, we provide technical details of the software development,
-utilizing interactive graphics in R. We then share results from our
-study which validates `You Draw It' as a method for graphical testing
-and apply an appropriate data analysis method to the participant data.
+{\textcolor{RedOrange}{We set out to implement}} `You Draw It', adapted
+from the New York Times feature, as a way to measure the patterns we see
+in data. Here, we provide technical details of the software development,
+utilizing interactive graphics in R\cite{r-software}. We then share
+results from our study which validates `You Draw It' as a method for
+graphical testing {\textcolor{RedOrange}{on linear trends}} and apply an
+appropriate data analysis method to the participant data.
+{\textcolor{RedOrange}{We also briefly demonstrate the use of the 'You Draw It' method and analysis on nonlinear data.}}
 
@@
 Prior to study participation, example gifs are shown and users are asked
-to complete practice plots to ease the learning curve of the
-task.\footnote{Visit \url{emily-robinson.shinyapps.io/can-you-draw-it}
-  for a test applet.}
+to complete practice plots to ease the learning curve of the task.
+\footnote{Visit \url{emily-robinson.shinyapps.io/can-you-draw-it} for a
+  test applet.}
 
@@
 specify the buffer space allowed for the \(x\) and \(y\) axes to avoid
-users to anchor their lines to the axes limits.\footnote{For
+users to anchor their lines to the axes limits. \footnote{For
   \texttt{D3.js} source code, visit

### 2022-07-05 2022-sdss-you-draw-it-manuscript.Rmd
Tweaking phrasing

@@
 \svp{One of the most common charts created is a scatter plot of points over time; these charts show up regularly in news articles and in scientific publications alike.
-These charts rely on our }ability to identify and detect trends in data. 
-Our visual system is naturally built to look for structure and identify patterns, \svp{including patterns and trends over time}. 
-\svp{As shown in \cref{fig:gas-prices}, a viewer engaging with a plot of weekly average gas prices over time will likely perform several cognitive operations.
+These charts rely on our} ability to identify and detect trends in data. 
+Our visual system is naturally built to look for structure and identify patterns, \svp{including patterns and trends over time; many times we do not even notice this process happening subconsciously}.
+\svp{As shown in \cref{fig:gas-prices}, a viewer engaging with a plot of weekly average gas prices over time may perform several cognitive operations.
 First, they scan the plot and assesses points to determine if there are any outliers or otherwise remarkable points.
 Then, they may fit a rough mental smooth/trend line to the points to summarize the useful information and remove variability.
-Finally, the viewer may pull in additional contextual information from long-term memory, seeking to explain variation in the mental 'trend' with supplemental information, such as COVID lock downs which reduced gasoline demand and the beginning of the war in Ukraine, which wreaked havoc on the supply and demand for global energy sources.}
+Finally, an interested and engaged viewer may pull in additional contextual information from long-term memory, seeking to explain variation in the mental 'trend' with supplemental information, such as COVID lock downs which reduced gasoline demand and the beginning of the war in Ukraine, which wreaked havoc on the supply and demand for global energy sources.}
 

### 2022-07-05 2022-sdss-you-draw-it-manuscript.tex
Tweaking phrasing

@@
 {\textcolor{RedOrange}{One of the most common charts created is a scatter plot of points over time; these charts show up regularly in news articles and in scientific publications alike.
-These charts rely on our }}ability to identify and detect trends in
+These charts rely on our}} ability to identify and detect trends in
 data. Our visual system is naturally built to look for structure and
 identify patterns,
-{\textcolor{RedOrange}{including patterns and trends over time}}.
-{\textcolor{RedOrange}{As shown in \cref{fig:gas-prices}, a viewer engaging with a plot of weekly average gas prices over time will likely perform several cognitive operations.
+{\textcolor{RedOrange}{including patterns and trends over time; many times we do not even notice this process happening subconsciously}}.
+{\textcolor{RedOrange}{As shown in \cref{fig:gas-prices}, a viewer engaging with a plot of weekly average gas prices over time may perform several cognitive operations.
 First, they scan the plot and assesses points to determine if there are any outliers or otherwise remarkable points.
 Then, they may fit a rough mental smooth/trend line to the points to summarize the useful information and remove variability.
-Finally, the viewer may pull in additional contextual information from long-term memory, seeking to explain variation in the mental 'trend' with supplemental information, such as COVID lock downs which reduced gasoline demand and the beginning of the war in Ukraine, which wreaked havoc on the supply and demand for global energy sources.}}
+Finally, an interested and engaged viewer may pull in additional contextual information from long-term memory, seeking to explain variation in the mental 'trend' with supplemental information, such as COVID lock downs which reduced gasoline demand and the beginning of the war in Ukraine, which wreaked havoc on the supply and demand for global energy sources.}}
 

### 2022-07-05 2022-sdss-you-draw-it-manuscript.tex
Phrasing

@@
 {\textcolor{RedOrange}{We set out to implement}} `You Draw It', adapted
-from the New York Times feature, as a way to measure the patterns we see
-in data. Here, we provide technical details of the software development,
-utilizing interactive graphics in R\cite{r-software}. We then share
-results from our study which validates `You Draw It' as a method for
-graphical testing {\textcolor{RedOrange}{on linear trends}} and apply an
-appropriate data analysis method to the participant data.
+from the New York Times feature, as a way to experimentally assess the
+patterns we see in data. Here, we provide technical details of the
+software development, utilizing interactive graphics in
+R\cite{r-software}. We then share results from our study which validates
+`You Draw It' as a method for graphical testing
+{\textcolor{RedOrange}{on linear trends}} and apply an appropriate data
+analysis method to the participant data.
 {\textcolor{RedOrange}{We also briefly demonstrate the use of the 'You Draw It' method and analysis on nonlinear data.}}

### 2022-07-05 2022-sdss-you-draw-it-manuscript.qmd
Switch to Quarto, single column format, with default references. Add an edits.tex file that can be easily transferred to other docs to add the commands for inline editing.

--- 2022-07-05 2022-sdss-you-draw-it-manuscript.Rmd
@@
 keywords: 'graphics, user interaction, regression'
-output: 
-  pdf_document:
-    keep_tex: true
+format: 
+  pdf:
+    include-in-header:
+      - edits.tex
+      - refs.tex
+    keep-tex: true
     # template: template-blinded.tex
-    template: template.tex
-header-includes:
-   - \usepackage[dvipsnames]{xcolor} % colors
-   - \newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
-   - \newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
-   - \newcommand{\rh}[1]{{\textcolor{Green}{#1}}}
-   - \usepackage[capitalise]{cleveref}
-   - \newcommand\pcref[1]{(\cref{#1})}
-
+    # template: template.tex
 ---
@@
 
-\section{Introduction}
+# Introduction
 
@@
 <!-- All of these types of tests require different levels of use and manipulation of the information being presented in the chart. -->
-<!-- Efforts in the field of statistical graphics have developed graphical testing tools and methods, such as the lineup protocol \cite{buja2009statistical}.  -->
+<!-- Efforts in the field of statistical graphics have developed graphical testing tools and methods, such as the lineup protocol @buja2009statistical.  -->
 <!-- The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics. -->
@@
 
-\subsection{Measuring Patterns and Trends}
 \svp{One of the most common charts created is a scatter plot of points over time; these charts show up regularly in news articles and in scientific publications alike.
@@
 Our visual system is naturally built to look for structure and identify patterns, \svp{including patterns and trends over time; many times we do not even notice this process happening subconsciously}.
-\svp{As shown in \cref{fig:gas-prices}, a viewer engaging with a plot of weekly average gas prices over time may perform several cognitive operations.
+\svp{As shown in @fig:gas-prices, a viewer engaging with a plot of weekly average gas prices over time may perform several cognitive operations.
 First, they scan the plot and assesses points to determine if there are any outliers or otherwise remarkable points.
@@
 Additional mental operations a viewer might perform while looking at the plot are annotated in grey.
-\\cite{usenergyinformationadministrationWeeklyAllGrades2022}."
-#| out-width: "\\columnwidth"
+@usenergyinformationadministrationWeeklyAllGrades2022"
+#| out-width: \columnwidth
 #| label: gas-prices
@@
 
-Initial studies in the 20th century explored the use of fitting lines by eye through a set of points \cite{finney1951subjective,mosteller1981eye}.
+Initial studies in the 20th century explored the use of fitting lines by eye through a set of points [@finney1951subjective,mosteller1981eye].
 Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit was suitable, then drawing the line through the set of points.
-Recently, Ciccione and Dehaene \shortcite{ciccione2021can} conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.\svp{XXX and found what?
+Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.\svp{XXX and found what?
 Describe the results XXX}
 
-\svp{While psychologists and statisticians have been using eye-fitting techniques to assess our innate perceptual statistical modeling abilities, news organizations have used similar techniques in order to draw readers in and demonstrate the difference between readers' assumptions and reality.} In 2015, the New York Times introduced an interactive feature, called 'You Draw It' \cite{aisch2015you}, where readers input their own assumptions about various \svp{metrics of news interest} and compare these assumptions to reality.
+\svp{While psychologists and statisticians have been using eye-fitting techniques to assess our innate perceptual statistical modeling abilities, news organizations have used similar techniques in order to draw readers in and demonstrate the difference between readers' assumptions and reality.} In 2015, the New York Times introduced an interactive feature, called 'You Draw It' [@aisch2015you], where readers input their own assumptions about various \svp{metrics of news interest} and compare these assumptions to reality.
 The New York Times team used Data Driven Documents (D3), \svp{a JavaScript library} that allows readers to \svp{interact with a chart directly by} drawing a line on their computer screen with a mouse.
@@
 
-\svp{We set out to implement} ‘You Draw It’, adapted from the New York Times feature, as a way to measure the patterns we see in data. 
-Here, we provide technical details of the software development, utilizing interactive graphics in R\cite{r-software}. 
+\svp{We set out to implement} ‘You Draw It’, adapted from the New York Times feature, as a way to experimentally assess the patterns we see in data. 
+Here, we provide technical details of the software development, utilizing interactive graphics in R[@r-software]. 
 We then share results from our study which validates 'You Draw It' as a method for graphical testing \svp{on linear trends} and apply an appropriate data analysis method to the participant data.
@@
 
-\section{Development}
+# Development
 
-\subsection{`You Draw It' Task}
+## You Draw It v0
 
-When completing the graphical task, users are shown an interactive scatter-plot (Figure 1) along with the prompt, "Use your mouse to fill in the trend in the yellow box region." The yellow box region moves along as the user draws their trend-line, providing a visual cue which indicates where the user still needs to complete a trend line.
-After the entire domain has been visually estimated or predicted, the yellow region disappears, indicating the participant has completed the task.
-Prior to study participation, example gifs are shown and users are asked to complete practice plots to ease the learning curve of the task.
-^[Visit \url{emily-robinson.shinyapps.io/can-you-draw-it} for a test applet.]
+\svp{Describe the NYT code - what it does, how it works (for loop to connect mouse coordinates to closest point via inverting the scales)}
 
-<!-- # ```{r, you-draw-it-example, echo = F, out.width = "100%", fig.align='center', fig.cap = "'You Draw It' interaction task plot as shown to user.
-\\textbf{Left:} illustrates what user first sees with the prompt \\textit{`Use your mouse to fill in the trend in the yellow box region.'} \\textbf{Middle:} illustrates what the user sees while completing the task.
-\\textbf{Right:} illustrates the users finished trend line."} -->
-<!-- # knitr::include_graphics("images/ydi-stimuli.png") -->
-<!-- # ``` -->
+\svp{The New York Times uses D3, Data Driven Documents [@bostock2011d3], to create many of the interactive data graphics that users rely on, from the famous "election needle" in 2016 to their covid dashboards.
+Naturally, this same framework was used} to create the 'You Draw It' \svp{feature}.
+<!-- Major news and research organizations such as the New York Times, FiveThirtyEight, Washington Post, and the pew Research Center use D3 to create and customize graphics.  -->
 
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/ydi-stimuli}}
-\caption{You Draw It' task plot as shown to user. \\
-\textbf{left:} illustrates what user first sees with the prompt \textit{Use your mouse to fill in the trend in the yellow box region.} \\
-\textbf{middle:} illustrates what the user sees while completing the task.\\
-\textbf{right:} illustrates the users finished trend line.}
-\label{you-draw-it-task-plot}
-\end{center}
-\end{figure}
+A challenge of working with D3 is the environment necessary to display the graphics and images.
+\svp{XXX What environment, why is this a challenge, and how did you overcome it?}
 
-\subsection{Code Sketch}
+## Generating D3 Plots in R
 
-Using Shiny \cite{shinypkg} and JavaScript, we modified the New York Times 'You Draw It' feature for the purpose of testing statistical graphics allowing us to incorporate user interaction, conduct studies online, and store participant responses.
+\svp{Describe the process of using `r2d3`}
+<!-- The `r2d3` package [@r2d3pkg] provides an efficient integration of D3 visuals into R HTML formats.
+We integrate the D3 visual source code into an R Shiny [@shinypkg] application in order to allow for user interaction and data collection.
+-->
+\svp{We leveraged the `r2d3` package [@r2d3pkg] to take data randomly generated in R and create a D3 plot that could be modified with the You Draw It script.
+This step allowed us to more easily connect the data generation process with the resulting D3 code without having to generate new code manually each time we re-generated the data.
+While not strictly necessary for the validation test of the You Draw It method compared to past linear regression methods presented in this paper, this abstraction makes it much easier to test arbitrary or model-generated data which is unique to each participant.}
 
-Data Driven Documents (D3) \cite{bostock2011d3}, a JavaScript-based graphing framework that facilitates user interaction, was used to create the 'You Draw It' visual.
-Major news and research organizations such as the New York Times, FiveThirtyEight, Washington Post, and the pew Research Center use D3 to create and customize graphics.
-A challenge of working with D3 is the environment necessary to display the graphics and images.
-The `r2d3` package \cite{r2d3pkg} provides an efficient integration of D3 visuals into R HTML formats.
-We integrate the D3 visual source code into an R Shiny \cite{shinypkg} application in order to allow for user interaction and data collection.
+We conducted all data simulation and processing in R and output two data sets - *point data* and *line data* - containing (x, y) coordinates corresponding to either a simulated point or fitted value predicted by a statistical model respectively.
+Then, the `r2d3` package converts the data sets in R to JavaScript Object Notation (JSON) to be interpreted by JavaScript code included with the `r2d3` package.
+Parameters for aesthetic design choices are defined in a list of options in the `r2d3` function call; `r2d3` passes these to the generated JavaScript code. 
+For instance, we can specify the buffer space allowed for the $x$ and $y$ axes to avoid users \svp{anchoring} their lines to the axes limits.
 
-Figure 2 illustrates the initial setup of the visual stimuli along with the iterative process between the user interaction and plotting.
-We conducted all data simulation and processing in R and output two data sets - *point data* and *line data* - containing (x, y) coordinates corresponding to either a simulated point or fitted value predicted by a statistical model respectively.
-Then, the r2d3 package converts the data sets in R to JavaScript Object Notation (JSON) to be interpreted by the `D3.js` code.
-We define functions in `D3.js` to draw the initial plot and set up drawable points for the user drawn line.
-Drag events in `D3.js` are utilized to observe and react to user input.
-Shiny Messages are used to communicate the user interaction between the `D3.js` code and the R environment.
-The plot is then rendered and updated on user interaction into the R shiny application with the `RenderD3` and `d3Output` functions.
-Once the user is done drawing the line, we saved the results of the drawn line to a database for analysis.
+## Modifying You Draw It
+\svp{
+In order to make use of the original You Draw It code for perceptual experiments, we first needed to accommodate an additional layer.
+The original NYT features asked participants to draw on a blank coordinate grid, as shown in @fig-nyt-screenshot.
+}
 
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/code-sketch-2}}
-\caption{Sketch of underlying code for 'You Draw It', illustrating the data simulation conducted in R, the initial setup of the visual stimuli with D3 source code, along with the iterative process between the user interaction and plotting in Shiny.
-Once the user is done drawing the line, we saved the results of the drawn line to a database for analysis.}
-\label{you-draw-it-code-sketch}
-\end{center}
-\end{figure}
+![Screenshot of the [You Draw It application](https://www.nytimes.com/interactive/2015/05/28/upshot/you-draw-it-how-family-income-affects-childrens-college-chances.html) originally developed by the New York Times.](images/NYT-You-Draw-It-Screenshot.png){#fig-nyt-screenshot
+width='\columnwidth'}
 
-Parameters for aesthetic design choices are defined in a list of options and `r2d3` passes these to the D3.js code.
-For instance, we can specify the buffer space allowed for the $x$ and $y$ axes to avoid users to anchor their lines to the axes limits.
-^[For `D3.js` source code, visit \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.]
+\svp{When testing perception of graphics, however, we want to pre-populate the chart with additional information - points, and in some cases, portions of a trend line.
+This requires that we modify the original JavaScript code to accommodate these additional elements, which we generate using `r2d3`, as described above.}
 
-\subsection{Challenges}
+We defined JavaScript functions to draw the initial plot and set up drawable points for the user drawn line.
+Drag events in `D3.js` are utilized to observe and react to user input, \svp{as in the original You Draw It script from the NYT}.
+^[The full source code we developed for the YouDrawIt task is available at \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.]
 
-During the development process, there were a few challenges we had to overcome. 
-We documented these challenges with GitHub commits and outline places where frustration occurred.
-Data Driven documents uses Scalable Vector Graphics (SVG), thus requiring careful transformation between the pixels and plot coordinates to align the simulated data and the user drawn line appropriately.
-With the layered framework of SVG's, it was important to place the layers in the right order so certain features would appear where desired. 
-For example, we wanted the opacity of the yellow box region to appear under the points while still showing the grid lines.
-As previously mentioned, `r2d3` automatically converts the data set to a JSON file to be interpreted by the underlying source code, however we provided two data sets (point data and line data) and converted the two data sets to a JSON file before passing them to the `r2d3` argument so that we could reference both data sets within the `D3.js` code.
-One constraint of the development code to date is that it is only built to work with one-to-one functions and does not allow for multiple estimates or predictions for one $x$-value.
+<!-- Note, we didn't modify d3.js directly, we just added additional functions that would interpret our JSON file and draw the appropriate plot.
+These functions were modified from the defaults automatically generated by `r2d3` -->
 
-\section{Application}
+\svp{One constraint we inherited from the NYT code is that users can only draw one-to-one functions; because the code works with drag events, each point in $x$ can correspond to only one point in $y$, at least as the code is currently written.
+While this is not particularly problematic for our applications to date, it might limit applications of this type of user interaction when assessing user drawn confidence bands, ribbons, and other situations where two or more vertical points are required.}
+<!-- \svp{XXX additional details of how we accomplished this XXX} -->
 
-\subsection{Tool Validation}
+## Visual Cues
 
-We conducted a study in order to validate 'You Draw It' as a method for graphical testing, comparing results to the less technological method utilized in Mosteller et al.
-\shortcite{mosteller1981eye}.
+\svp{During testing of our modifications to the You Draw It script, we discovered that frequently the JavaScript code would ``skip" from point to point out of sequence.
+This resulted in a jagged line (visually) with missing values in the underlying stored array of data; as a result, the for-loop controlling the user-drawn line never exited and the user's data was not recorded.
+While it would have been possible to fix this using some sort of interpolation algorithm, we did not want to compromise user results by introducing interpolation artifacts, so we opted instead for a visual cue that fixed the problem from a psychological angle.}
+
+![You Draw It task plot as shown to user.    
+**left:** Initial state, with instructions *Use your mouse to fill in the trend in the yellow box region.*    
+**middle:** User view during task completion.    
+**right:** Finished state](images/ydi-stimuli.png){#fig-you-draw-it-task-plot}
+
+As shown in @fig-you-draw-it-task-plot, the user-filled portion of the plot is represented with a yellow rectangle, which adapts to the user's input and spans any missing values.
+When the box disappears, the array is filled in and the user can submit their response.
+
+\svp{One challenge introduced by adding this visual cue is that D3 uses Scalable Vector Graphics (SVG) to render elements.
+Adding an additional layer meant that we had to ensure that not only the grids, lines, and points of the default scatter plot and trend line rendered in the correct order, but that the yellow box rendered behind all of these points as well so that no information was masked.}
+
+## Connecting to Shiny
+Shiny Messages are used to communicate the user interaction between the JavaScript code and the R environment. 
+\svp{The initial plot is rendered using Shiny's `RenderD3` and `d3Output` functions; subsequent user interactions are controlled via the JavaScript code.
+Once the user is finished modifying the plot, they can submit their response, so long as they have filled in all of the points.
+This is enforced using Shiny's message-passing interface and a hook in the JavaScript code that notifies Shiny when the array of user-drawn points is completely filled in.}
+<!-- The plot is then rendered and updated on user interaction into the R shiny application with the `RenderD3` and `d3Output` functions. -->
+Once the user is done drawing the line, we save the results of the drawn line to a SQLite database, using Shiny to pass data from JavaScript to R. 
+The connections between each portion of the app are shown in @fig-you-draw-it-code-sketch.
+
+![Sketch of underlying code for 'You Draw It', illustrating the data simulation conducted in R, the initial setup of the visual stimuli with D3 source code, along with the iterative process between the user interaction and plotting in Shiny.
+Once the user is done drawing the line, we saved the results of the drawn line to a database for analysis.](images/code-sketch-2.png){#fig-you-draw-it-code-sketch}
+
+\svp{One additional challenge when integrating the D3 graphics into the Shiny application is that by default, most Shiny applications use a reactive framework that adjusts to the user's browser size.
+When working with D3, however, this can be problematic: most D3 parameters are specified at the pixel level.
+The discontinuity between the assumptions of Shiny and D3 meant that we had to fix the size of the Shiny element, and then perform a check to ensure that the user's screen was sufficiently big to render the plot.
+While in an ideal world, users could participate using cell phones, tablets, and traditional laptop/desktop computers, functionally our checks limited users' ability to participate using smaller screen sizes found in cell phones and some tablets.
+}
+
+\svp{Additionally, after some feedback by laptop users during pilot testing, we included an additional requirement that participants have a computer mouse available; the results from touchpad users were qualitatively different (more jagged) in ways suggesting that the recorded data did not adequately describe users' perceptions.}
+
+# Application
+
+## Tool Validation
+
+We conducted a study in order to validate 'You Draw It' as a method for graphical testing, comparing results to the less technological method utilized in @mosteller1981eye.
 Results from our study were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the ordinary least-squares regression line (Figure 3).
@@
 
-\subsection{Data Analysis}
+## Data Analysis
 
@@
 
-\section{Future Work}
+# Future Work
 

### 2022-07-05 2022-sdss-you-draw-it-manuscript.tex
Switch to Quarto, single column format, with default references. Add an edits.tex file that can be easily transferred to other docs to add the commands for inline editing.

--- 2022-07-05 2022-sdss-you-draw-it-manuscript.tex
@@
-\documentclass[10pt]{article}
+% Options for packages loaded elsewhere
+\PassOptionsToPackage{unicode}{hyperref}
+\PassOptionsToPackage{hyphens}{url}
+\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
+%
+\documentclass[
+  letterpaper,
+  DIV=11,
+  numbers=noendperiod]{scrartcl}
 
-\usepackage{sdss2020} % Uses Times Roman font (either newtx or times package)
-\usepackage{url}
-\usepackage{latexsym}
-\usepackage{amsmath, amsthm, amsfonts}
-\usepackage{algorithm, algorithmic}  
+\usepackage{amsmath,amssymb}
+\usepackage{lmodern}
+\usepackage{iftex}
+\ifPDFTeX
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+  \usepackage{textcomp} % provide euro and other symbols
+\else % if luatex or xetex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Scale=MatchLowercase}
+  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+% Use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+\IfFileExists{microtype.sty}{% use microtype if available
+  \usepackage[]{microtype}
+  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\makeatletter
+\@ifundefined{KOMAClassName}{% if non-KOMA class
+  \IfFileExists{parskip.sty}{%
+    \usepackage{parskip}
+  }{% else
+    \setlength{\parindent}{0pt}
+    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
+}{% if KOMA class
+  \KOMAoptions{parskip=half}}
+\makeatother
+\usepackage{xcolor}
+\setlength{\emergencystretch}{3em} % prevent overfull lines
+\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
+% Make \paragraph and \subparagraph free-standing
+\ifx\paragraph\undefined\else
+  \let\oldparagraph\paragraph
+  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
+\fi
+\ifx\subparagraph\undefined\else
+  \let\oldsubparagraph\subparagraph
+  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
+\fi
+
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
 \usepackage{graphicx}
+\makeatletter
+\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
+\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
+\makeatother
+% Scale images if necessary, so that they will not overflow the page
+% margins by default, and it is still possible to overwrite the defaults
+% using explicit options in \includegraphics[width, height, ...]{}
+\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
+% Set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 \newcommand{\rh}[1]{{\textcolor{Green}{#1}}}
-\usepackage[capitalise]{cleveref}
-\newcommand\pcref[1]{(\cref{#1})}
+% \usepackage[capitalise]{cleveref}
+% \newcommand\pcref[1]{(\cref{#1})}
+
+\usepackage{natbib}
+\bibliographystyle{unsrtnat}
+\KOMAoption{captions}{tableheading}
+\makeatletter
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\AtBeginDocument{%
+\ifdefined\contentsname
+  \renewcommand*\contentsname{Table of contents}
+\else
+  \newcommand\contentsname{Table of contents}
+\fi
+\ifdefined\listfigurename
+  \renewcommand*\listfigurename{List of Figures}
+\else
+  \newcommand\listfigurename{List of Figures}
+\fi
+\ifdefined\listtablename
+  \renewcommand*\listtablename{List of Tables}
+\else
+  \newcommand\listtablename{List of Tables}
+\fi
+\ifdefined\figurename
+  \renewcommand*\figurename{Figure}
+\else
+  \newcommand\figurename{Figure}
+\fi
+\ifdefined\tablename
+  \renewcommand*\tablename{Table}
+\else
+  \newcommand\tablename{Table}
+\fi
+}
+\@ifpackageloaded{float}{}{\usepackage{float}}
+\floatstyle{ruled}
+\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
+\floatname{codelisting}{Listing}
+\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
+\makeatother
+\makeatletter
+\@ifpackageloaded{caption}{}{\usepackage{caption}}
+\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
+\makeatother
+\makeatletter
+\@ifpackageloaded{tcolorbox}{}{\usepackage[many]{tcolorbox}}
+\makeatother
+\makeatletter
+\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
+\makeatother
+\makeatletter
+\makeatother
+\ifLuaTeX
+  \usepackage{selnolig}  % disable illegal ligatures
+\fi
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={`You Draw It': Implementation of visually fitted trends with r2d3},
+  pdfkeywords={graphics, user interaction, regression},
+  colorlinks=true,
+  linkcolor={blue},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={Blue},
+  pdfcreator={LaTeX via pandoc}}
 
@@
 \texttt{r2d3}}
-
-%\author{by }
-\author{
-  Emily A. Robinson \\
-  \small{University of Nebraska - Lincoln}\\
-  \small{Lincoln, NE} \\
-  \small{\tt emily.robinson@huskers.unl.edu} \\\And
- Reka Howard \\
-  \small{University of Nebraska - Lincoln}\\
-  \small{Lincoln, NE} \\
-  \small{\tt rekahoward@unl.edu} \\\And
-  Susan VanderPlas \\
-  \small{University of Nebraska - Lincoln}\\
-  \small{Lincoln, NE} \\
-  \small{\tt susan.vanderplas@unl.edu} \\}
-
+\author{}
 \date{}
@@
 \end{abstract}
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
-{\bf Keywords:} graphics, user interaction, regression
-
-\section{Introduction}
+\hypertarget{introduction}{%
+\section{Introduction}\label{introduction}}
 
 We all use statistical graphics, but how do we know that the graphics we
-use are communicating {\textcolor{RedOrange}{effectively}}? Through
-experimentation, graphical testing methods allow researchers to conduct
-studies
-{\textcolor{RedOrange}{designed to understand how we perceive graphics and perform graphical tasks}}
+use are communicating \svp{effectively}? Through experimentation,
+graphical testing methods allow researchers to conduct studies
+\svp{designed to understand how we perceive graphics and perform graphical tasks}
 such as differentiation, prediction, estimation, and extrapolation.
-{\textcolor{RedOrange}{Each of these levels of interaction with a graph require a different method of engagement with and use of the information presented in a chart.}}
-{\textcolor{RedOrange}{In this paper, we describe the adaptation of an old tool for graphical testing and evaluation, eye-fitting, for use in modern web-applications suitable for testing statistical graphics.
-We present an empirical evaluation of this testing method for linear regression, and briefly discuss an extension of this method to non-linear applications.}}
+\svp{Each of these levels of interaction with a graph require a different method of engagement with and use of the information presented in a chart.}
+\svp{In this paper, we describe the adaptation of an old tool for graphical testing and evaluation, eye-fitting, for use in modern web-applications suitable for testing statistical graphics. 
+We present an empirical evaluation of this testing method for linear regression, and briefly discuss an extension of this method to non-linear applications.}
 
-\subsection{Measuring Patterns and Trends}
-
-{\textcolor{RedOrange}{One of the most common charts created is a scatter plot of points over time; these charts show up regularly in news articles and in scientific publications alike.
-These charts rely on our }}ability to identify and detect trends in
-data. Our visual system is naturally built to look for structure and
-identify patterns,
-{\textcolor{RedOrange}{including patterns and trends over time}}.
-{\textcolor{RedOrange}{As shown in \cref{fig:gas-prices}, a viewer engaging with a plot of weekly average gas prices over time will likely perform several cognitive operations.
+\svp{One of the most common charts created is a scatter plot of points over time; these charts show up regularly in news articles and in scientific publications alike.
+These charts rely on our} ability to identify and detect trends in data.
+Our visual system is naturally built to look for structure and identify
+patterns,
+\svp{including patterns and trends over time; many times we do not even notice this process happening subconsciously}.
+\svp{As shown in @fig:gas-prices, a viewer engaging with a plot of weekly average gas prices over time may perform several cognitive operations.
 First, they scan the plot and assesses points to determine if there are any outliers or otherwise remarkable points.
 Then, they may fit a rough mental smooth/trend line to the points to summarize the useful information and remove variability.
-Finally, the viewer may pull in additional contextual information from long-term memory, seeking to explain variation in the mental 'trend' with supplemental information, such as COVID lock downs which reduced gasoline demand and the beginning of the war in Ukraine, which wreaked havoc on the supply and demand for global energy sources.}}
+Finally, an interested and engaged viewer may pull in additional contextual information from long-term memory, seeking to explain variation in the mental 'trend' with supplemental information, such as COVID lock downs which reduced gasoline demand and the beginning of the war in Ukraine, which wreaked havoc on the supply and demand for global energy sources.}
 
@@
 
-{\centering \includegraphics[width=\columnwidth]{./images/gas-prices-1}
+{\centering \includegraphics{./images/gas-prices-1.pdf}
 
@@
 \caption{Weekly average gas prices in the United States, 2019-June 2022.
-Additional mental operations a viewer might perform while looking at the plot are annotated in grey.
-\cite{usenergyinformationadministrationWeeklyAllGrades2022}.}\label{fig:gas-prices}
+Additional mental operations a viewer might perform while looking at the
+plot are annotated in grey. US Energy Information Administration (2022)}
+
 \end{figure}
@@
 Initial studies in the 20th century explored the use of fitting lines by
-eye through a set of points
-\cite{finney1951subjective,mosteller1981eye}. Common methods of fitting
-trends by eye involved maneuvering a string, black thread, or ruler
-until the fit was suitable, then drawing the line through the set of
-points. Recently, Ciccione and Dehaene \shortcite{ciccione2021can}
+eye through a set of points (Finney 1951, mosteller1981eye). Common
+methods of fitting trends by eye involved maneuvering a string, black
+thread, or ruler until the fit was suitable, then drawing the line
+through the set of points. Recently, Ciccione and Dehaene (2021)
 conducted a comprehensive set of studies investigating human ability to
 detect trends in graphical representations from a psychophysical
-approach.{\textcolor{RedOrange}{XXX and found what? Describe the results XXX}}
+approach.\svp{XXX and found what? Describe the results XXX}
 
-{\textcolor{RedOrange}{While psychologists and statisticians have been using eye-fitting techniques to assess our innate perceptual statistical modeling abilities, news organizations have used similar techniques in order to draw readers in and demonstrate the difference between readers' assumptions and reality.}}
+\svp{While psychologists and statisticians have been using eye-fitting techniques to assess our innate perceptual statistical modeling abilities, news organizations have used similar techniques in order to draw readers in and demonstrate the difference between readers' assumptions and reality.}
 In 2015, the New York Times introduced an interactive feature, called
-`You Draw It' \cite{aisch2015you}, where readers input their own
-assumptions about various
-{\textcolor{RedOrange}{metrics of news interest}} and compare these
-assumptions to reality. The New York Times team used Data Driven
-Documents (D3), {\textcolor{RedOrange}{a JavaScript library}} that
-allows readers to
-{\textcolor{RedOrange}{interact with a chart directly by}} drawing a
-line on their computer screen with a mouse.
-{\textcolor{RedOrange}{Despite the somewhat different purpose behind this feature, the D3 driven method used by the NYT is wonderfully intuitive, and does not require the assumption of linearity, making it much more adaptable to testing how viewers perceive and predict when presented with non-linear data.}}
+`You Draw It' (Aisch, Cox, and Quealy 2015), where readers input their
+own assumptions about various \svp{metrics of news interest} and compare
+these assumptions to reality. The New York Times team used Data Driven
+Documents (D3), \svp{a JavaScript library} that allows readers to
+\svp{interact with a chart directly by} drawing a line on their computer
+screen with a mouse.
+\svp{Despite the somewhat different purpose behind this feature, the D3 driven method used by the NYT is wonderfully intuitive, and does not require the assumption of linearity, making it much more adaptable to testing how viewers perceive and predict when presented with non-linear data.}
 
-{\textcolor{RedOrange}{We set out to implement}} `You Draw It', adapted
-from the New York Times feature, as a way to measure the patterns we see
-in data. Here, we provide technical details of the software development,
-utilizing interactive graphics in R\cite{r-software}. We then share
+\svp{We set out to implement} `You Draw It', adapted from the New York
+Times feature, as a way to experimentally assess the patterns we see in
+data. Here, we provide technical details of the software development,
+utilizing interactive graphics in R(R Core Team 2022). We then share
 results from our study which validates `You Draw It' as a method for
-graphical testing {\textcolor{RedOrange}{on linear trends}} and apply an
-appropriate data analysis method to the participant data.
-{\textcolor{RedOrange}{We also briefly demonstrate the use of the 'You Draw It' method and analysis on nonlinear data.}}
+graphical testing \svp{on linear trends} and apply an appropriate data
+analysis method to the participant data.
+\svp{We also briefly demonstrate the use of the 'You Draw It' method and analysis on nonlinear data.}
 
-\section{Development}
+\hypertarget{development}{%
+\section{Development}\label{development}}
 
-\subsection{`You Draw It' Task}
+\hypertarget{you-draw-it-v0}{%
+\subsection{You Draw It v0}\label{you-draw-it-v0}}
 
-When completing the graphical task, users are shown an interactive
-scatter-plot (Figure 1) along with the prompt, ``Use your mouse to fill
-in the trend in the yellow box region.'' The yellow box region moves
-along as the user draws their trend-line, providing a visual cue which
-indicates where the user still needs to complete a trend line. After the
-entire domain has been visually estimated or predicted, the yellow
-region disappears, indicating the participant has completed the task.
-Prior to study participation, example gifs are shown and users are asked
-to complete practice plots to ease the learning curve of the task.
-\footnote{Visit \url{emily-robinson.shinyapps.io/can-you-draw-it} for a
-  test applet.}
+\svp{Describe the NYT code - what it does, how it works (for loop to connect mouse coordinates to closest point via inverting the scales)}
 
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/ydi-stimuli}}
-\caption{You Draw It' task plot as shown to user. \\
-\textbf{left:} illustrates what user first sees with the prompt \textit{Use your mouse to fill in the trend in the yellow box region.} \\
-\textbf{middle:} illustrates what the user sees while completing the task.\\
-\textbf{right:} illustrates the users finished trend line.}
-\label{you-draw-it-task-plot}
-\end{center}
+\svp{The New York Times uses D3, Data Driven Documents [@bostock2011d3], to create many of the interactive data graphics that users rely on, from the famous "election needle" in 2016 to their covid dashboards.
+Naturally, this same framework was used}
+to create the `You Draw It' \svp{feature}.
+
+A challenge of working with D3 is the environment necessary to display
+the graphics and images.
+\svp{XXX What environment, why is this a challenge, and how did you overcome it?}
+
+\hypertarget{generating-d3-plots-in-r}{%
+\subsection{Generating D3 Plots in R}\label{generating-d3-plots-in-r}}
+
+\svp{Describe the process of using `r2d3`}
+
+\svp{We leveraged the `r2d3` package [@r2d3pkg] to take data randomly generated in R and create a D3 plot that could be modified with the You Draw It script.
+This step allowed us to more easily connect the data generation process with the resulting D3 code without having to generate new code manually each time we re-generated the data.
+While not strictly necessary for the validation test of the You Draw It method compared to past linear regression methods presented in this paper, this abstraction makes it much easier to test arbitrary or model-generated data which is unique to each participant.}
+
+We conducted all data simulation and processing in R and output two data
+sets - \emph{point data} and \emph{line data} - containing (x, y)
+coordinates corresponding to either a simulated point or fitted value
+predicted by a statistical model respectively. Then, the \texttt{r2d3}
+package converts the data sets in R to JavaScript Object Notation (JSON)
+to be interpreted by JavaScript code included with the \texttt{r2d3}
+package. Parameters for aesthetic design choices are defined in a list
+of options in the \texttt{r2d3} function call; \texttt{r2d3} passes
+these to the generated JavaScript code. For instance, we can specify the
+buffer space allowed for the \(x\) and \(y\) axes to avoid users
+\svp{anchoring} their lines to the axes limits.
+
+\hypertarget{modifying-you-draw-it}{%
+\subsection{Modifying You Draw It}\label{modifying-you-draw-it}}
+
+\svp{
+In order to make use of the original You Draw It code for perceptual experiments, we first needed to accommodate an additional layer.
+The original NYT features asked participants to draw on a blank coordinate grid, as shown in @fig-nyt-screenshot.
+}
+
+\begin{figure}
+
+{\centering \includegraphics{images/NYT-You-Draw-It-Screenshot.png}
+
+}
+
+\caption{\label{fig-nyt-screenshot}Screenshot of the
+\href{https://www.nytimes.com/interactive/2015/05/28/upshot/you-draw-it-how-family-income-affects-childrens-college-chances.html}{You
+Draw It application} originally developed by the New York Times.}
+
 \end{figure}
 
-\subsection{Code Sketch}
+\svp{When testing perception of graphics, however, we want to pre-populate the chart with additional information - points, and in some cases, portions of a trend line.
+This requires that we modify the original JavaScript code to accommodate these additional elements, which we generate using `r2d3`, as described above.}
 
-Using Shiny \cite{shinypkg} and JavaScript, we modified the New York
-Times `You Draw It' feature for the purpose of testing statistical
-graphics allowing us to incorporate user interaction, conduct studies
-online, and store participant responses.
+We defined JavaScript functions to draw the initial plot and set up
+drawable points for the user drawn line. Drag events in \texttt{D3.js}
+are utilized to observe and react to user input,
+\svp{as in the original You Draw It script from the NYT}. \footnote{The
+  full source code we developed for the YouDrawIt task is available at
+  \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.}
 
-Data Driven Documents (D3) \cite{bostock2011d3}, a JavaScript-based
-graphing framework that facilitates user interaction, was used to create
-the `You Draw It' visual. Major news and research organizations such as
-the New York Times, FiveThirtyEight, Washington Post, and the pew
-Research Center use D3 to create and customize graphics. A challenge of
-working with D3 is the environment necessary to display the graphics and
-images. The \texttt{r2d3} package \cite{r2d3pkg} provides an efficient
-integration of D3 visuals into R HTML formats. We integrate the D3
-visual source code into an R Shiny \cite{shinypkg} application in order
-to allow for user interaction and data collection.
+\svp{One constraint we inherited from the NYT code is that users can only draw one-to-one functions; because the code works with drag events, each point in $x$ can correspond to only one point in $y$, at least as the code is currently written.
+While this is not particularly problematic for our applications to date, it might limit applications of this type of user interaction when assessing user drawn confidence bands, ribbons, and other situations where two or more vertical points are required.}
 
-Figure 2 illustrates the initial setup of the visual stimuli along with
-the iterative process between the user interaction and plotting. We
-conducted all data simulation and processing in R and output two data
-sets - \emph{point data} and \emph{line data} - containing (x, y)
-coordinates corresponding to either a simulated point or fitted value
-predicted by a statistical model respectively. Then, the r2d3 package
-converts the data sets in R to JavaScript Object Notation (JSON) to be
-interpreted by the \texttt{D3.js} code. We define functions in
-\texttt{D3.js} to draw the initial plot and set up drawable points for
-the user drawn line. Drag events in \texttt{D3.js} are utilized to
-observe and react to user input. Shiny Messages are used to communicate
-the user interaction between the \texttt{D3.js} code and the R
-environment. The plot is then rendered and updated on user interaction
-into the R shiny application with the \texttt{RenderD3} and
-\texttt{d3Output} functions. Once the user is done drawing the line, we
-saved the results of the drawn line to a database for analysis.
+\hypertarget{visual-cues}{%
+\subsection{Visual Cues}\label{visual-cues}}
 
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/code-sketch-2}}
-\caption{Sketch of underlying code for 'You Draw It', illustrating the data simulation conducted in R, the initial setup of the visual stimuli with D3 source code, along with the iterative process between the user interaction and plotting in Shiny.
-Once the user is done drawing the line, we saved the results of the drawn line to a database for analysis.}
-\label{you-draw-it-code-sketch}
-\end{center}
+\svp{During testing of our modifications to the You Draw It script, we discovered that frequently the JavaScript code would ``skip" from point to point out of sequence.
+This resulted in a jagged line (visually) with missing values in the underlying stored array of data; as a result, the for-loop controlling the user-drawn line never exited and the user's data was not recorded.
+While it would have been possible to fix this using some sort of interpolation algorithm, we did not want to compromise user results by introducing interpolation artifacts, so we opted instead for a visual cue that fixed the problem from a psychological angle.}
+
+\begin{figure}
+
+{\centering \includegraphics{images/ydi-stimuli.png}
+
+}
+
+\caption{\label{fig-you-draw-it-task-plot}You Draw It task plot as shown
+to user.\\
+\textbf{left:} Initial state, with instructions \emph{Use your mouse to
+fill in the trend in the yellow box region.}\\
+\textbf{middle:} User view during task completion.\\
+\textbf{right:} Finished state}
+
 \end{figure}
 
-Parameters for aesthetic design choices are defined in a list of options
-and \texttt{r2d3} passes these to the D3.js code. For instance, we can
-specify the buffer space allowed for the \(x\) and \(y\) axes to avoid
-users to anchor their lines to the axes limits. \footnote{For
-  \texttt{D3.js} source code, visit
-  \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.}
+As shown in Figure~\ref{fig-you-draw-it-task-plot}, the user-filled
+portion of the plot is represented with a yellow rectangle, which adapts
+to the user's input and spans any missing values. When the box
+disappears, the array is filled in and the user can submit their
+response.
 
-\subsection{Challenges}
+\svp{One challenge introduced by adding this visual cue is that D3 uses Scalable Vector Graphics (SVG) to render elements.
+Adding an additional layer meant that we had to ensure that not only the grids, lines, and points of the default scatter plot and trend line rendered in the correct order, but that the yellow box rendered behind all of these points as well so that no information was masked.}
 
-During the development process, there were a few challenges we had to
-overcome. We documented these challenges with GitHub commits and outline
-places where frustration occurred. Data Driven documents uses Scalable
-Vector Graphics (SVG), thus requiring careful transformation between the
-pixels and plot coordinates to align the simulated data and the user
-drawn line appropriately. With the layered framework of SVG's, it was
-important to place the layers in the right order so certain features
-would appear where desired. For example, we wanted the opacity of the
-yellow box region to appear under the points while still showing the
-grid lines. As previously mentioned, \texttt{r2d3} automatically
-converts the data set to a JSON file to be interpreted by the underlying
-source code, however we provided two data sets (point data and line
-data) and converted the two data sets to a JSON file before passing them
-to the \texttt{r2d3} argument so that we could reference both data sets
-within the \texttt{D3.js} code. One constraint of the development code
-to date is that it is only built to work with one-to-one functions and
-does not allow for multiple estimates or predictions for one
-\(x\)-value.
+\hypertarget{connecting-to-shiny}{%
+\subsection{Connecting to Shiny}\label{connecting-to-shiny}}
 
-\section{Application}
+Shiny Messages are used to communicate the user interaction between the
+JavaScript code and the R environment.
+\svp{The initial plot is rendered using Shiny's `RenderD3` and `d3Output` functions; subsequent user interactions are controlled via the JavaScript code.
+Once the user is finished modifying the plot, they can submit their response, so long as they have filled in all of the points.
+This is enforced using Shiny's message-passing interface and a hook in the JavaScript code that notifies Shiny when the array of user-drawn points is completely filled in.}
+Once the user is done drawing the line, we save the results of the drawn
+line to a SQLite database, using Shiny to pass data from JavaScript to
+R. The connections between each portion of the app are shown in
+Figure~\ref{fig-you-draw-it-code-sketch}.
 
-\subsection{Tool Validation}
+\begin{figure}
+
+{\centering \includegraphics{images/code-sketch-2.png}
+
+}
+
+\caption{\label{fig-you-draw-it-code-sketch}Sketch of underlying code
+for `You Draw It', illustrating the data simulation conducted in R, the
+initial setup of the visual stimuli with D3 source code, along with the
+iterative process between the user interaction and plotting in Shiny.
+Once the user is done drawing the line, we saved the results of the
+drawn line to a database for analysis.}
+
+\end{figure}
+
+\svp{One additional challenge when integrating the D3 graphics into the Shiny application is that by default, most Shiny applications use a reactive framework that adjusts to the user's browser size.
+When working with D3, however, this can be problematic: most D3 parameters are specified at the pixel level.
+The discontinuity between the assumptions of Shiny and D3 meant that we had to fix the size of the Shiny element, and then perform a check to ensure that the user's screen was sufficiently big to render the plot.
+While in an ideal world, users could participate using cell phones, tablets, and traditional laptop/desktop computers, functionally our checks limited users' ability to participate using smaller screen sizes found in cell phones and some tablets.
+}
+
+\svp{Additionally, after some feedback by laptop users during pilot testing, we included an additional requirement that participants have a computer mouse available; the results from touchpad users were qualitatively different (more jagged) in ways suggesting that the recorded data did not adequately describe users' perceptions.}
+
+\hypertarget{application}{%
+\section{Application}\label{application}}
+
+\hypertarget{tool-validation}{%
+\subsection{Tool Validation}\label{tool-validation}}
 
@@
 graphical testing, comparing results to the less technological method
-utilized in Mosteller et al.~\shortcite{mosteller1981eye}. Results from
-our study were consistent with those found in the previous study; when
-shown points following a linear trend, participants tended to fit the
-slope of the first principal component over the slope of the ordinary
-least-squares regression line (Figure 3). This trend was most prominent
-when shown data simulated with larger variances. This study reinforced
-the differences between intuitive visual model fitting and statistical
-model fitting, providing information about human perception as it
-relates to the use of statistical graphics.
+utilized in Mosteller et al. (1981). Results from our study were
+consistent with those found in the previous study; when shown points
+following a linear trend, participants tended to fit the slope of the
+first principal component over the slope of the ordinary least-squares
+regression line (Figure 3). This trend was most prominent when shown
+data simulated with larger variances. This study reinforced the
+differences between intuitive visual model fitting and statistical model
+fitting, providing information about human perception as it relates to
+the use of statistical graphics.
 
@@
 
-\subsection{Data Analysis}
+\hypertarget{data-analysis}{%
+\subsection{Data Analysis}\label{data-analysis}}
 
@@
 
-\section{Future Work}
+\hypertarget{future-work}{%
+\section{Future Work}\label{future-work}}
 
@@
 
-\bibliographystyle{sdss2020}
-\bibliography{references.bib}
-% \bibliography{references}
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-aisch2015you}{}}%
+Aisch, Gregor, Amanda Cox, and Kevin Quealy. 2015. {``You Draw It: How
+Family Income Predicts Children's College Chances.''} \emph{The New York
+Times} 28.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ciccione2021can}{}}%
+Ciccione, Lorenzo, and Stanislas Dehaene. 2021. {``Can Humans Perform
+Mental Regression on a Graph? Accuracy and Bias in the Perception of
+Scatterplots.''} \emph{Cognitive Psychology} 128: 101406.
+
+\leavevmode\vadjust pre{\hypertarget{ref-finney1951subjective}{}}%
+Finney, DJ. 1951. {``Subjective Judgment in Statistical Analysis: An
+Experimental Study.''} \emph{Journal of the Royal Statistical Society:
+Series B (Methodological)} 13 (2): 284--97.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mosteller1981eye}{}}%
+Mosteller, Frederick, Andrew F Siegel, Edward Trapido, and Cleo Youtz.
+1981. {``Eye Fitting Straight Lines.''} \emph{The American Statistician}
+35 (3): 150--52.
+
+\leavevmode\vadjust pre{\hypertarget{ref-r-software}{}}%
+R Core Team. 2022. \emph{R: A Language and Environment for Statistical
+Computing}. Vienna, Austria: R Foundation for Statistical Computing.
+\url{https://www.R-project.org/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-usenergyinformationadministrationWeeklyAllGrades2022}{}}%
+US Energy Information Administration. 2022. {``Weekly {U}.{S}. {All}
+{Grades} {All} {Formulations} {Retail} {Gasoline} {Prices} ({Dollars}
+Per {Gallon}).''} \emph{US Energy Information Administration Independent
+Statistics \& Analysis}.
+\url{https://www.eia.gov/dnav/pet/hist/LeafHandler.ashx?n=PET\&s=EMM_EPM0_PTE_NUS_DPG\&f=W}.
+
+\end{CSLReferences}
 

### 2022-07-05 2022-sdss-you-draw-it-manuscript.qmd
A couple more comments

@@
 
+\svp{Describe the experiment formally.
+Don't be afraid to quote Mosteller's paper if you think it's necessary.
+Focus on one set of plots from the Mosteller paper and one set of plots from our exponential stuff, so you're demonstrating linear and nonlinear analysis.
+Include basic demographics, and compare to Mosteller's paper - we also had a sample of highly educated individuals.
+Right now you have a very short summary, but you need to actually explain the full process behind it, with all of the details you'd expect from a full experimental writeup.}
+
 ## Tool Validation
@@
 ## Data Analysis
+
+\svp{Explain why our method works so well here - you can actually test whether the PCA residuals are closer than the OLS residuals formally, where Mosteller et al just said "seems like".
+Your method also extends beyond OLS using GAMMs.
+Fully describe the models you fit - assumptions, equations, etc.}
 
@@
 
-# Future Work
+\svp{It's also worth remarking on the effectiveness of the spaghetti plots and GAMM plots for visual analysis of the results. Gotta keep pounding the "visual analytics is statistics" drum a bit :). }
 
-In this work, we implemented and validated ‘You Draw It’ as a new method to measure the patterns we see in data.
-We demonstrated the use of generalized additive models to statistically model participant feedback data and generalized the analysis method to nonlinear trends.
-While technical details of the development process are presented here, we intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers.
-Further investigation is necessary to implement this method with real data in order to facilitate scientific communication.
+# Conclusion
+
+\svp{The data presented in this paper are part of a much broader experiment on the perception of log scales; while this paper focuses primarily on the computational implementation of the You Draw It method and its adaptation to testing graphics, we are currently finishing the analysis and description of the broader experiment, which uses this You Draw It method to assess user prediction of exponential trends.}
+
+\svp{By introducing an interactive method for assessing eye-fit data summaries, along with an analysis method which allows us to test competing hypotheses to determine whether user responses are similar to particular statistical models, we have provided another tool for experimentally testing statistical charts.
+In the future, we hope to create an R package to more easily facilitate these types of user experiments, making this technique available to other researchers who may not be willing to tinker with JavaScript directly.}
+
+\svp{As with any method for testing statistical graphics, there are a host of additional studies which would be useful to understand how users react to the You Draw It method and what parameters are most important for the researcher to control.
+One avenue of future exploration is to investigate the effect of axis limits on user anchoring.
+We know that the perceptual experience is heavily affected by anchoring to e.g.
+axis breaks, and we would expect that a similar anchoring effect might be present based on the limits of the plot and the amount of space in x and y provided for the user to draw.}
+
+XXX add references to the anchoring bit XXX
+
+# References

### 2022-07-05 2022-sdss-you-draw-it-manuscript.tex
A couple more comments

@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, frame hidden, enhanced, boxrule=0pt, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable]}{\end{tcolorbox}}\fi
 
@@
 
+\svp{Describe the experiment formally.
+Don't be afraid to quote Mosteller's paper if you think it's necessary.
+Focus on one set of plots from the Mosteller paper and one set of plots from our exponential stuff, so you're demonstrating linear and nonlinear analysis.
+Include basic demographics, and compare to Mosteller's paper - we also had a sample of highly educated individuals.
+Right now you have a very short summary, but you need to actually explain the full process behind it, with all of the details you'd expect from a full experimental writeup.}
+
 \hypertarget{tool-validation}{%
@@
 \subsection{Data Analysis}\label{data-analysis}}
+
+\svp{Explain why our method works so well here - you can actually test whether the PCA residuals are closer than the OLS residuals formally, where Mosteller et al just said "seems like".
+Your method also extends beyond OLS using GAMMs.
+Fully describe the models you fit - assumptions, equations, etc.}
 
@@
 
-\hypertarget{future-work}{%
-\section{Future Work}\label{future-work}}
+\svp{It's also worth remarking on the effectiveness of the spaghetti plots and GAMM plots for visual analysis of the results. Gotta keep pounding the "visual analytics is statistics" drum a bit :). }
 
-In this work, we implemented and validated `You Draw It' as a new method
-to measure the patterns we see in data. We demonstrated the use of
-generalized additive models to statistically model participant feedback
-data and generalized the analysis method to nonlinear trends. While
-technical details of the development process are presented here, we
-intend to create an R package designed for easy implementation of `You
-Draw It' task plots in order to make this tool accessible to other
-researchers. Further investigation is necessary to implement this method
-with real data in order to facilitate scientific communication.
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
+
+\svp{The data presented in this paper are part of a much broader experiment on the perception of log scales; while this paper focuses primarily on the computational implementation of the You Draw It method and its adaptation to testing graphics, we are currently finishing the analysis and description of the broader experiment, which uses this You Draw It method to assess user prediction of exponential trends.}
+
+\svp{By introducing an interactive method for assessing eye-fit data summaries, along with an analysis method which allows us to test competing hypotheses to determine whether user responses are similar to particular statistical models, we have provided another tool for experimentally testing statistical charts.
+In the future, we hope to create an R package to more easily facilitate these types of user experiments, making this technique available to other researchers who may not be willing to tinker with JavaScript directly.}
+
+\svp{As with any method for testing statistical graphics, there are a host of additional studies which would be useful to understand how users react to the You Draw It method and what parameters are most important for the researcher to control.
+One avenue of future exploration is to investigate the effect of axis limits on user anchoring.
+We know that the perceptual experience is heavily affected by anchoring to e.g.
+axis breaks, and we would expect that a similar anchoring effect might be present based on the limits of the plot and the amount of space in x and y provided for the user to draw.}
+
+XXX add references to the anchoring bit XXX
+
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
 

### 2022-07-06 2022-sdss-you-draw-it-manuscript.qmd
added more to the validation study and addressed a few comments above.

--- 2022-07-05 2022-sdss-you-draw-it-manuscript.qmd
@@
 
+```{r eyefitting-data, message=FALSE, warning=FALSE, echo = F}
+eyefitting_model_data <- read_csv("data/eyefitting-model-data.csv") %>%
+  dplyr::rename(`Parameter Choice` = parm_id)
+
+eyefitting_simulated_data <- read_csv("data/eyefitting-simulated-data.csv") %>%
+  dplyr::rename(`Parameter Choice` = parm_id)
+
+eyefitting_parameters <- read_csv("data/eyefitting-parameters.csv") %>%
+  dplyr::rename(`Parameter Choice` = parm_id)
+
+eyefitting_lmer_preds <- read_csv("data/eyefitting-lmer-preds.csv") %>%
+  dplyr::rename(`Parameter Choice` = parm_id)
+
+eyefitting_gamm_preds <- read_csv("data/eyefitting-gamm-preds.csv") %>%
+  dplyr::rename(`Parameter Choice` = parm_id)
+
+eyefitting_example_sim <- read.csv("data/eyefitting-simdata-example.csv")
+```
+
 # Introduction
 
-We all use statistical graphics, but how do we know that the graphics we use are communicating \svp{effectively}?
-Through experimentation, graphical testing methods allow researchers to conduct studies \svp{designed to understand how we perceive graphics and perform graphical tasks} such as differentiation, prediction, estimation, and extrapolation.
-\svp{Each of these levels of interaction with a graph require a different method of engagement with and use of the information presented in a chart.}
-<!-- All of these types of tests require different levels of use and manipulation of the information being presented in the chart. -->
-<!-- Efforts in the field of statistical graphics have developed graphical testing tools and methods, such as the lineup protocol @buja2009statistical.  -->
-<!-- The advancement of graphing software provides the tools necessary to develop new methods of testing statistical graphics. -->
-\svp{In this paper, we describe the adaptation of an old tool for graphical testing and evaluation, eye-fitting, for use in modern web-applications suitable for testing statistical graphics. 
-We present an empirical evaluation of this testing method for linear regression, and briefly discuss an extension of this method to non-linear applications.}
+We all use statistical graphics, but how do we know that the graphics we use are communicating effectively?
+Through experimentation, graphical testing methods allow researchers to conduct studies designed to understand how we perceive graphics and perform graphical tasks such as differentiation, prediction, estimation, and extrapolation.
+Each of these levels of interaction with a graph require a different method of engagement with and use of the information presented in a chart.
+In this paper, we describe the adaptation of an old tool for graphical testing and evaluation, eye-fitting, for use in modern web-applications suitable for testing statistical graphics. 
+We present an empirical evaluation of this testing method for linear regression, and briefly discuss an extension of this method to non-linear applications.
 
-\svp{One of the most common charts created is a scatter plot of points over time; these charts show up regularly in news articles and in scientific publications alike.
-These charts rely on our} ability to identify and detect trends in data. 
-Our visual system is naturally built to look for structure and identify patterns, \svp{including patterns and trends over time; many times we do not even notice this process happening subconsciously}.
-\svp{As shown in @fig:gas-prices, a viewer engaging with a plot of weekly average gas prices over time may perform several cognitive operations.
+One of the most common charts created is a scatter plot of points over time; these charts show up regularly in news articles and in scientific publications alike.
+These charts rely on our ability to identify and detect trends in data. 
+Our visual system is naturally built to look for structure and identify patterns, including patterns and trends over time; many times we do not even notice this process happening subconsciously.
+As shown in @fig-gas-prices, a viewer engaging with a plot of weekly average gas prices over time may perform several cognitive operations.
 First, they scan the plot and assesses points to determine if there are any outliers or otherwise remarkable points.
 Then, they may fit a rough mental smooth/trend line to the points to summarize the useful information and remove variability.
-Finally, an interested and engaged viewer may pull in additional contextual information from long-term memory, seeking to explain variation in the mental 'trend' with supplemental information, such as COVID lock downs which reduced gasoline demand and the beginning of the war in Ukraine, which wreaked havoc on the supply and demand for global energy sources.}
+Finally, an interested and engaged viewer may pull in additional contextual information from long-term memory, seeking to explain variation in the mental 'trend' with supplemental information, such as COVID lock downs which reduced gasoline demand and the beginning of the war in Ukraine, which wreaked havoc on the supply and demand for global energy sources.
 
@@
 #| out-width: \columnwidth
-#| label: gas-prices
+#| label: fig-gas-prices
 #| fig-res: 300
@@
 
-Initial studies in the 20th century explored the use of fitting lines by eye through a set of points [@finney1951subjective,mosteller1981eye].
+Initial studies in the 20th century explored the use of fitting lines by eye through a set of points [@finney1951subjective, @mosteller1981eye].
 Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit was suitable, then drawing the line through the set of points.
-Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.\svp{XXX and found what?
-Describe the results XXX}
+\ear{Results from these early studies provided groundwork for visually judging slopes and selecting accurate starting values for common iterative calculations.} Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.
+<!-- \svp{XXX and found what? Describe the results XXX} -->
+\ear{This set of studies asked participants to judge trends, estimate slopes, and conduct extrapolation by using a track-pad to adjust the tilt of a line on the screen. 
+Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increased with noise and with number of points.
+This supported the results found in @mosteller1981eye and suggested that participants might use Deming regression [@deming1943statistical], which minimizes the Euclidean distance of points from the line, when fitting a line to a noisy scatter-plot.}
 
-\svp{While psychologists and statisticians have been using eye-fitting techniques to assess our innate perceptual statistical modeling abilities, news organizations have used similar techniques in order to draw readers in and demonstrate the difference between readers' assumptions and reality.} In 2015, the New York Times introduced an interactive feature, called 'You Draw It' [@aisch2015you], where readers input their own assumptions about various \svp{metrics of news interest} and compare these assumptions to reality.
-The New York Times team used Data Driven Documents (D3), \svp{a JavaScript library} that allows readers to \svp{interact with a chart directly by} drawing a line on their computer screen with a mouse.
-\svp{Despite the somewhat different purpose behind this feature, the D3 driven method used by the NYT is wonderfully intuitive, and does not require the assumption of linearity, making it much more adaptable to testing how viewers perceive and predict when presented with non-linear data.}
+While psychologists and statisticians have been using eye-fitting techniques to assess our innate perceptual statistical modeling abilities, news organizations have used similar techniques in order to draw readers in and demonstrate the difference between readers' assumptions and reality.
+In 2015, the New York Times introduced an interactive feature, called 'You Draw It' [@aisch2015you], where readers input their own assumptions about various metrics of news interest and compare these assumptions to reality.
+The New York Times team used Data Driven Documents (D3), a JavaScript library that allows readers to interact with a chart directly by drawing a line on their computer screen with a mouse.
+Despite the somewhat different purpose behind this feature, the D3 driven method used by the NYT is wonderfully intuitive, and does not require the assumption of linearity, making it much more adaptable to testing how viewers perceive and predict when presented with non-linear data.
 
-\svp{We set out to implement} ‘You Draw It’, adapted from the New York Times feature, as a way to experimentally assess the patterns we see in data. 
-Here, we provide technical details of the software development, utilizing interactive graphics in R[@r-software]. 
-We then share results from our study which validates 'You Draw It' as a method for graphical testing \svp{on linear trends} and apply an appropriate data analysis method to the participant data.
-\svp{We also briefly demonstrate the use of the 'You Draw It' method and analysis on nonlinear data.}
+We set out to implement ‘You Draw It’, adapted from the New York Times feature, as a way to experimentally assess the patterns we see in data. 
+Here, we provide technical details of the software development, utilizing interactive graphics in R [@r-software]. 
+We then share results from our study which validates 'You Draw It' as a method for graphical testing on linear trends and apply an appropriate data analysis method to the participant data.
+We also briefly demonstrate the use of the 'You Draw It' method and analysis on nonlinear data.
 
@@
 
-\svp{The New York Times uses D3, Data Driven Documents [@bostock2011d3], to create many of the interactive data graphics that users rely on, from the famous "election needle" in 2016 to their covid dashboards.
-Naturally, this same framework was used} to create the 'You Draw It' \svp{feature}.
-<!-- Major news and research organizations such as the New York Times, FiveThirtyEight, Washington Post, and the pew Research Center use D3 to create and customize graphics.  -->
+The New York Times uses D3, Data Driven Documents [@bostock2011d3], to create many of the interactive data graphics that users rely on, from the famous "election needle" in 2016 to their COVID dashboards.
+Naturally, this same framework was used to create the 'You Draw It' feature.
 
 A challenge of working with D3 is the environment necessary to display the graphics and images.
-\svp{XXX What environment, why is this a challenge, and how did you overcome it?}
+<!-- \svp{XXX What environment, why is this a challenge, and how did you overcome it?} -->
+\ear{Typically, a server setup is needed in order to render D3 visuals and show output from JavaScript on the webpage.
+This can be a barrier to getting started with D3 for someone unfamiliar with server setup and minimal JavaScript coding background.
+Resources, such as @wattenberger-fullstack, help novice users overcome this challenge by walking them step by step through setting up a local server and creating their first plot in D3.
+In the next section, we will see how the `r2d3` package [@r2d3pkg] in R can be used to display D3 visuals in familiar R HTML formats such as Rmarkdown and Shiny.}
 
@@
 \svp{Describe the process of using `r2d3`}
-<!-- The `r2d3` package [@r2d3pkg] provides an efficient integration of D3 visuals into R HTML formats.
-We integrate the D3 visual source code into an R Shiny [@shinypkg] application in order to allow for user interaction and data collection.
--->
-\svp{We leveraged the `r2d3` package [@r2d3pkg] to take data randomly generated in R and create a D3 plot that could be modified with the You Draw It script.
+
+We leveraged the `r2d3` package to take data randomly generated in R and create a D3 plot that could be modified with the You Draw It script.
 This step allowed us to more easily connect the data generation process with the resulting D3 code without having to generate new code manually each time we re-generated the data.
-While not strictly necessary for the validation test of the You Draw It method compared to past linear regression methods presented in this paper, this abstraction makes it much easier to test arbitrary or model-generated data which is unique to each participant.}
+While not strictly necessary for the validation test of the 'You Draw It' method compared to past linear regression methods presented in this paper, this abstraction makes it much easier to test arbitrary or model-generated data which is unique to each participant.
 
@@
 Parameters for aesthetic design choices are defined in a list of options in the `r2d3` function call; `r2d3` passes these to the generated JavaScript code. 
-For instance, we can specify the buffer space allowed for the $x$ and $y$ axes to avoid users \svp{anchoring} their lines to the axes limits.
+For instance, we can specify the buffer space allowed for the $x$ and $y$ axes to avoid users anchoring their lines to the axes limits.
 
 ## Modifying You Draw It
-\svp{
-In order to make use of the original You Draw It code for perceptual experiments, we first needed to accommodate an additional layer.
+
+In order to make use of the original 'You Draw It' code for perceptual experiments, we first needed to accommodate an additional layer.
 The original NYT features asked participants to draw on a blank coordinate grid, as shown in @fig-nyt-screenshot.
-}
 
@@
 
-\svp{When testing perception of graphics, however, we want to pre-populate the chart with additional information - points, and in some cases, portions of a trend line.
-This requires that we modify the original JavaScript code to accommodate these additional elements, which we generate using `r2d3`, as described above.}
+When testing perception of graphics, however, we want to pre-populate the chart with additional information - points, and in some cases, portions of a trend line.
+This requires that we modify the original JavaScript \ear{source} code to accommodate these additional elements, which we generate using `r2d3`, as described above.
 
 We defined JavaScript functions to draw the initial plot and set up drawable points for the user drawn line. 
-Drag events in `D3.js` are utilized to observe and react to user input, \svp{as in the original You Draw It script from the NYT}.
+Drag events in `D3.js` are utilized to observe and react to user input, as in the original 'You Draw It' script from the NYT.
 ^[The full source code we developed for the YouDrawIt task is available at \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.]
@@
 
-\svp{One constraint we inherited from the NYT code is that users can only draw one-to-one functions; because the code works with drag events, each point in $x$ can correspond to only one point in $y$, at least as the code is currently written.
-While this is not particularly problematic for our applications to date, it might limit applications of this type of user interaction when assessing user drawn confidence bands, ribbons, and other situations where two or more vertical points are required.}
+One constraint we inherited from the NYT code is that users can only draw one-to-one functions; because the code works with drag events, each point in $x$ can correspond to only one point in $y$, at least as the code is currently written.
+While this is not particularly problematic for our applications to date, it might limit applications of this type of user interaction when assessing user drawn confidence bands, ribbons, and other situations where two or more vertical points are required.
 <!-- \svp{XXX additional details of how we accomplished this XXX} -->
@@
 
-\svp{During testing of our modifications to the You Draw It script, we discovered that frequently the JavaScript code would ``skip" from point to point out of sequence.
+During testing of our modifications to the You Draw It' script, we discovered that frequently the JavaScript code would ``skip" from point to point out of sequence.
 This resulted in a jagged line (visually) with missing values in the underlying stored array of data; as a result, the for-loop controlling the user-drawn line never exited and the user's data was not recorded.
-While it would have been possible to fix this using some sort of interpolation algorithm, we did not want to compromise user results by introducing interpolation artifacts, so we opted instead for a visual cue that fixed the problem from a psychological angle.}
+While it would have been possible to fix this using some sort of interpolation algorithm, we did not want to compromise user results by introducing interpolation artifacts, so we opted instead for a visual cue that fixed the problem from a psychological angle.
 
@@
 
-\svp{One challenge introduced by adding this visual cue is that D3 uses Scalable Vector Graphics (SVG) to render elements.
-Adding an additional layer meant that we had to ensure that not only the grids, lines, and points of the default scatter plot and trend line rendered in the correct order, but that the yellow box rendered behind all of these points as well so that no information was masked.}
+One challenge introduced by adding this visual cue is that D3 uses Scalable Vector Graphics (SVG) to render elements. 
+Adding an additional layer meant that we had to ensure that not only the grids, lines, and points of the default scatter plot and trend line rendered in the correct order, but that the yellow box rendered behind all of these points as well so that no information was masked.
 
@@
 Shiny Messages are used to communicate the user interaction between the JavaScript code and the R environment. 
-\svp{The initial plot is rendered using Shiny's `RenderD3` and `d3Output` functions; subsequent user interactions are controlled via the JavaScript code.
+The initial plot is rendered using Shiny's `RenderD3` and `d3Output` functions; subsequent user interactions are controlled via the JavaScript code.
 Once the user is finished modifying the plot, they can submit their response, so long as they have filled in all of the points.
-This is enforced using Shiny's message-passing interface and a hook in the JavaScript code that notifies Shiny when the array of user-drawn points is completely filled in.}
-<!-- The plot is then rendered and updated on user interaction into the R shiny application with the `RenderD3` and `d3Output` functions. -->
+This is enforced using Shiny's message-passing interface and a hook in the JavaScript code that notifies Shiny when the array of user-drawn points is completely filled in.
 Once the user is done drawing the line, we save the results of the drawn line to a SQLite database, using Shiny to pass data from JavaScript to R. 
@@
 
-\svp{One additional challenge when integrating the D3 graphics into the Shiny application is that by default, most Shiny applications use a reactive framework that adjusts to the user's browser size.
+One additional challenge when integrating the D3 graphics into the Shiny application is that by default, most Shiny applications use a reactive framework that adjusts to the user's browser size.
 When working with D3, however, this can be problematic: most D3 parameters are specified at the pixel level.
@@
 While in an ideal world, users could participate using cell phones, tablets, and traditional laptop/desktop computers, functionally our checks limited users' ability to participate using smaller screen sizes found in cell phones and some tablets.
-}
-
-\svp{Additionally, after some feedback by laptop users during pilot testing, we included an additional requirement that participants have a computer mouse available; the results from touchpad users were qualitatively different (more jagged) in ways suggesting that the recorded data did not adequately describe users' perceptions.}
+Additionally, after some feedback by laptop users during pilot testing, we included an additional requirement that participants have a computer mouse available; the results from touchpad users were qualitatively different (more jagged) in ways suggesting that the recorded data did not adequately describe users' perceptions.
 
@@
 
-## Tool Validation
+<!-- \svp{Describe the experiment formally.
+Don't be afraid to quote Mosteller's paper if you think it's necessary.
+Focus on one set of plots from the Mosteller paper and one set of plots from our exponential stuff, so you're demonstrating linear and nonlinear analysis.
+Include basic demographics, and compare to Mosteller's paper - we also had a sample of highly educated individuals.
+Right now you have a very short summary, but you need to actually explain the full process behind it, with all of the details you'd expect from a full experimental writeup.} -->
+
+## Validation Study
 
 We conducted a study in order to validate 'You Draw It' as a method for graphical testing, comparing results to the less technological method utilized in @mosteller1981eye.
-Results from our study were consistent with those found in the previous study; when shown points following a linear trend, participants tended to fit the slope of the first principal component over the slope of the ordinary least-squares regression line (Figure 3).
-This trend was most prominent when shown data simulated with larger variances.
-This study reinforced the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.
+The original study asked 153 graduate students and postdoctoral researchers in an Introductory Biostatistics course to fit lines by eye to a set of four points (S, F, V, N), each set with differing slope and variance properties, using an 8.5 x 11 inch transparency with a straight line etched across the middle.
+Researchers conducted the study with a latin square experimental design to test for a practice effect on the task.
+Results indicated participants tended to fit the slope of the first principal component (minimizes orthogonal residuals) as opposed to the slope from the ordinary least squares regression (minimizes vertical residuals) [@fig-pca-plot] and found no effect of order.
 
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/pca-plot}}
-\caption{Comparison between an OLS regression equation which minimizes the vertical distance of points from the line and a regression equation with a slope calculated by the first principal component which minimizes the smallest distance of points from the line.}
-\label{pca-plot}
-\end{center}
-\end{figure}
+```{r ols-vs-pca-example, message = F, warning = F}
+#| fig-cap: "Comparison between an OLS regression equation which minimizes the vertical distance of points from the line and a regression equation with a slope calculated by the first principal component which minimizes the smallest distance of points from the line."
+#| out-width: \columnwidth
+#| label: fig-pca-plot
+#| fig-res: 300
+#| fig-width: 6
+#| fig-height: 4
+#| fig-env: figure
 
-## Data Analysis
+library(ggplot2)
+library(magrittr)
+library(plyr)
 
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-trial-plot}}
-\caption{Example of validation feedback data where three trend lines show the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points}
-\label{eyefitting-trial-plot}
-\end{center}
-\end{figure}
+set.seed(2)
+corrCoef = 0.5 # sample from a multivariate normal, 10 datapoints
+dat = MASS::mvrnorm(10,c(0,0),Sigma = matrix(c(1,corrCoef,2,corrCoef),2,2))
+dat[,1] = dat[,1] - mean(dat[,1]) # it makes life easier for the princomp
+dat[,2] = dat[,2] - mean(dat[,2])
 
-Feedback data from conducted studies were collected and stored in a database for analysis (Figure 4).
-Within the collected feedback data, we knew the simulated data points, the predicted values from the statistical model, and the predicted values from the user drawn line.
-In our initial studies, a unique data set was simulated independently for each participant.
-Therefore, we evaluate the accuracy of the user drawn line by observing the deviation, vertical residuals, between the user drawn line and the predicted values from the statistical model.
-Due to participant variation, we use mixed models to evaluate the vertical residuals in order to statistically compare visually fitted trends to actual metrics, simulated data models, or statistical regression results.
-In our validation study, we first analyzed the residual trend using a linear mixed model (LMM), thus constraining the fit to a linear trend (Figure 5).
-In addition, we fit a Generalized Additive Mixed Model (GAMM) to allow for flexibility in the residual trend and to generalized the method to nonlinear models (Figure 6).
+dat = data.frame(x1 = dat[,1],x2 = dat[,2])
 
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-lmer-plot}}
-\caption{Validation study estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a linear mixed model.}
-\label{eyefitting-lmer-plot}
-\end{center}
-\end{figure}
+# Calculate the first principle component
+# see http://stats.stackexchange.com/questions/13152/how-to-perform-orthogonal-regression-total-least-squares-via-pca
+v = dat%>%prcomp%$%rotation
+x1x2cor = bCor = v[2,1]/v[1,1]
 
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-gamm-plot}}
-\caption{Validation study estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a generalized additive mixed model.}
-\label{eyefitting-gamm-plot}
-\end{center}
-\end{figure}
+x1tox2 = coef(lm(x1~x2,dat))
+x2tox1 = coef(lm(x2~x1,dat))
+slopeData = data.frame(slope = c(x1x2cor,x2tox1[2]),
+                       type=c("Principal Component", "Ordinary Least Squares"))
+
+# We want this to draw the neat orthogonal lines.
+pointOnLine = function(inp){
+  # y = a*x + c (c=0)
+  # yOrth = -(1/a)*x + d
+  # yOrth = b*x + d
+  x0 = inp[1] 
+  y0 = inp[2] 
+  a = x1x2cor
+  b = -(1/a)
+  c = 0
+  d = y0 - b*x0
+  x = (d-c)/(a-b)
+  y = -(1/a)*x+d
+  return(c(x,y))
+}
+
+points = apply(dat,1,FUN=pointOnLine)
+
+segmeData = rbind(data.frame(x=dat[,1],y=dat[,2],xend=points[1,],yend=points[2,],type = "Principal Component"),
+                  data.frame(x=dat[,1],y=dat[,2],yend=dat[,1]*x2tox1[2],xend=dat[,1],type="Ordinary Least Squares"))
+
+ols_pca_plot <- dat %>%
+ggplot(aes(x1,x2))+
+  geom_point()+
+  geom_abline(data=slopeData,aes(slope = slope,intercept=0,color=type, linetype=type), size = 1.2)+
+  geom_segment(data=segmeData,aes(x=x,y=y,xend=xend,yend=yend,color=type, linetype=type))+
+  facet_grid(.~type)+
+  coord_equal()+
+  scale_x_continuous("x") +
+  scale_y_continuous("y") +
+  theme_bw(base_size = 14) +
+  theme(aspect.ratio = 1,
+        legend.position = "none",
+        axis.text    = element_text(size = 12),
+        axis.title   = element_text(size = 12),
+        legend.title = element_blank(),
+        # legend.text  = element_text(size = 10),
+        # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
+        # strip.background = element_rect(size = 0.8),
+        legend.key.size = unit(1, "line")
+        ) +
+  scale_color_manual(values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
+  scale_linetype_manual(values = c("solid", "dashed"), labels = c("OLS", "PCA"))
+ols_pca_plot
+```
+
+In our study, we replicated @mosteller1981eye using the 'You Draw It' method and simulated data with parameter coefficients selected to reflect those from the four data sets in the original study.
+Data were simulated based on a linear model with additive errors:
+\begin{align}
+y_i & = \beta_0 + \beta_1 x_i + e_i \\
+\text{with } e_i & \sim N(0, \sigma^2). \nonumber
+\end{align} 
+For each participant, a unique data set was randomly and independently generated from the underlying parameters at the beginning of the study and mapped to a scatter-plot graphic.
+@fig-eyefitting-simplot illustrates an example of simulated data for all four parameter choices intended to reflect the trends in the original study.
+When participants started the study, they were first asked to complete two practice plots - accompanied by instructions and a .gif demonstrating the task - in order to train them in the skills associated with executing the task.
+The four 'You Draw It' task plots associated with the selected parameters followed the practice plots in a random order for each individual.
+
+```{r eyefitting-simplot, warning = F, message = F}
+#| fig-cap: "Example of simulated data points displayed in a scatter-plot illustrating the trends associated with the four selected parameter choices."
+#| out-width: \columnwidth
+#| label: fig-eyefitting-simplot
+#| fig-res: 300
+#| fig-width: 9
+#| fig-height: 9
+#| fig-env: figure
+
+eyefitting_example_simplot <- eyefitting_example_sim %>%
+  filter(data == "point_data") %>%
+  filter(dataset %in% c("F", "N", "S") | (x < 16 & x > 4)) %>%
+  mutate(dataset = factor(dataset, levels = c("S", "F", "V", "N"))) %>%
+  dplyr::rename(`Parameter Choice` = dataset) %>%
+  ggplot(aes(x = x, y = y)) +
+  geom_point(size = 1) +
+  facet_wrap(~`Parameter Choice`, ncol = 2) +
+  theme_bw(base_size = 14) +
+  theme(aspect.ratio = 1,
+  legend.position = "none",
+  plot.title   = element_text(size = 12, hjust = 0),
+  axis.text    = element_text(size = 12),
+  axis.title   = element_text(size = 12),
+  legend.title = element_text(size = 12),
+  legend.text  = element_text(size = 12),
+  # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
+  # strip.background = element_rect(size = 0.5),
+  legend.key.size = unit(1, "line")
+) +
+  scale_y_continuous(breaks = seq(-10, 20, 5))
+eyefitting_example_simplot
+```
+
+Participants were recruited via Prolific in March 2022; while this study does utilize a convenience sample, as this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context [CITE SPATIAL PAPER].
+The study was conducted and distributed via a Shiny application and participants completed the tasks on their own computer, with a computer mouse, and in an environment of their choosing.
+The data from this study were collected to validate this method of graphical testing, with the hopes of providing a new tool to assess graphical perception interactively.
+
+## Data Analysis and Results
+
+<!-- \svp{Explain why our method works so well here - you can actually test whether the PCA residuals are closer than the OLS residuals formally, where Mosteller et al just said "seems like".
+Your method also extends beyond OLS using GAMMs.
+Fully describe the models you fit - assumptions, equations, etc.} -->
+
+As a part of the validation study, we collected data for each participant drawn trend line, allowing us to conduct formal analyses which compared the intuitive visually fit lines to statistical regression results - the original study lacked these formal comparisons, providing only summaries of averages, variances, and actual (least squares) values for the slope and intercept of each data set.
+The feedback data, stored in a SQLite database, contained the simulated data points} - $(x_{ijk}, y_{ijk})$ -, the predicted values from the statistical regression models - $\hat y_{ijk,OLS}$, and $\hat y_{ijk,PCA}$ -, and the predicted values from the user drawn line - $y_{ijk,drawn}$ for parameter choice $i = 1,2,3,4$, $j = 1,...N_\text{participant}$, and $x_{ijk}$ value $k = 1, ...,4 x_{max} + 1$.
+@fig-eyefitting-trial-plot displays an example of all three fitted trend lines for parameter choice F.
+
+```{r eyefitting-trial-plot, warning = F, message = F}
+#| fig-cap: "Example of validation feedback data where three trend lines show the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points."
+#| out-width: \columnwidth
+#| label: fig-eyefitting-trial-plot
+#| fig-res: 300
+#| fig-width: 4
+#| fig-height: 4
+#| fig-env: figure
+#|
+
+eyefitting_example <- eyefitting_model_data %>%
+  # filter(participant_id == "65c10d498eae365e108efcd3dcb75287", parm_id == "N") %>%
+  filter(participant_id == "60b16b9bd5a122c1457d31055df51a45", `Parameter Choice` == "F") %>%
+  ggplot(aes(x = x)) +
+  geom_line(aes(y = yols, color = "OLS", linetype = "OLS")) +
+  geom_line(aes(y = ypca, color = "PCA", linetype = "PCA")) +
+  geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
+  geom_point(data = eyefitting_simulated_data %>%
+               filter(dataset == "point_data", participant_id == "60b16b9bd5a122c1457d31055df51a45", `Parameter Choice` == "F"),
+             aes(x = x, y = y)) +
+  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both)) +
+  theme_bw(base_size = 14) +
+  theme(aspect.ratio = 1,
+        legend.position = "bottom",
+        axis.text    = element_text(size = 12),
+        axis.title   = element_text(size = 12),
+        legend.title = element_text(size = 12),
+        legend.text  = element_text(size = 12),
+        # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
+        # strip.background = element_rect(size = 0.8),
+        legend.key.size = unit(1, "line")
+        ) +
+  scale_x_continuous(limits = c(0,20)) +
+  scale_color_manual("", values = c("black", "steelblue", "orange")) +
+  scale_linetype_manual("", values = c("dashed", "solid", "solid")) +
+  scale_y_continuous("y")
+
+eyefitting_example
+```
+
+A unique data set was simulated independently for each participant, therefore, comparisons of vertical residuals between the user drawn line and the OLS fitted values ($e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}$) and PCA fitted values ($e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}$) were used to assess the accuracy of participant drawn lines.
+We use mixed model methods to analyze residual trends and capture the variability due to participant differences.
+
+We first fit a linear mixed model (LMM) to the OLS and PCA residuals separately using the `lmer` function in the `lme4` package [@lme4-pkg], constraining the fit to a linear trend.
+Parameter choice, $x$, and the interaction between $x$ and parameter choice were treated as fixed effects with a random participant effect accounting for variation due to participant.
+The LMM equation for each fit (OLS and PCA) residuals is given by:
+\begin{equation}
+y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
+\end{equation}
+\noindent where
+
++ $y_{ijk,drawn}$ is the drawn y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$-value
++ $\hat y_{ijk,fit}$ is the fitted y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$-value corresponding to either the OLS or PCA fit
++ $e_{ijk,fit}$ is the residual between the drawn and fitted y-values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$-value corresponding to either the OLS or PCA fit
++ $\gamma_0$ is the overall intercept
++ $\alpha_i$ is the effect of the $i^{th}$ parameter choice (F, S, V, N) on the intercept
++ $\gamma_1$ is the overall slope for $x$
++ $\gamma_{2i}$ is the effect of the parameter choice on the slope
++ $x_{ijk}$ is the $x$-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
++ $p_{j} \sim N(0, \sigma^2_\text{participant})$ is the random error due to the $j^{th}$ participant's characteristics
++ $\epsilon_{ijk} \sim N(0, \sigma^2)$ is the residual error.
+
+```{r eyefitting-lmer-residualplots, warning = F, message = F}
+#| fig-cap: "Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by the linear mixed model.
+A random sample of 75 participants was selected to display the individual participant residuals behind the overall trend."
+#| out-width: \columnwidth
+#| label: fig-eyefitting-lmer-residualplots
+#| fig-res: 300
+#| fig-width: 9
+#| fig-height: 9
+#| fig-env: figure
+#|
+
+set.seed(68505)
+participant_sample <- sample(unique(eyefitting_model_data$prolific_id), 75)
+
+# Plot Predictions
+eyefitting_lmer_plot <- eyefitting_lmer_preds %>%
+  filter((`Parameter Choice` %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
+  mutate(`Parameter Choice` = factor(`Parameter Choice`, levels = c("S", "F", "V", "N"))) %>%
+  ggplot(aes(x = x)) +
+  geom_line(data = eyefitting_model_data %>% 
+              filter(prolific_id %in% participant_sample), 
+            aes(x = x, y = residual_ols_loess, group = plot_id, color = "OLS"), alpha = 0.1) +
+  geom_line(data = eyefitting_model_data %>% 
+              filter(prolific_id %in% participant_sample), 
+            aes(x = x, y = residual_pca_loess, group = plot_id, color = "PCA"), alpha = 0.1) +
+  geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.4) +
+  geom_line(aes(y = emmean.ols, color = "OLS")) +
+  geom_ribbon(aes(ymin = asymp.LCL.pca, ymax = asymp.UCL.pca, fill = "PCA"), color = NA, alpha = 0.4) +
+  geom_line(aes(y = emmean.pca, color = "PCA")) +
+  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
+  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both), scales = "free") +
+  theme_bw(base_size = 14) +
+  theme(aspect.ratio = 1,
+        legend.position = "right",
+        plot.title   = element_text(size = 12, hjust = 0),
+        axis.text    = element_text(size = 12),
+        axis.title   = element_text(size = 12),
+        legend.title = element_text(size = 12),
+        legend.text  = element_text(size = 12),
+        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
+        # strip.background = element_rect(size = 0.5),
+        legend.key.size = unit(1, "line")
+        ) +
+  scale_y_continuous("Residual", limits = c(-5,5), breaks = seq(-5,5,2.5)) +
+  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
+  scale_fill_manual("LMER fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA"))  +
+  guides(color = guide_legend(override.aes = list(alpha = 1)),
+         fill = guide_legend(override.aes = list(alpha = 1)))
+
+eyefitting_lmer_plot
+```
+
+Eliminating the linear trend constraint, we can extend the method and analysis beyond ordinary least squares by using generalized additive mixed models (GAMM) to estimate smoothing splines and allow for a more flexible residual trend.
+The `bam` function in the `mgcv` package was used to fit a GAMM separately to the OLS and PCA.
+Parameter choice was treated as a fixed effect with no estimated intercept and a separate smoothing spline for $x$ was estimated for each parameter choice. 
+A random participant effect accounting for variation due to participant and a random spline for each participant accounted for variation in spline for each participant.
+The GAMM equation for each fit (OLS and PCA) residuals is given by:
+\begin{equation}
+y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
+\end{equation}
+\noindent where
+
++ $y_{ijk,drawn}$ is the drawn y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$-value
++ $\hat y_{ijk,fit}$ is the fitted y-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$-value corresponding to either the OLS or PCA fit
++ $e_{ijk,fit}$ is the residual between the drawn and fitted y-values for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment of $x$-value corresponding to either the OLS or PCA fit
++ $\alpha_i$ is the intercept for the parameter choice $i$
++ $s_{i}$ is the smoothing spline for the $i^{th}$ parameter choice
++ $x_{ijk}$ is the $x$-value for the $i^{th}$ parameter choice, $j^{th}$ participant, and $k^{th}$ increment
++ $p_{j} \sim N(0, \sigma^2_\text{participant})$ is the error due to participant variation
++ $s_{j}$ is the random smoothing spline for each participant.
+
+```{r eyefitting-gamm-residualplots, warning = F, message = F}
+#| fig-cap: "Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by the generalized additive mixed model.
+A random sample of 75 participants was selected to display the individual participant residuals behind the overall trend."
+#| out-width: \columnwidth
+#| label: fig-eyefitting-gamm-residualplots
+#| fig-res: 300
+#| fig-width: 9
+#| fig-height: 9
+#| fig-env: figure
+#|
+
+eyefitting_gamm_plot <- eyefitting_gamm_preds %>%
+  filter((`Parameter Choice` %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
+  mutate(`Parameter Choice` = factor(`Parameter Choice`, levels = c("S", "F", "V", "N"))) %>%
+  ggplot(aes(x = x)) +
+  geom_line(data = eyefitting_model_data %>% 
+              filter(prolific_id %in% participant_sample), 
+            aes(x = x, y = residual_ols_loess, group = plot_id, color = "OLS"), alpha = 0.1) +
+  geom_line(data = eyefitting_model_data %>% 
+              filter(prolific_id %in% participant_sample), 
+            aes(x = x, y = residual_pca_loess, group = plot_id, color = "PCA"), alpha = 0.1) +
+  geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.4) +
+  geom_line(aes(y = ols.pred, color = "OLS")) +
+  geom_ribbon(aes(ymin = pca.lower, ymax = pca.upper, fill = "PCA"), color = NA, alpha = 0.4) +
+  geom_line(aes(y = pca.pred, color = "PCA")) +
+  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
+  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both), scales = "free") +
+  theme_bw(base_size = 14) +
+  theme(aspect.ratio = 1,
+        legend.position = "right",
+        plot.title   = element_text(size = 12, hjust = 0),
+        axis.text    = element_text(size = 12),
+        axis.title   = element_text(size = 12),
+        legend.title = element_text(size = 12),
+        legend.text  = element_text(size = 12),
+        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
+        # strip.background = element_rect(size = 0.5),
+        legend.key.size = unit(1, "line")
+        ) +
+  scale_y_continuous("Residual", limits = c(-5,5), breaks = seq(-5,5,2.5)) +
+  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
+  scale_fill_manual("GAMM fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA"))   +
+  guides(color = guide_legend(override.aes = list(alpha = 1)),
+         fill = guide_legend(override.aes = list(alpha = 1)))
+
+eyefitting_gamm_plot
+```
+
+@fig-eyefitting-lmer-residualplots and @fig-eyefitting-gamm-residualplots show a visual analysis of the estimated trends of residuals (vertical deviation of participant drawn points from both the OLS and PCA fitted points) as modeled by a LMM and GAMM respectively.
+A random sample of 75 participants was selected to display individual participant residuals behind the overall residual trend.
+Examining the plots, the estimated trends of PCA residuals (orange) appear to align more parallel and closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue). 
+In particular, this trend is more prominent in parameter choices with large variances (F and N).
+Results from our study were consistent with those found in the original study; when shown points following a linear trend, participants tended to fit the slope of the first principal component Our study established 'You Draw It' as a method for graphical testing and reinforced the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.
+
+## Extension to Nonlinear Data
 
 After validating the 'You Draw It' method, we conducted a study utilizing the new method to evaluate participants ability to make forecasts for exponentially increasing data on log and linear scales.
-Along with analyzing the feedback data with the GAMM method for flexibility due to nonlinear data, we used spaghetti plots to visualize participants forecasts compared to the nonlinear least squares statistical model and make comparisons between two chart design features (Figure 6).
+Along with analyzing the feedback data with the GAMM method for flexibility due to nonlinear data, we used spaghetti plots to conduct a visual analysis of participant forecasts compared to the nonlinear least squares statistical model and make comparisons between two chart design features [@fig-exponential-yloess-spaghetti-plot-2-1].
+This study demonstrated the advantage of using the 'You Draw It' method paired with GAMM and visual analysis methods for measuring perception and visual patterns of nonlinear data.
 
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/exponential-yloess-spaghetti-plot-2-1}}
-\caption{Spaghetti plot of results from a study which asked participants to forcast trends of exponentially increasing data.
+![Spaghetti plot of results from a study which asked participants to forcast trends of exponentially increasing data.
 Participants drawn lines on the linear scale are shown in blue and the log scale are shown in orange.
-Variability in the statistically fitted regression lines occured due to a unique data set being simulated for each individual; the gray band shows the range fitted values from the statistically fitted regression lines.}
-\label{exponential-yloess-spaghetti-plot-2-1}
-\end{center}
-\end{figure}
+Variability in the statistically fitted regression lines occured due to a unique data set being simulated for each individual; the gray band shows the range fitted values from the statistically fitted regression lines.](images/exponential-yloess-spaghetti-plot-2-1.png){#fig-exponential-yloess-spaghetti-plot-2-1}
 
-# Future Work
+<!-- \svp{It's also worth remarking on the effectiveness of the spaghetti plots and GAMM plots for visual analysis of the results.
+Gotta keep pounding the "visual analytics is statistics" drum a bit :).
+} -->
 
-In this work, we implemented and validated ‘You Draw It’ as a new method to measure the patterns we see in data.
-We demonstrated the use of generalized additive models to statistically model participant feedback data and generalized the analysis method to nonlinear trends.
-While technical details of the development process are presented here, we intend to create an R package designed for easy implementation of 'You Draw It' task plots in order to make this tool accessible to other researchers.
-Further investigation is necessary to implement this method with real data in order to facilitate scientific communication.
+# Conclusion
+
+The data presented in this paper are part of a much broader experiment on the perception of log scales; while this paper focuses primarily on the computational implementation of the 'You Draw It' method and its adaptation to testing graphics, we are currently finishing the analysis and description of the broader experiment, which uses this 'You Draw It' method to assess user prediction of exponential trends.
+
+By introducing an interactive method for assessing eye-fit data summaries, along with an analysis method which allows us to test competing hypotheses to determine whether user responses are similar to particular statistical models, we have provided another tool for experimentally testing statistical charts.
+In the future, we hope to create an R package to more easily facilitate these types of user experiments, making this technique available to other researchers who may not be willing to tinker with JavaScript directly.
+
+As with any method for testing statistical graphics, there are a host of additional studies which would be useful to understand how users react to the 'You Draw It' method and what parameters are most important for the researcher to control.
+One avenue of future exploration is to investigate the effect of axis limits on user anchoring.
+We know that the perceptual experience is heavily affected by anchoring to e.g.
+axis breaks, and we would expect that a similar anchoring effect might be present based on the limits of the plot and the amount of space in $x$ and $y$ provided for the user to draw.
+
+\svp{XXX add references to the anchoring bit XXX}
+
+# References

### 2022-07-06 2022-sdss-you-draw-it-manuscript.tex
added more to the validation study and addressed a few comments above.

--- 2022-07-05 2022-sdss-you-draw-it-manuscript.tex
@@
 \makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable]}{\end{tcolorbox}}\fi
 
@@
 We all use statistical graphics, but how do we know that the graphics we
-use are communicating \svp{effectively}? Through experimentation,
-graphical testing methods allow researchers to conduct studies
-\svp{designed to understand how we perceive graphics and perform graphical tasks}
-such as differentiation, prediction, estimation, and extrapolation.
-\svp{Each of these levels of interaction with a graph require a different method of engagement with and use of the information presented in a chart.}
-\svp{In this paper, we describe the adaptation of an old tool for graphical testing and evaluation, eye-fitting, for use in modern web-applications suitable for testing statistical graphics. 
-We present an empirical evaluation of this testing method for linear regression, and briefly discuss an extension of this method to non-linear applications.}
+use are communicating effectively? Through experimentation, graphical
+testing methods allow researchers to conduct studies designed to
+understand how we perceive graphics and perform graphical tasks such as
+differentiation, prediction, estimation, and extrapolation. Each of
+these levels of interaction with a graph require a different method of
+engagement with and use of the information presented in a chart. In this
+paper, we describe the adaptation of an old tool for graphical testing
+and evaluation, eye-fitting, for use in modern web-applications suitable
+for testing statistical graphics. We present an empirical evaluation of
+this testing method for linear regression, and briefly discuss an
+extension of this method to non-linear applications.
 
-\svp{One of the most common charts created is a scatter plot of points over time; these charts show up regularly in news articles and in scientific publications alike.
-These charts rely on our} ability to identify and detect trends in data.
-Our visual system is naturally built to look for structure and identify
-patterns,
-\svp{including patterns and trends over time; many times we do not even notice this process happening subconsciously}.
-\svp{As shown in @fig:gas-prices, a viewer engaging with a plot of weekly average gas prices over time may perform several cognitive operations.
-First, they scan the plot and assesses points to determine if there are any outliers or otherwise remarkable points.
-Then, they may fit a rough mental smooth/trend line to the points to summarize the useful information and remove variability.
-Finally, an interested and engaged viewer may pull in additional contextual information from long-term memory, seeking to explain variation in the mental 'trend' with supplemental information, such as COVID lock downs which reduced gasoline demand and the beginning of the war in Ukraine, which wreaked havoc on the supply and demand for global energy sources.}
+One of the most common charts created is a scatter plot of points over
+time; these charts show up regularly in news articles and in scientific
+publications alike. These charts rely on our ability to identify and
+detect trends in data. Our visual system is naturally built to look for
+structure and identify patterns, including patterns and trends over
+time; many times we do not even notice this process happening
+subconsciously. As shown in Figure~\ref{fig-gas-prices}, a viewer
+engaging with a plot of weekly average gas prices over time may perform
+several cognitive operations. First, they scan the plot and assesses
+points to determine if there are any outliers or otherwise remarkable
+points. Then, they may fit a rough mental smooth/trend line to the
+points to summarize the useful information and remove variability.
+Finally, an interested and engaged viewer may pull in additional
+contextual information from long-term memory, seeking to explain
+variation in the mental `trend' with supplemental information, such as
+COVID lock downs which reduced gasoline demand and the beginning of the
+war in Ukraine, which wreaked havoc on the supply and demand for global
+energy sources.
 
@@
 
-{\centering \includegraphics{./images/gas-prices-1.pdf}
+{\centering \includegraphics{./images/fig-gas-prices-1.pdf}
 
@@
 
-\caption{Weekly average gas prices in the United States, 2019-June 2022.
-Additional mental operations a viewer might perform while looking at the
-plot are annotated in grey. US Energy Information Administration (2022)}
+\caption{\label{fig-gas-prices}Weekly average gas prices in the United
+States, 2019-June 2022. Additional mental operations a viewer might
+perform while looking at the plot are annotated in grey. US Energy
+Information Administration (2022)}
 
@@
 Initial studies in the 20th century explored the use of fitting lines by
-eye through a set of points (Finney 1951, mosteller1981eye). Common
-methods of fitting trends by eye involved maneuvering a string, black
-thread, or ruler until the fit was suitable, then drawing the line
-through the set of points. Recently, Ciccione and Dehaene (2021)
-conducted a comprehensive set of studies investigating human ability to
-detect trends in graphical representations from a psychophysical
-approach.\svp{XXX and found what? Describe the results XXX}
+eye through a set of points Mosteller et al. (1981). Common methods of
+fitting trends by eye involved maneuvering a string, black thread, or
+ruler until the fit was suitable, then drawing the line through the set
+of points.
+\ear{Results from these early studies provided groundwork for visually judging slopes and selecting accurate starting values for common iterative calculations.}
+Recently, Ciccione and Dehaene (2021) conducted a comprehensive set of
+studies investigating human ability to detect trends in graphical
+representations from a psychophysical approach.
+\ear{This set of studies asked participants to judge trends, estimate slopes, and conduct extrapolation by using a track-pad to adjust the tilt of a line on the screen. 
+Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increased with noise and with number of points.
+This supported the results found in @mosteller1981eye and suggested that participants might use Deming regression [@deming1943statistical], which minimizes the Euclidean distance of points from the line, when fitting a line to a noisy scatter-plot.}
 
-\svp{While psychologists and statisticians have been using eye-fitting techniques to assess our innate perceptual statistical modeling abilities, news organizations have used similar techniques in order to draw readers in and demonstrate the difference between readers' assumptions and reality.}
-In 2015, the New York Times introduced an interactive feature, called
-`You Draw It' (Aisch, Cox, and Quealy 2015), where readers input their
-own assumptions about various \svp{metrics of news interest} and compare
-these assumptions to reality. The New York Times team used Data Driven
-Documents (D3), \svp{a JavaScript library} that allows readers to
-\svp{interact with a chart directly by} drawing a line on their computer
-screen with a mouse.
-\svp{Despite the somewhat different purpose behind this feature, the D3 driven method used by the NYT is wonderfully intuitive, and does not require the assumption of linearity, making it much more adaptable to testing how viewers perceive and predict when presented with non-linear data.}
+While psychologists and statisticians have been using eye-fitting
+techniques to assess our innate perceptual statistical modeling
+abilities, news organizations have used similar techniques in order to
+draw readers in and demonstrate the difference between readers'
+assumptions and reality. In 2015, the New York Times introduced an
+interactive feature, called `You Draw It' (Aisch, Cox, and Quealy 2015),
+where readers input their own assumptions about various metrics of news
+interest and compare these assumptions to reality. The New York Times
+team used Data Driven Documents (D3), a JavaScript library that allows
+readers to interact with a chart directly by drawing a line on their
+computer screen with a mouse. Despite the somewhat different purpose
+behind this feature, the D3 driven method used by the NYT is wonderfully
+intuitive, and does not require the assumption of linearity, making it
+much more adaptable to testing how viewers perceive and predict when
+presented with non-linear data.
 
-\svp{We set out to implement} `You Draw It', adapted from the New York
-Times feature, as a way to experimentally assess the patterns we see in
-data. Here, we provide technical details of the software development,
-utilizing interactive graphics in R(R Core Team 2022). We then share
+We set out to implement `You Draw It', adapted from the New York Times
+feature, as a way to experimentally assess the patterns we see in data.
+Here, we provide technical details of the software development,
+utilizing interactive graphics in R (R Core Team 2022). We then share
 results from our study which validates `You Draw It' as a method for
-graphical testing \svp{on linear trends} and apply an appropriate data
-analysis method to the participant data.
-\svp{We also briefly demonstrate the use of the 'You Draw It' method and analysis on nonlinear data.}
+graphical testing on linear trends and apply an appropriate data
+analysis method to the participant data. We also briefly demonstrate the
+use of the `You Draw It' method and analysis on nonlinear data.
 
@@
 
-\svp{The New York Times uses D3, Data Driven Documents [@bostock2011d3], to create many of the interactive data graphics that users rely on, from the famous "election needle" in 2016 to their covid dashboards.
-Naturally, this same framework was used}
-to create the `You Draw It' \svp{feature}.
+The New York Times uses D3, Data Driven Documents (Bostock, Ogievetsky,
+and Heer 2011), to create many of the interactive data graphics that
+users rely on, from the famous ``election needle'' in 2016 to their
+COVID dashboards. Naturally, this same framework was used to create the
+`You Draw It' feature.
 
@@
 the graphics and images.
-\svp{XXX What environment, why is this a challenge, and how did you overcome it?}
+\ear{Typically, a server setup is needed in order to render D3 visuals and show output from JavaScript on the webpage.
+This can be a barrier to getting started with D3 for someone unfamiliar with server setup and minimal JavaScript coding background.
+Resources, such as @wattenberger-fullstack, help novice users overcome this challenge by walking them step by step through setting up a local server and creating their first plot in D3.
+In the next section, we will see how the `r2d3` package [@r2d3pkg] in R can be used to display D3 visuals in familiar R HTML formats such as Rmarkdown and Shiny.}
 
@@
 
-\svp{We leveraged the `r2d3` package [@r2d3pkg] to take data randomly generated in R and create a D3 plot that could be modified with the You Draw It script.
-This step allowed us to more easily connect the data generation process with the resulting D3 code without having to generate new code manually each time we re-generated the data.
-While not strictly necessary for the validation test of the You Draw It method compared to past linear regression methods presented in this paper, this abstraction makes it much easier to test arbitrary or model-generated data which is unique to each participant.}
+We leveraged the \texttt{r2d3} package to take data randomly generated
+in R and create a D3 plot that could be modified with the You Draw It
+script. This step allowed us to more easily connect the data generation
+process with the resulting D3 code without having to generate new code
+manually each time we re-generated the data. While not strictly
+necessary for the validation test of the `You Draw It' method compared
+to past linear regression methods presented in this paper, this
+abstraction makes it much easier to test arbitrary or model-generated
+data which is unique to each participant.
 
@@
 buffer space allowed for the \(x\) and \(y\) axes to avoid users
-\svp{anchoring} their lines to the axes limits.
+anchoring their lines to the axes limits.
 
@@
 
-\svp{
-In order to make use of the original You Draw It code for perceptual experiments, we first needed to accommodate an additional layer.
-The original NYT features asked participants to draw on a blank coordinate grid, as shown in @fig-nyt-screenshot.
-}
+In order to make use of the original `You Draw It' code for perceptual
+experiments, we first needed to accommodate an additional layer. The
+original NYT features asked participants to draw on a blank coordinate
+grid, as shown in Figure~\ref{fig-nyt-screenshot}.
 
@@
 
-\svp{When testing perception of graphics, however, we want to pre-populate the chart with additional information - points, and in some cases, portions of a trend line.
-This requires that we modify the original JavaScript code to accommodate these additional elements, which we generate using `r2d3`, as described above.}
+When testing perception of graphics, however, we want to pre-populate
+the chart with additional information - points, and in some cases,
+portions of a trend line. This requires that we modify the original
+JavaScript \ear{source} code to accommodate these additional elements,
+which we generate using \texttt{r2d3}, as described above.
 
@@
 drawable points for the user drawn line. Drag events in \texttt{D3.js}
-are utilized to observe and react to user input,
-\svp{as in the original You Draw It script from the NYT}. \footnote{The
-  full source code we developed for the YouDrawIt task is available at
+are utilized to observe and react to user input, as in the original `You
+Draw It' script from the NYT. \footnote{The full source code we
+  developed for the YouDrawIt task is available at
   \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.}
 
-\svp{One constraint we inherited from the NYT code is that users can only draw one-to-one functions; because the code works with drag events, each point in $x$ can correspond to only one point in $y$, at least as the code is currently written.
-While this is not particularly problematic for our applications to date, it might limit applications of this type of user interaction when assessing user drawn confidence bands, ribbons, and other situations where two or more vertical points are required.}
+One constraint we inherited from the NYT code is that users can only
+draw one-to-one functions; because the code works with drag events, each
+point in \(x\) can correspond to only one point in \(y\), at least as
+the code is currently written. While this is not particularly
+problematic for our applications to date, it might limit applications of
+this type of user interaction when assessing user drawn confidence
+bands, ribbons, and other situations where two or more vertical points
+are required.
 
@@
 
-\svp{During testing of our modifications to the You Draw It script, we discovered that frequently the JavaScript code would ``skip" from point to point out of sequence.
-This resulted in a jagged line (visually) with missing values in the underlying stored array of data; as a result, the for-loop controlling the user-drawn line never exited and the user's data was not recorded.
-While it would have been possible to fix this using some sort of interpolation algorithm, we did not want to compromise user results by introducing interpolation artifacts, so we opted instead for a visual cue that fixed the problem from a psychological angle.}
+During testing of our modifications to the You Draw It' script, we
+discovered that frequently the JavaScript code would ``skip'' from point
+to point out of sequence. This resulted in a jagged line (visually) with
+missing values in the underlying stored array of data; as a result, the
+for-loop controlling the user-drawn line never exited and the user's
+data was not recorded. While it would have been possible to fix this
+using some sort of interpolation algorithm, we did not want to
+compromise user results by introducing interpolation artifacts, so we
+opted instead for a visual cue that fixed the problem from a
+psychological angle.
 
@@
 
-\svp{One challenge introduced by adding this visual cue is that D3 uses Scalable Vector Graphics (SVG) to render elements.
-Adding an additional layer meant that we had to ensure that not only the grids, lines, and points of the default scatter plot and trend line rendered in the correct order, but that the yellow box rendered behind all of these points as well so that no information was masked.}
+One challenge introduced by adding this visual cue is that D3 uses
+Scalable Vector Graphics (SVG) to render elements. Adding an additional
+layer meant that we had to ensure that not only the grids, lines, and
+points of the default scatter plot and trend line rendered in the
+correct order, but that the yellow box rendered behind all of these
+points as well so that no information was masked.
 
@@
 Shiny Messages are used to communicate the user interaction between the
-JavaScript code and the R environment.
-\svp{The initial plot is rendered using Shiny's `RenderD3` and `d3Output` functions; subsequent user interactions are controlled via the JavaScript code.
-Once the user is finished modifying the plot, they can submit their response, so long as they have filled in all of the points.
-This is enforced using Shiny's message-passing interface and a hook in the JavaScript code that notifies Shiny when the array of user-drawn points is completely filled in.}
-Once the user is done drawing the line, we save the results of the drawn
-line to a SQLite database, using Shiny to pass data from JavaScript to
-R. The connections between each portion of the app are shown in
-Figure~\ref{fig-you-draw-it-code-sketch}.
+JavaScript code and the R environment. The initial plot is rendered
+using Shiny's \texttt{RenderD3} and \texttt{d3Output} functions;
+subsequent user interactions are controlled via the JavaScript code.
+Once the user is finished modifying the plot, they can submit their
+response, so long as they have filled in all of the points. This is
+enforced using Shiny's message-passing interface and a hook in the
+JavaScript code that notifies Shiny when the array of user-drawn points
+is completely filled in. Once the user is done drawing the line, we save
+the results of the drawn line to a SQLite database, using Shiny to pass
+data from JavaScript to R. The connections between each portion of the
+app are shown in Figure~\ref{fig-you-draw-it-code-sketch}.
 
@@
 
-\svp{One additional challenge when integrating the D3 graphics into the Shiny application is that by default, most Shiny applications use a reactive framework that adjusts to the user's browser size.
-When working with D3, however, this can be problematic: most D3 parameters are specified at the pixel level.
-The discontinuity between the assumptions of Shiny and D3 meant that we had to fix the size of the Shiny element, and then perform a check to ensure that the user's screen was sufficiently big to render the plot.
-While in an ideal world, users could participate using cell phones, tablets, and traditional laptop/desktop computers, functionally our checks limited users' ability to participate using smaller screen sizes found in cell phones and some tablets.
-}
-
-\svp{Additionally, after some feedback by laptop users during pilot testing, we included an additional requirement that participants have a computer mouse available; the results from touchpad users were qualitatively different (more jagged) in ways suggesting that the recorded data did not adequately describe users' perceptions.}
+One additional challenge when integrating the D3 graphics into the Shiny
+application is that by default, most Shiny applications use a reactive
+framework that adjusts to the user's browser size. When working with D3,
+however, this can be problematic: most D3 parameters are specified at
+the pixel level. The discontinuity between the assumptions of Shiny and
+D3 meant that we had to fix the size of the Shiny element, and then
+perform a check to ensure that the user's screen was sufficiently big to
+render the plot. While in an ideal world, users could participate using
+cell phones, tablets, and traditional laptop/desktop computers,
+functionally our checks limited users' ability to participate using
+smaller screen sizes found in cell phones and some tablets.
+Additionally, after some feedback by laptop users during pilot testing,
+we included an additional requirement that participants have a computer
+mouse available; the results from touchpad users were qualitatively
+different (more jagged) in ways suggesting that the recorded data did
+not adequately describe users' perceptions.
 
@@
 
-\hypertarget{tool-validation}{%
-\subsection{Tool Validation}\label{tool-validation}}
+\hypertarget{validation-study}{%
+\subsection{Validation Study}\label{validation-study}}
 
@@
 graphical testing, comparing results to the less technological method
-utilized in Mosteller et al. (1981). Results from our study were
-consistent with those found in the previous study; when shown points
-following a linear trend, participants tended to fit the slope of the
-first principal component over the slope of the ordinary least-squares
-regression line (Figure 3). This trend was most prominent when shown
-data simulated with larger variances. This study reinforced the
-differences between intuitive visual model fitting and statistical model
-fitting, providing information about human perception as it relates to
-the use of statistical graphics.
+utilized in Mosteller et al. (1981). The original study asked 153
+graduate students and postdoctoral researchers in an Introductory
+Biostatistics course to fit lines by eye to a set of four points (S, F,
+V, N), each set with differing slope and variance properties, using an
+8.5 x 11 inch transparency with a straight line etched across the
+middle. Researchers conducted the study with a latin square experimental
+design to test for a practice effect on the task. Results indicated
+participants tended to fit the slope of the first principal component
+(minimizes orthogonal residuals) as opposed to the slope from the
+ordinary least squares regression (minimizes vertical residuals)
+Figure~\ref{fig-pca-plot} and found no effect of order.
 
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/pca-plot}}
-\caption{Comparison between an OLS regression equation which minimizes the vertical distance of points from the line and a regression equation with a slope calculated by the first principal component which minimizes the smallest distance of points from the line.}
-\label{pca-plot}
-\end{center}
+\begin{figure}
+
+{\centering \includegraphics{./images/fig-pca-plot-1.pdf}
+
+}
+
+\caption{\label{fig-pca-plot}Comparison between an OLS regression
+equation which minimizes the vertical distance of points from the line
+and a regression equation with a slope calculated by the first principal
+component which minimizes the smallest distance of points from the
+line.}
+
 \end{figure}
 
-\hypertarget{data-analysis}{%
-\subsection{Data Analysis}\label{data-analysis}}
+In our study, we replicated Mosteller et al. (1981) using the `You Draw
+It' method and simulated data with parameter coefficients selected to
+reflect those from the four data sets in the original study. Data were
+simulated based on a linear model with additive errors: \begin{align}
+y_i & = \beta_0 + \beta_1 x_i + e_i \\
+\text{with } e_i & \sim N(0, \sigma^2). \nonumber
+\end{align} For each participant, a unique data set was randomly and
+independently generated from the underlying parameters at the beginning
+of the study and mapped to a scatter-plot graphic.
+Figure~\ref{fig-eyefitting-simplot} illustrates an example of simulated
+data for all four parameter choices intended to reflect the trends in
+the original study. When participants started the study, they were first
+asked to complete two practice plots - accompanied by instructions and a
+.gif demonstrating the task - in order to train them in the skills
+associated with executing the task. The four `You Draw It' task plots
+associated with the selected parameters followed the practice plots in a
+random order for each individual.
 
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-trial-plot}}
-\caption{Example of validation feedback data where three trend lines show the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points}
-\label{eyefitting-trial-plot}
-\end{center}
+\begin{figure}
+
+{\centering \includegraphics{./images/fig-eyefitting-simplot-1.pdf}
+
+}
+
+\caption{\label{fig-eyefitting-simplot}Example of simulated data points
+displayed in a scatter-plot illustrating the trends associated with the
+four selected parameter choices.}
+
 \end{figure}
 
-Feedback data from conducted studies were collected and stored in a
-database for analysis (Figure 4). Within the collected feedback data, we
-knew the simulated data points, the predicted values from the
-statistical model, and the predicted values from the user drawn line. In
-our initial studies, a unique data set was simulated independently for
-each participant. Therefore, we evaluate the accuracy of the user drawn
-line by observing the deviation, vertical residuals, between the user
-drawn line and the predicted values from the statistical model. Due to
-participant variation, we use mixed models to evaluate the vertical
-residuals in order to statistically compare visually fitted trends to
-actual metrics, simulated data models, or statistical regression
-results. In our validation study, we first analyzed the residual trend
-using a linear mixed model (LMM), thus constraining the fit to a linear
-trend (Figure 5). In addition, we fit a Generalized Additive Mixed Model
-(GAMM) to allow for flexibility in the residual trend and to generalized
-the method to nonlinear models (Figure 6).
+Participants were recruited via Prolific in March 2022; while this study
+does utilize a convenience sample, as this is primarily a perceptual
+task, previous results have found few differences between expert and
+non-expert participants in this context {[}CITE SPATIAL PAPER{]}. The
+study was conducted and distributed via a Shiny application and
+participants completed the tasks on their own computer, with a computer
+mouse, and in an environment of their choosing. The data from this study
+were collected to validate this method of graphical testing, with the
+hopes of providing a new tool to assess graphical perception
+interactively.
 
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-lmer-plot}}
-\caption{Validation study estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a linear mixed model.}
-\label{eyefitting-lmer-plot}
-\end{center}
+\hypertarget{data-analysis-and-results}{%
+\subsection{Data Analysis and Results}\label{data-analysis-and-results}}
+
+As a part of the validation study, we collected data for each
+participant drawn trend line, allowing us to conduct formal analyses
+which compared the intuitive visually fit lines to statistical
+regression results - the original study lacked these formal comparisons,
+providing only summaries of averages, variances, and actual (least
+squares) values for the slope and intercept of each data set. The
+feedback data, stored in a SQLite database, contained the simulated data
+points\} - \((x_{ijk}, y_{ijk})\) -, the predicted values from the
+statistical regression models - \(\hat y_{ijk,OLS}\), and
+\(\hat y_{ijk,PCA}\) -, and the predicted values from the user drawn
+line - \(y_{ijk,drawn}\) for parameter choice \(i = 1,2,3,4\),
+\(j = 1,...N_\text{participant}\), and \(x_{ijk}\) value
+\(k = 1, ...,4 x_{max} + 1\). Figure~\ref{fig-eyefitting-trial-plot}
+displays an example of all three fitted trend lines for parameter choice
+F.
+
+\begin{figure}
+
+{\centering \includegraphics{./images/fig-eyefitting-trial-plot-1.pdf}
+
+}
+
+\caption{\label{fig-eyefitting-trial-plot}Example of validation feedback
+data where three trend lines show the the OLS fitted, PCA fitted, and
+participant drawn values overlaid on the simulated data points.}
+
 \end{figure}
 
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/eyefitting-gamm-plot}}
-\caption{Validation study estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by a generalized additive mixed model.}
-\label{eyefitting-gamm-plot}
-\end{center}
+A unique data set was simulated independently for each participant,
+therefore, comparisons of vertical residuals between the user drawn line
+and the OLS fitted values
+(\(e_{ijk,OLS} = y_{ijk,drawn} - \hat y_{ijk,OLS}\)) and PCA fitted
+values (\(e_{ijk,PCA} = y_{ijk,drawn} - \hat y_{ijk,PCA}\)) were used to
+assess the accuracy of participant drawn lines. We use mixed model
+methods to analyze residual trends and capture the variability due to
+participant differences.
+
+We first fit a linear mixed model (LMM) to the OLS and PCA residuals
+separately using the \texttt{lmer} function in the \texttt{lme4} package
+(Bates et al. 2015), constraining the fit to a linear trend. Parameter
+choice, \(x\), and the interaction between \(x\) and parameter choice
+were treated as fixed effects with a random participant effect
+accounting for variation due to participant. The LMM equation for each
+fit (OLS and PCA) residuals is given by: \begin{equation}
+y_{ijk,drawn} - \hat y_{ijk,fit} = e_{ijk,fit} = \left[\gamma_0 + \alpha_i\right] + \left[\gamma_{1} x_{ijk} + \gamma_{2i} x_{ijk}\right] + p_{j} + \epsilon_{ijk}
+\end{equation} \noindent where
+
+\begin{itemize}
+\tightlist
+\item
+  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
+  choice, \(j^{th}\) participant, and \(k^{th}\) increment of
+  \(x\)-value
+\item
+  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
+  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
+  \(x\)-value corresponding to either the OLS or PCA fit
+\item
+  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
+  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
+  \(k^{th}\) increment of \(x\)-value corresponding to either the OLS or
+  PCA fit
+\item
+  \(\gamma_0\) is the overall intercept
+\item
+  \(\alpha_i\) is the effect of the \(i^{th}\) parameter choice (F, S,
+  V, N) on the intercept
+\item
+  \(\gamma_1\) is the overall slope for \(x\)
+\item
+  \(\gamma_{2i}\) is the effect of the parameter choice on the slope
+\item
+  \(x_{ijk}\) is the \(x\)-value for the \(i^{th}\) parameter choice,
+  \(j^{th}\) participant, and \(k^{th}\) increment
+\item
+  \(p_{j} \sim N(0, \sigma^2_\text{participant})\) is the random error
+  due to the \(j^{th}\) participant's characteristics
+\item
+  \(\epsilon_{ijk} \sim N(0, \sigma^2)\) is the residual error.
+\end{itemize}
+
+\begin{figure}
+
+{\centering \includegraphics{./images/fig-eyefitting-lmer-residualplots-1.pdf}
+
+}
+
+\caption{\label{fig-eyefitting-lmer-residualplots}Estimated trends of
+residuals (vertical deviation of participant drawn points from both the
+OLS (blue) and PCA (orange) fitted points) as fit by the linear mixed
+model. A random sample of 75 participants was selected to display the
+individual participant residuals behind the overall trend.}
+
 \end{figure}
+
+Eliminating the linear trend constraint, we can extend the method and
+analysis beyond ordinary least squares by using generalized additive
+mixed models (GAMM) to estimate smoothing splines and allow for a more
+flexible residual trend. The \texttt{bam} function in the \texttt{mgcv}
+package was used to fit a GAMM separately to the OLS and PCA. Parameter
+choice was treated as a fixed effect with no estimated intercept and a
+separate smoothing spline for \(x\) was estimated for each parameter
+choice. A random participant effect accounting for variation due to
+participant and a random spline for each participant accounted for
+variation in spline for each participant. The GAMM equation for each fit
+(OLS and PCA) residuals is given by: \begin{equation}
+y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
+\end{equation} \noindent where
+
+\begin{itemize}
+\tightlist
+\item
+  \(y_{ijk,drawn}\) is the drawn y-value for the \(i^{th}\) parameter
+  choice, \(j^{th}\) participant, and \(k^{th}\) increment of
+  \(x\)-value
+\item
+  \(\hat y_{ijk,fit}\) is the fitted y-value for the \(i^{th}\)
+  parameter choice, \(j^{th}\) participant, and \(k^{th}\) increment of
+  \(x\)-value corresponding to either the OLS or PCA fit
+\item
+  \(e_{ijk,fit}\) is the residual between the drawn and fitted y-values
+  for the \(i^{th}\) parameter choice, \(j^{th}\) participant, and
+  \(k^{th}\) increment of \(x\)-value corresponding to either the OLS or
+  PCA fit
+\item
+  \(\alpha_i\) is the intercept for the parameter choice \(i\)
+\item
+  \(s_{i}\) is the smoothing spline for the \(i^{th}\) parameter choice
+\item
+  \(x_{ijk}\) is the \(x\)-value for the \(i^{th}\) parameter choice,
+  \(j^{th}\) participant, and \(k^{th}\) increment
+\item
+  \(p_{j} \sim N(0, \sigma^2_\text{participant})\) is the error due to
+  participant variation
+\item
+  \(s_{j}\) is the random smoothing spline for each participant.
+\end{itemize}
+
+\begin{figure}
+
+{\centering \includegraphics{./images/fig-eyefitting-gamm-residualplots-1.pdf}
+
+}
+
+\caption{\label{fig-eyefitting-gamm-residualplots}Estimated trends of
+residuals (vertical deviation of participant drawn points from both the
+OLS (blue) and PCA (orange) fitted points) as fit by the generalized
+additive mixed model. A random sample of 75 participants was selected to
+display the individual participant residuals behind the overall trend.}
+
+\end{figure}
+
+Figure~\ref{fig-eyefitting-lmer-residualplots} and
+Figure~\ref{fig-eyefitting-gamm-residualplots} show a visual analysis of
+the estimated trends of residuals (vertical deviation of participant
+drawn points from both the OLS and PCA fitted points) as modeled by a
+LMM and GAMM respectively. A random sample of 75 participants was
+selected to display individual participant residuals behind the overall
+residual trend. Examining the plots, the estimated trends of PCA
+residuals (orange) appear to align more parallel and closer to the
+\(y=0\) horizontal (dashed) line than the OLS residuals (blue). In
+particular, this trend is more prominent in parameter choices with large
+variances (F and N). Results from our study were consistent with those
+found in the original study; when shown points following a linear trend,
+participants tended to fit the slope of the first principal component
+Our study established `You Draw It' as a method for graphical testing
+and reinforced the differences between intuitive visual model fitting
+and statistical model fitting, providing information about human
+perception as it relates to the use of statistical graphics.
+
+\hypertarget{extension-to-nonlinear-data}{%
+\subsection{Extension to Nonlinear
+Data}\label{extension-to-nonlinear-data}}
 
@@
 Along with analyzing the feedback data with the GAMM method for
-flexibility due to nonlinear data, we used spaghetti plots to visualize
-participants forecasts compared to the nonlinear least squares
-statistical model and make comparisons between two chart design features
-(Figure 6).
+flexibility due to nonlinear data, we used spaghetti plots to conduct a
+visual analysis of participant forecasts compared to the nonlinear least
+squares statistical model and make comparisons between two chart design
+features Figure~\ref{fig-exponential-yloess-spaghetti-plot-2-1}. This
+study demonstrated the advantage of using the `You Draw It' method
+paired with GAMM and visual analysis methods for measuring perception
+and visual patterns of nonlinear data.
 
-\begin{figure}[ht]
-\begin{center}
-\centerline{\includegraphics[width=\columnwidth]{images/exponential-yloess-spaghetti-plot-2-1}}
-\caption{Spaghetti plot of results from a study which asked participants to forcast trends of exponentially increasing data.
-Participants drawn lines on the linear scale are shown in blue and the log scale are shown in orange.
-Variability in the statistically fitted regression lines occured due to a unique data set being simulated for each individual; the gray band shows the range fitted values from the statistically fitted regression lines.}
-\label{exponential-yloess-spaghetti-plot-2-1}
-\end{center}
+\begin{figure}
+
+{\centering \includegraphics{images/exponential-yloess-spaghetti-plot-2-1.png}
+
+}
+
+\caption{\label{fig-exponential-yloess-spaghetti-plot-2-1}Spaghetti plot
+of results from a study which asked participants to forcast trends of
+exponentially increasing data. Participants drawn lines on the linear
+scale are shown in blue and the log scale are shown in orange.
+Variability in the statistically fitted regression lines occured due to
+a unique data set being simulated for each individual; the gray band
+shows the range fitted values from the statistically fitted regression
+lines.}
+
 \end{figure}
 
-\hypertarget{future-work}{%
-\section{Future Work}\label{future-work}}
+\hypertarget{conclusion}{%
+\section{Conclusion}\label{conclusion}}
 
-In this work, we implemented and validated `You Draw It' as a new method
-to measure the patterns we see in data. We demonstrated the use of
-generalized additive models to statistically model participant feedback
-data and generalized the analysis method to nonlinear trends. While
-technical details of the development process are presented here, we
-intend to create an R package designed for easy implementation of `You
-Draw It' task plots in order to make this tool accessible to other
-researchers. Further investigation is necessary to implement this method
-with real data in order to facilitate scientific communication.
+The data presented in this paper are part of a much broader experiment
+on the perception of log scales; while this paper focuses primarily on
+the computational implementation of the `You Draw It' method and its
+adaptation to testing graphics, we are currently finishing the analysis
+and description of the broader experiment, which uses this `You Draw It'
+method to assess user prediction of exponential trends.
+
+By introducing an interactive method for assessing eye-fit data
+summaries, along with an analysis method which allows us to test
+competing hypotheses to determine whether user responses are similar to
+particular statistical models, we have provided another tool for
+experimentally testing statistical charts. In the future, we hope to
+create an R package to more easily facilitate these types of user
+experiments, making this technique available to other researchers who
+may not be willing to tinker with JavaScript directly.
+
+As with any method for testing statistical graphics, there are a host of
+additional studies which would be useful to understand how users react
+to the `You Draw It' method and what parameters are most important for
+the researcher to control. One avenue of future exploration is to
+investigate the effect of axis limits on user anchoring. We know that
+the perceptual experience is heavily affected by anchoring to e.g.~axis
+breaks, and we would expect that a similar anchoring effect might be
+present based on the limits of the plot and the amount of space in \(x\)
+and \(y\) provided for the user to draw.
+
+\svp{XXX add references to the anchoring bit XXX}
+
+\hypertarget{references}{%
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
 
@@
 Times} 28.
+
+\leavevmode\vadjust pre{\hypertarget{ref-lme4-pkg}{}}%
+Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker. 2015.
+{``Fitting Linear Mixed-Effects Models Using {lme4}.''} \emph{Journal of
+Statistical Software} 67 (1): 1--48.
+\url{https://doi.org/10.18637/jss.v067.i01}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-bostock2011d3}{}}%
+Bostock, Michael, Vadim Ogievetsky, and Jeffrey Heer. 2011. {``D\(^3\)
+Data-Driven Documents.''} \emph{IEEE Transactions on Visualization and
+Computer Graphics} 17 (12): 2301--9.
 

### 2022-07-11 2022-sdss-you-draw-it-manuscript.qmd
updated template from SDSS email. Adjusted figures to display only one set.

@@
       - edits.tex
-      - refs.tex
+      # - refs.tex
     keep-tex: true
-    # template: template-blinded.tex
     # template: template.tex
+documentclass: jds
 ---
@@
 library(lubridate)
+library(patchwork)
 ```
@@
 eyefitting_example_sim <- read.csv("data/eyefitting-simdata-example.csv")
+
+youdrawit_model_data      <- read_csv("data/youdrawit-model-data.csv")  %>%
+  mutate(`Points Truncated` = ifelse(points_truncated == 10, "50%", "75%"),
+         `Growth Rate` = ifelse(beta == "beta0.1", "Low", "High")) %>%
+  mutate(`Growth Rate` = factor(`Growth Rate`, levels = c("Low", "High"))) %>%
+  filter(`Points Truncated` == "50%", `Growth Rate` == "High")
+
+youdrawit_simulated_band <- youdrawit_model_data %>%
+  group_by(`Growth Rate`, `Points Truncated`, x) %>%
+  summarize(min_ynls = min(ynls),
+            max_ynls = max(ynls))
+```
+
+```{r eyefitting-simplot, eval = F, warning = F, message = F}
+#| fig-cap: "Example of simulated data points displayed in a scatter-plot illustrating the trends associated with the four selected parameter choices."
+#| out-width: \columnwidth
+#| label: fig-eyefitting-simplot
+#| fig-res: 300
+#| fig-width: 9
+#| fig-height: 9
+#| fig-env: figure
+
+eyefitting_example_simplot <- eyefitting_example_sim %>%
+  filter(data == "point_data") %>%
+  filter(dataset %in% c("F", "N", "S") | (x < 16 & x > 4)) %>%
+  mutate(dataset = factor(dataset, levels = c("S", "F", "V", "N"))) %>%
+  filter(`Parameter Choice` == "F") %>%
+  dplyr::rename(`Parameter Choice` = dataset) %>%
+  ggplot(aes(x = x, y = y)) +
+  geom_point(size = 1) +
+  facet_wrap(~`Parameter Choice`, ncol = 2) +
+  theme_bw(base_size = 14) +
+  theme(aspect.ratio = 1,
+  legend.position = "none",
+  plot.title   = element_text(size = 12, hjust = 0),
+  axis.text    = element_text(size = 12),
+  axis.title   = element_text(size = 12),
+  legend.title = element_text(size = 12),
+  legend.text  = element_text(size = 12),
+  # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
+  # strip.background = element_rect(size = 0.5),
+  legend.key.size = unit(1, "line")
+) +
+  scale_y_continuous(breaks = seq(-10, 20, 5))
+# eyefitting_example_simplot
+```
+
+```{r eyefitting-trial-plot, warning = F, message = F}
+#| fig-cap: "Example of validation feedback data where three trend lines show the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points."
+#| out-width: \columnwidth
+#| label: fig-eyefitting-trial-plot
+#| fig-res: 300
+#| fig-width: 4
+#| fig-height: 4
+#| fig-env: figure
+#|
+
+eyefitting_example <- eyefitting_model_data %>%
+  # filter(participant_id == "65c10d498eae365e108efcd3dcb75287", parm_id == "N") %>%
+  filter(participant_id == "60b16b9bd5a122c1457d31055df51a45", `Parameter Choice` == "F") %>%
+  ggplot(aes(x = x)) +
+  geom_line(aes(y = yols, color = "OLS", linetype = "OLS")) +
+  geom_line(aes(y = ypca, color = "PCA", linetype = "PCA")) +
+  geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
+  geom_point(data = eyefitting_simulated_data %>%
+               filter(dataset == "point_data", participant_id == "60b16b9bd5a122c1457d31055df51a45", `Parameter Choice` == "F"),
+             aes(x = x, y = y)) +
+  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both)) +
+  theme_bw(base_size = 14) +
+  theme(aspect.ratio = 1,
+        legend.position = "bottom",
+        plot.title   = element_text(size = 14, hjust = 0),
+        axis.text    = element_text(size = 12),
+        axis.title   = element_text(size = 12),
+        legend.title = element_text(size = 12),
+        legend.text  = element_text(size = 12),
+        # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
+        # strip.background = element_rect(size = 0.8),
+        legend.key.size = unit(1, "line")
+        ) +
+  scale_x_continuous(limits = c(0,20)) +
+  scale_color_manual("", values = c("black", "steelblue", "orange")) +
+  scale_linetype_manual("", values = c("dashed", "solid", "solid")) +
+  scale_y_continuous("y")
+
+# eyefitting_example
+```
+
+```{r eyefitting-lmer-residualplots, warning = F, message = F}
+#| fig-cap: "Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by the linear mixed model.
+A random sample of 75 participants was selected to display the individual participant residuals behind the overall trend."
+#| out-width: \columnwidth
+#| label: fig-eyefitting-lmer-residualplots
+#| fig-res: 300
+#| fig-width: 9
+#| fig-height: 9
+#| fig-env: figure
+#|
+
+set.seed(68505)
+participant_sample <- sample(unique(eyefitting_model_data$prolific_id), 75)
+
+# Plot Predictions
+eyefitting_lmer_plot <- eyefitting_lmer_preds %>%
+  filter((`Parameter Choice` %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
+  mutate(`Parameter Choice` = factor(`Parameter Choice`, levels = c("S", "F", "V", "N"))) %>%
+  filter(`Parameter Choice` == "F") %>%
+  ggplot(aes(x = x)) +
+  geom_line(data = eyefitting_model_data %>% 
+              filter(prolific_id %in% participant_sample)  %>% filter(`Parameter Choice` == "F"), 
+            aes(x = x, y = residual_ols_loess, group = plot_id, color = "OLS"), alpha = 0.1) +
+  geom_line(data = eyefitting_model_data %>% 
+              filter(prolific_id %in% participant_sample) %>% filter(`Parameter Choice` == "F"), 
+            aes(x = x, y = residual_pca_loess, group = plot_id, color = "PCA"), alpha = 0.1) +
+  geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.4) +
+  geom_line(aes(y = emmean.ols, color = "OLS")) +
+  geom_ribbon(aes(ymin = asymp.LCL.pca, ymax = asymp.UCL.pca, fill = "PCA"), color = NA, alpha = 0.4) +
+  geom_line(aes(y = emmean.pca, color = "PCA")) +
+  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
+  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both), scales = "free") +
+  theme_bw(base_size = 14) +
+  theme(aspect.ratio = 1,
+        legend.position = "bottom",
+        plot.title   = element_text(size = 14, hjust = 0),
+        axis.text    = element_text(size = 12),
+        axis.title   = element_text(size = 12),
+        legend.title = element_text(size = 12),
+        legend.text  = element_text(size = 12),
+        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
+        # strip.background = element_rect(size = 0.5),
+        legend.key.size = unit(1, "line"), legend.box="vertical", legend.margin=margin()
+        ) +
+  scale_y_continuous("Residual", limits = c(-5,5), breaks = seq(-5,5,2.5)) +
+  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
+  scale_fill_manual("LMER fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA"))  +
+  guides(color = guide_legend(override.aes = list(alpha = 1)),
+         fill = guide_legend(override.aes = list(alpha = 1)))
+
+# eyefitting_lmer_plot
+```
+
+```{r eyefitting-gamm-residualplots, warning = F, message = F}
+#| fig-cap: "Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by the generalized additive mixed model.
+A random sample of 75 participants was selected to display the individual participant residuals behind the overall trend."
+#| out-width: \columnwidth
+#| label: fig-eyefitting-gamm-residualplots
+#| fig-res: 300
+#| fig-width: 9
+#| fig-height: 9
+#| fig-env: figure
+#|
+
+eyefitting_gamm_plot <- eyefitting_gamm_preds %>%
+  filter((`Parameter Choice` %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
+  mutate(`Parameter Choice` = factor(`Parameter Choice`, levels = c("S", "F", "V", "N"))) %>%
+  filter(`Parameter Choice` == "F") %>%
+  ggplot(aes(x = x)) +
+  geom_line(data = eyefitting_model_data %>% 
+              filter(prolific_id %in% participant_sample)  %>% filter(`Parameter Choice` == "F"), 
+            aes(x = x, y = residual_ols_loess, group = plot_id, color = "OLS"), alpha = 0.1) +
+  geom_line(data = eyefitting_model_data %>% 
+              filter(prolific_id %in% participant_sample)  %>% filter(`Parameter Choice` == "F"), 
+            aes(x = x, y = residual_pca_loess, group = plot_id, color = "PCA"), alpha = 0.1) +
+  geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.4) +
+  geom_line(aes(y = ols.pred, color = "OLS")) +
+  geom_ribbon(aes(ymin = pca.lower, ymax = pca.upper, fill = "PCA"), color = NA, alpha = 0.4) +
+  geom_line(aes(y = pca.pred, color = "PCA")) +
+  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
+  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both), scales = "free") +
+  theme_bw(base_size = 14) +
+  theme(aspect.ratio = 1,
+        legend.position = "bottom",
+        plot.title   = element_text(size = 14, hjust = 0),
+        axis.text    = element_text(size = 12),
+        axis.title   = element_text(size = 12),
+        legend.title = element_text(size = 12),
+        legend.text  = element_text(size = 12),
+        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
+        # strip.background = element_rect(size = 0.5),
+        legend.key.size = unit(1, "line"), legend.box="vertical", legend.margin=margin()
+        ) +
+  scale_y_continuous("Residual", limits = c(-5,5), breaks = seq(-5,5,2.5)) +
+  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
+  scale_fill_manual("GAMM fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA"))   +
+  guides(color = guide_legend(override.aes = list(alpha = 1)),
+         fill = guide_legend(override.aes = list(alpha = 1)))
+
+# eyefitting_gamm_plot
 ```
@@
 Researchers conducted the study with a latin square experimental design to test for a practice effect on the task.
-Results indicated participants tended to fit the slope of the first principal component (minimizes orthogonal residuals) as opposed to the slope from the ordinary least squares regression (minimizes vertical residuals) [@fig-pca-plot] and found no effect of order.
+Results indicated participants tended to fit the slope of the first principal component (minimizes Euclidean distance) as opposed to the slope from the ordinary least squares regression (minimizes vertical distance) [@fig-pca-plot] and found no effect of order.
 
@@
 #| fig-cap: "Comparison between an OLS regression equation which minimizes the vertical distance of points from the line and a regression equation with a slope calculated by the first principal component which minimizes the smallest distance of points from the line."
-#| out-width: \columnwidth
+#| out-width: "70%"
 #| label: fig-pca-plot
@@
 \end{align} 
-For each participant, a unique data set was randomly and independently generated from the underlying parameters at the beginning of the study and mapped to a scatter-plot graphic.
-@fig-eyefitting-simplot illustrates an example of simulated data for all four parameter choices intended to reflect the trends in the original study.
+For each participant, four unique data sets were randomly and independently generated from the underlying parameters at the beginning of the study and mapped to a scatter-plot graphic.
+<!-- @fig-eyefitting-simplot illustrates an example of simulated data for all four parameter choices intended to reflect the trends in the original study. -->
 When participants started the study, they were first asked to complete two practice plots - accompanied by instructions and a .gif demonstrating the task - in order to train them in the skills associated with executing the task.
@@
 
-```{r eyefitting-simplot, warning = F, message = F}
-#| fig-cap: "Example of simulated data points displayed in a scatter-plot illustrating the trends associated with the four selected parameter choices."
-#| out-width: \columnwidth
-#| label: fig-eyefitting-simplot
-#| fig-res: 300
-#| fig-width: 9
-#| fig-height: 9
-#| fig-env: figure
-
-eyefitting_example_simplot <- eyefitting_example_sim %>%
-  filter(data == "point_data") %>%
-  filter(dataset %in% c("F", "N", "S") | (x < 16 & x > 4)) %>%
-  mutate(dataset = factor(dataset, levels = c("S", "F", "V", "N"))) %>%
-  dplyr::rename(`Parameter Choice` = dataset) %>%
-  ggplot(aes(x = x, y = y)) +
-  geom_point(size = 1) +
-  facet_wrap(~`Parameter Choice`, ncol = 2) +
-  theme_bw(base_size = 14) +
-  theme(aspect.ratio = 1,
-  legend.position = "none",
-  plot.title   = element_text(size = 12, hjust = 0),
-  axis.text    = element_text(size = 12),
-  axis.title   = element_text(size = 12),
-  legend.title = element_text(size = 12),
-  legend.text  = element_text(size = 12),
-  # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
-  # strip.background = element_rect(size = 0.5),
-  legend.key.size = unit(1, "line")
-) +
-  scale_y_continuous(breaks = seq(-10, 20, 5))
-eyefitting_example_simplot
-```
-
-Participants were recruited via Prolific in March 2022; while this study does utilize a convenience sample, as this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context [CITE SPATIAL PAPER].
+Participants were recruited via Prolific in March 2022; while this study does utilize a convenience sample, as this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context \ear{(CITE SPATIAL PAPER)}.
 The study was conducted and distributed via a Shiny application and participants completed the tasks on their own computer, with a computer mouse, and in an environment of their choosing.
@@
 The feedback data, stored in a SQLite database, contained the simulated data points} - $(x_{ijk}, y_{ijk})$ -, the predicted values from the statistical regression models - $\hat y_{ijk,OLS}$, and $\hat y_{ijk,PCA}$ -, and the predicted values from the user drawn line - $y_{ijk,drawn}$ for parameter choice $i = 1,2,3,4$, $j = 1,...N_\text{participant}$, and $x_{ijk}$ value $k = 1, ...,4 x_{max} + 1$.
-@fig-eyefitting-trial-plot displays an example of all three fitted trend lines for parameter choice F.
-
-```{r eyefitting-trial-plot, warning = F, message = F}
-#| fig-cap: "Example of validation feedback data where three trend lines show the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points."
-#| out-width: \columnwidth
-#| label: fig-eyefitting-trial-plot
-#| fig-res: 300
-#| fig-width: 4
-#| fig-height: 4
-#| fig-env: figure
-#|
-
-eyefitting_example <- eyefitting_model_data %>%
-  # filter(participant_id == "65c10d498eae365e108efcd3dcb75287", parm_id == "N") %>%
-  filter(participant_id == "60b16b9bd5a122c1457d31055df51a45", `Parameter Choice` == "F") %>%
-  ggplot(aes(x = x)) +
-  geom_line(aes(y = yols, color = "OLS", linetype = "OLS")) +
-  geom_line(aes(y = ypca, color = "PCA", linetype = "PCA")) +
-  geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
-  geom_point(data = eyefitting_simulated_data %>%
-               filter(dataset == "point_data", participant_id == "60b16b9bd5a122c1457d31055df51a45", `Parameter Choice` == "F"),
-             aes(x = x, y = y)) +
-  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both)) +
-  theme_bw(base_size = 14) +
-  theme(aspect.ratio = 1,
-        legend.position = "bottom",
-        axis.text    = element_text(size = 12),
-        axis.title   = element_text(size = 12),
-        legend.title = element_text(size = 12),
-        legend.text  = element_text(size = 12),
-        # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
-        # strip.background = element_rect(size = 0.8),
-        legend.key.size = unit(1, "line")
-        ) +
-  scale_x_continuous(limits = c(0,20)) +
-  scale_color_manual("", values = c("black", "steelblue", "orange")) +
-  scale_linetype_manual("", values = c("dashed", "solid", "solid")) +
-  scale_y_continuous("y")
-
-eyefitting_example
-```
+@fig-eyefitting-plots displays an example of all three fitted trend lines for parameter choice F.
 
@@
 We use mixed model methods to analyze residual trends and capture the variability due to participant differences.
+
+```{r eyefitting-plots, warning = F, message = F}
+#| fig-cap: "Visual analysis results for parameter choice set F.
+\\linebreak \\textbf{A.} Example of validation feedback data where three trend lines show the the OLS fitted, PCA fitted, and participant drawn values overlaid on the simulated data points.
+\\linebreak \\textbf{B \\& C.} Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by the (B) linear mixed model and (C) generalized additive mixed model.
+A random sample of 75 participants was selected to display the individual participant residuals behind the overall trend."
+#| out-width: \columnwidth
+#| label: fig-eyefitting-plots
+#| fig-res: 300
+#| fig-width: 12
+#| fig-height: 6
+#| fig-env: figure
+
+(eyefitting_example + ggtitle("A. Example")) + 
+  (eyefitting_lmer_plot + ggtitle("B. LMER Results")) + 
+  (eyefitting_gamm_plot + ggtitle("C. GAMM Results")) +
+  plot_layout(ncol =  3)
+```
 
@@
 
-```{r eyefitting-lmer-residualplots, warning = F, message = F}
-#| fig-cap: "Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by the linear mixed model.
-A random sample of 75 participants was selected to display the individual participant residuals behind the overall trend."
-#| out-width: \columnwidth
-#| label: fig-eyefitting-lmer-residualplots
-#| fig-res: 300
-#| fig-width: 9
-#| fig-height: 9
-#| fig-env: figure
-#|
-
-set.seed(68505)
-participant_sample <- sample(unique(eyefitting_model_data$prolific_id), 75)
-
-# Plot Predictions
-eyefitting_lmer_plot <- eyefitting_lmer_preds %>%
-  filter((`Parameter Choice` %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
-  mutate(`Parameter Choice` = factor(`Parameter Choice`, levels = c("S", "F", "V", "N"))) %>%
-  ggplot(aes(x = x)) +
-  geom_line(data = eyefitting_model_data %>% 
-              filter(prolific_id %in% participant_sample), 
-            aes(x = x, y = residual_ols_loess, group = plot_id, color = "OLS"), alpha = 0.1) +
-  geom_line(data = eyefitting_model_data %>% 
-              filter(prolific_id %in% participant_sample), 
-            aes(x = x, y = residual_pca_loess, group = plot_id, color = "PCA"), alpha = 0.1) +
-  geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.4) +
-  geom_line(aes(y = emmean.ols, color = "OLS")) +
-  geom_ribbon(aes(ymin = asymp.LCL.pca, ymax = asymp.UCL.pca, fill = "PCA"), color = NA, alpha = 0.4) +
-  geom_line(aes(y = emmean.pca, color = "PCA")) +
-  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
-  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both), scales = "free") +
-  theme_bw(base_size = 14) +
-  theme(aspect.ratio = 1,
-        legend.position = "right",
-        plot.title   = element_text(size = 12, hjust = 0),
-        axis.text    = element_text(size = 12),
-        axis.title   = element_text(size = 12),
-        legend.title = element_text(size = 12),
-        legend.text  = element_text(size = 12),
-        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
-        # strip.background = element_rect(size = 0.5),
-        legend.key.size = unit(1, "line")
-        ) +
-  scale_y_continuous("Residual", limits = c(-5,5), breaks = seq(-5,5,2.5)) +
-  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
-  scale_fill_manual("LMER fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA"))  +
-  guides(color = guide_legend(override.aes = list(alpha = 1)),
-         fill = guide_legend(override.aes = list(alpha = 1)))
-
-eyefitting_lmer_plot
-```
-
 Eliminating the linear trend constraint, we can extend the method and analysis beyond ordinary least squares by using generalized additive mixed models (GAMM) to estimate smoothing splines and allow for a more flexible residual trend.
@@
 
-```{r eyefitting-gamm-residualplots, warning = F, message = F}
-#| fig-cap: "Estimated trends of residuals (vertical deviation of participant drawn points from both the OLS (blue) and PCA (orange) fitted points) as fit by the generalized additive mixed model.
-A random sample of 75 participants was selected to display the individual participant residuals behind the overall trend."
-#| out-width: \columnwidth
-#| label: fig-eyefitting-gamm-residualplots
-#| fig-res: 300
-#| fig-width: 9
-#| fig-height: 9
-#| fig-env: figure
-#|
-
-eyefitting_gamm_plot <- eyefitting_gamm_preds %>%
-  filter((`Parameter Choice` %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
-  mutate(`Parameter Choice` = factor(`Parameter Choice`, levels = c("S", "F", "V", "N"))) %>%
-  ggplot(aes(x = x)) +
-  geom_line(data = eyefitting_model_data %>% 
-              filter(prolific_id %in% participant_sample), 
-            aes(x = x, y = residual_ols_loess, group = plot_id, color = "OLS"), alpha = 0.1) +
-  geom_line(data = eyefitting_model_data %>% 
-              filter(prolific_id %in% participant_sample), 
-            aes(x = x, y = residual_pca_loess, group = plot_id, color = "PCA"), alpha = 0.1) +
-  geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.4) +
-  geom_line(aes(y = ols.pred, color = "OLS")) +
-  geom_ribbon(aes(ymin = pca.lower, ymax = pca.upper, fill = "PCA"), color = NA, alpha = 0.4) +
-  geom_line(aes(y = pca.pred, color = "PCA")) +
-  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
-  facet_wrap(~`Parameter Choice`, labeller = labeller(`Parameter Choice` = label_both), scales = "free") +
-  theme_bw(base_size = 14) +
-  theme(aspect.ratio = 1,
-        legend.position = "right",
-        plot.title   = element_text(size = 12, hjust = 0),
-        axis.text    = element_text(size = 12),
-        axis.title   = element_text(size = 12),
-        legend.title = element_text(size = 12),
-        legend.text  = element_text(size = 12),
-        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
-        # strip.background = element_rect(size = 0.5),
-        legend.key.size = unit(1, "line")
-        ) +
-  scale_y_continuous("Residual", limits = c(-5,5), breaks = seq(-5,5,2.5)) +
-  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
-  scale_fill_manual("GAMM fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA"))   +
-  guides(color = guide_legend(override.aes = list(alpha = 1)),
-         fill = guide_legend(override.aes = list(alpha = 1)))
-
-eyefitting_gamm_plot
-```
-
-@fig-eyefitting-lmer-residualplots and @fig-eyefitting-gamm-residualplots show a visual analysis of the estimated trends of residuals (vertical deviation of participant drawn points from both the OLS and PCA fitted points) as modeled by a LMM and GAMM respectively.
+While the statistical models were fit to all four parameter choices, @fig-eyefitting-plots displays the estimated trends of residuals (vertical deviation of participant drawn points from both the OLS and PCA fitted points) as modeled by a LMM and GAMM respectively for parameter choice set F.
 A random sample of 75 participants was selected to display individual participant residuals behind the overall residual trend.
 Examining the plots, the estimated trends of PCA residuals (orange) appear to align more parallel and closer to the $y=0$ horizontal (dashed) line than the OLS residuals (blue). 
-In particular, this trend is more prominent in parameter choices with large variances (F and N).
+In particular, this trend was more prominent in parameter choices with large variances.
 Results from our study were consistent with those found in the original study; when shown points following a linear trend, participants tended to fit the slope of the first principal component Our study established 'You Draw It' as a method for graphical testing and reinforced the differences between intuitive visual model fitting and statistical model fitting, providing information about human perception as it relates to the use of statistical graphics.
@@
 After validating the 'You Draw It' method, we conducted a study utilizing the new method to evaluate participants ability to make forecasts for exponentially increasing data on log and linear scales.
-Along with analyzing the feedback data with the GAMM method for flexibility due to nonlinear data, we used spaghetti plots to conduct a visual analysis of participant forecasts compared to the nonlinear least squares statistical model and make comparisons between two chart design features [@fig-exponential-yloess-spaghetti-plot-2-1].
+Along with analyzing the feedback data with the GAMM method for flexibility due to nonlinear data, we used spaghetti plots to conduct a visual analysis of participant forecasts compared to the nonlinear least squares statistical model and make comparisons between two chart design features [@fig-exponential-spaghetti-plot].
 This study demonstrated the advantage of using the 'You Draw It' method paired with GAMM and visual analysis methods for measuring perception and visual patterns of nonlinear data.
 
-![Spaghetti plot of results from a study which asked participants to forcast trends of exponentially increasing data.
+```{r eyefitting-plots, warning = F, message = F}
+#| fig-cap: "Spaghetti plot of results from a study which asked participants to forecast trends of exponentially increasing data.
 Participants drawn lines on the linear scale are shown in blue and the log scale are shown in orange.
-Variability in the statistically fitted regression lines occured due to a unique data set being simulated for each individual; the gray band shows the range fitted values from the statistically fitted regression lines.](images/exponential-yloess-spaghetti-plot-2-1.png){#fig-exponential-yloess-spaghetti-plot-2-1}
+Variability in the statistically fitted regression lines occurred due to a unique data set being simulated for each individual; the gray band shows the range fitted values from the statistically fitted regression lines."
+#| out-width: \columnwidth
+#| label: fig-exponential-spaghetti-plot
+#| fig-res: 300
+#| fig-width: 6
+#| fig-height: 4
+#| fig-env: figure
+
+spaghetti_plot <- youdrawit_model_data %>%
+  ggplot(aes(x = x)) +
+  geom_line(aes(y = yloess, group = plot_id, color = scale), alpha = 0.2) +
+  geom_ribbon(data = youdrawit_simulated_band, aes(ymin = min_ynls, ymax = max_ynls, fill = "Fitted NLS", group = NA), color = NA, alpha = 0.35) +
+  # facet_grid(`Growth Rate` ~ `Points Truncated`, scales = "free", labeller = labeller(`Growth Rate` = label_both, `Points Truncated` = label_both)) +
+  theme_bw() +
+  theme(aspect.ratio = 1,
+        legend.position = "right"
+        ) +
+  scale_color_manual("", values = c("steelblue", "orange"), labels = c("Visual fit, linear scale \n (drawn, loess)", "Visual fit, log scale \n (drawn, loess)")) +
+  scale_fill_manual("", values = c("black"), labels = c("Range of fitted values \n from statistically fitted regression \n lines (NLS)")) +
+  scale_x_continuous(limits = c(10, 20)) +
+  scale_y_continuous("y", limits = c(0, 300), breaks = seq(0,300,50)) +
+  guides(color = guide_legend(override.aes = list(alpha = 1)),
+         fill = guide_legend(override.aes = list(alpha = 0.3)))
+spaghetti_plot
+```
 

### 2022-07-11 2022-sdss-you-draw-it-manuscript.qmd
added article types

@@
 ---
+
+<!-- Article Types -->
+<!-- https://jds-online.org/journal/JDS/information/Article%20Types -->
+
+<!-- Statistical Data Science -->
+
+<!-- This section is the home base of the reformed journal covering statistical methods that are motivated by real-world applications.
+It is not for papers with technical proofs that push the frontiers of theoretical developments.
+In addition to classic topics in Statistics, cutting-edge works on big data, visualization, machine learning, and artificial intelligence are also welcome.
+-->
+
+<!-- Computing in Data Science -->
+
+<!-- Computing is an indispensable component of all data science and big data applications.
+As more journals on data science and big data emerge, it is of great interest for the data science community to have a highly regarded outlet with a specialization in computing, covering a wide spectrum from methods, algorithms, software implementations, to case studies.
+Existing journals on statistical computing have their own traditions and may not meet the increasing demands.
+Some research works on cutting-edge problems may not fit well in any existing journal.
+-->
+
+<!-- This section covers the following types of articles. -->
+<!-- 1.
+Software: Articles here are similar to those in the Journal of Statistical Software.
+They are not referencing manuals but vignettes that introduce the methods being implemented as well as the usage of the software with reproducible code chunks.
+The software implementation can be in any computer language with a sufficiently large user base.
+-->
+<!-- 2.
+Algorithms: Articles here focus on the performance side of the computing needs arising from domain applications.
+For example, one can propose algorithms that make infeasible tasks feasible or speed up existing algorithms.
+-->
+<!-- 3.
+Methods: Articles here are similar to those in Journal of Computational and Graphic Statistics or Statistics and Computing.
+The computing methods need to be motivated by a domain application with the properties carefully studied.
+-->
 

### 2022-07-27 2022-sdss-you-draw-it-manuscript.qmd
edits. can't get author names added to .qmd

@@
 This research is intended to implement 'You Draw It', adapted from the New York Times, as a way to measure the patterns we see in data."
-
+author:
+  - Emily A. Robinson
 keywords: 'graphics, user interaction, regression'
@@
 ```{r eyefitting-simplot, eval = F, warning = F, message = F}
-#| fig-cap: "Example of simulated data points displayed in a scatter-plot illustrating the trends associated with the four selected parameter choices."
+#| fig-cap: "Example of simulated data points displayed in a scatterplot illustrating the trends associated with the four selected parameter choices."
 #| out-width: \columnwidth
@@
 
-One of the most common charts created is a scatter plot of points over time; these charts show up regularly in news articles and in scientific publications alike.
+One of the most common charts created is a scatterplot of points over time; these charts show up regularly in news articles and in scientific publications alike.
 These charts rely on our ability to identify and detect trends in data. 
@@
 #| fig-cap: "Weekly average gas prices in the United States, 2019-June 2022.
-Additional mental operations a viewer might perform while looking at the plot are annotated in grey.
-@usenergyinformationadministrationWeeklyAllGrades2022"
+Additional mental operations a viewer might perform while looking at the plot are annotated in grey [@usenergyinformationadministrationWeeklyAllGrades2022]."
 #| out-width: \columnwidth
@@
 
-Initial studies in the 20th century explored the use of fitting lines by eye through a set of points [@finney1951subjective, @mosteller1981eye].
+Initial studies in the 20th century explored the use of fitting lines by eye through a set of points [@finney1951subjective; @mosteller1981eye; @unwin1988eyeballing].
 Common methods of fitting trends by eye involved maneuvering a string, black thread, or ruler until the fit was suitable, then drawing the line through the set of points.
-\ear{Results from these early studies provided groundwork for visually judging slopes and selecting accurate starting values for common iterative calculations.} Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.
+Results from these early studies provided groundwork for visually judging slopes and selecting accurate starting values for common iterative calculations.
+Recently, @ciccione2021can conducted a comprehensive set of studies investigating human ability to detect trends in graphical representations from a psychophysical approach.
 <!-- \svp{XXX and found what? Describe the results XXX} -->
-\ear{This set of studies asked participants to judge trends, estimate slopes, and conduct extrapolation by using a track-pad to adjust the tilt of a line on the screen. 
+This set of studies asked participants to judge trends, estimate slopes, and conduct extrapolation by using a track-pad to adjust the tilt of a line on the screen. 
 Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increased with noise and with number of points.
-This supported the results found in @mosteller1981eye and suggested that participants might use Deming regression [@deming1943statistical], which minimizes the Euclidean distance of points from the line, when fitting a line to a noisy scatter-plot.}
+This supported the results found in @mosteller1981eye and suggested that participants might use Deming regression [@deming1943statistical], which minimizes the Euclidean distance of points from the line, when fitting a line to a noisy scatterplot.
 
 While psychologists and statisticians have been using eye-fitting techniques to assess our innate perceptual statistical modeling abilities, news organizations have used similar techniques in order to draw readers in and demonstrate the difference between readers' assumptions and reality.
-In 2015, the New York Times introduced an interactive feature, called 'You Draw It' [@aisch2015you], where readers input their own assumptions about various metrics of news interest and compare these assumptions to reality.
-The New York Times team used Data Driven Documents (D3), a JavaScript library that allows readers to interact with a chart directly by drawing a line on their computer screen with a mouse.
+In 2015, the New York Times (NYT) introduced an interactive feature, called 'You Draw It' [@aisch2015you], where readers input their own assumptions about various metrics of news interest and compare these assumptions to reality.
+The NYT team used Data Driven Documents (D3), a JavaScript library that allows readers to interact with a chart directly by drawing a line on their computer screen with a mouse.
 Despite the somewhat different purpose behind this feature, the D3 driven method used by the NYT is wonderfully intuitive, and does not require the assumption of linearity, making it much more adaptable to testing how viewers perceive and predict when presented with non-linear data.
 
-We set out to implement ‘You Draw It’, adapted from the New York Times feature, as a way to experimentally assess the patterns we see in data. 
+We set out to implement ‘You Draw It’, adapted from the NYT feature, as a way to experimentally assess the patterns we see in data. 
 Here, we provide technical details of the software development, utilizing interactive graphics in R [@r-software]. 
@@
 
-## You Draw It v0
+## 'You Draw It' v0
 
-\svp{Describe the NYT code - what it does, how it works (for loop to connect mouse coordinates to closest point via inverting the scales)}
+<!-- \svp{Describe the NYT code - what it does, how it works (for loop to connect mouse coordinates to closest point via inverting the scales)} -->
 
-The New York Times uses D3, Data Driven Documents [@bostock2011d3], to create many of the interactive data graphics that users rely on, from the famous "election needle" in 2016 to their COVID dashboards.
+The NYT uses D3, Data Driven Documents [@bostock2011d3], to create many of the interactive data graphics that users rely on, from the famous "election needle" in 2016 to their COVID dashboards. 
 Naturally, this same framework was used to create the 'You Draw It' feature.
+The NYT code used to create the initial D3 plot and 'You Draw It' interactive feature includes multiple working parts.
+First, the relationship between coordinate mappings and pixels sets up the chart and the initial data is plotted.
+Next, the `d3.drag` function adds user interactivity and observes the movement of the user's mouse.
+Additionally, the `forEach` loop finds the closest point in the coordinate system to the user's mouse.
+Finally, the code detects if all points are filled by the user and updates the status of the task.
 
-A challenge of working with D3 is the environment necessary to display the graphics and images.
-<!-- \svp{XXX What environment, why is this a challenge, and how did you overcome it?} -->
-\ear{Typically, a server setup is needed in order to render D3 visuals and show output from JavaScript on the webpage.
-This can be a barrier to getting started with D3 for someone unfamiliar with server setup and minimal JavaScript coding background.
-Resources, such as @wattenberger-fullstack, help novice users overcome this challenge by walking them step by step through setting up a local server and creating their first plot in D3.
-In the next section, we will see how the `r2d3` package [@r2d3pkg] in R can be used to display D3 visuals in familiar R HTML formats such as Rmarkdown and Shiny.}
+When working with a new programming language, setting up a new development environment can be challenging.
+As R users, we found the software overhead required to develop straight JavaScript code to be intimidating [@wattenberger-fullstack], but luckily, the R community anticipated this barrier.
+In the next section, we will see how the `r2d3` package [@r2d3pkg] in R can be used to render the JavaScript code and display D3 visuals in familiar R HTML formats such as Rmarkdown and Shiny.
 
@@
 
-\svp{Describe the process of using `r2d3`}
-
-We leveraged the `r2d3` package to take data randomly generated in R and create a D3 plot that could be modified with the You Draw It script.
+<!-- \svp{Describe the process of using `r2d3`} -->
+We leveraged the `r2d3` package to take data randomly generated in R and create a D3 plot that could be modified with the 'You Draw It' script.
 This step allowed us to more easily connect the data generation process with the resulting D3 code without having to generate new code manually each time we re-generated the data.
@@
 
-We conducted all data simulation and processing in R and output two data sets - *point data* and *line data* - containing (x, y) coordinates corresponding to either a simulated point or fitted value predicted by a statistical model respectively.
+We conducted all data simulation and processing in R and output two data sets - *point data* and *line data* - containing $(x, y)$ coordinates corresponding to either a simulated point or fitted value predicted by a statistical model respectively.
 Then, the `r2d3` package converts the data sets in R to JavaScript Object Notation (JSON) to be interpreted by JavaScript code included with the `r2d3` package.
@@
 
-## Modifying You Draw It
+## Adding 'You Draw It' Functionality
 
@@
 When testing perception of graphics, however, we want to pre-populate the chart with additional information - points, and in some cases, portions of a trend line.
-This requires that we modify the original JavaScript \ear{source} code to accommodate these additional elements, which we generate using `r2d3`, as described above.
+This requires that we modify the original JavaScript source code to accommodate these additional elements, which we generate using `r2d3`, as described above.
 
-We defined JavaScript functions to draw the initial plot and set up drawable points for the user drawn line. 
+As previously discussed, we modified `r2d3` code to draw the initial plot and data.
 Drag events in `D3.js` are utilized to observe and react to user input, as in the original 'You Draw It' script from the NYT.
-^[The full source code we developed for the YouDrawIt task is available at \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.]
+^[The full source code we developed for the 'You Draw It' task is available at \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js}.] We adapted initial code from @pearce_2019 and @furic_2017.
+<!-- \ear{Rename js code to you-draw-it}. -->
 
@@
 These functions were modified from the defaults automatically generated by `r2d3` -->
+
+<!-- \ear{Reference https://bl.ocks.org/1wheel/07d9040c3422dac16bd5be741433ff1e and https://github.com/BenoitFuric/you-draw-it-graph.} -->
 
@@
 
-During testing of our modifications to the You Draw It' script, we discovered that frequently the JavaScript code would ``skip" from point to point out of sequence.
+During testing of our modifications to the 'You Draw It' script, we discovered that frequently the JavaScript code would ``skip" from point to point out of sequence.
 This resulted in a jagged line (visually) with missing values in the underlying stored array of data; as a result, the for-loop controlling the user-drawn line never exited and the user's data was not recorded.
-While it would have been possible to fix this using some sort of interpolation algorithm, we did not want to compromise user results by introducing interpolation artifacts, so we opted instead for a visual cue that fixed the problem from a psychological angle.
+While it would have been possible to fix this using some sort of interpolation algorithm, we did not want to compromise user results by introducing interpolation artifacts, so we opted instead for a visual cue that fixed the problem by modifying the human behavior.
 
@@
 One challenge introduced by adding this visual cue is that D3 uses Scalable Vector Graphics (SVG) to render elements. 
-Adding an additional layer meant that we had to ensure that not only the grids, lines, and points of the default scatter plot and trend line rendered in the correct order, but that the yellow box rendered behind all of these points as well so that no information was masked.
+Adding an additional layer meant that we had to ensure that not only the grids, lines, and points of the default scatterplot and trend line rendered in the correct order, but that the yellow box rendered behind all of these points as well so that no information was masked; and that this layer order updated in real time.
 
@@
 Once the user is finished modifying the plot, they can submit their response, so long as they have filled in all of the points.
-This is enforced using Shiny's message-passing interface and a hook in the JavaScript code that notifies Shiny when the array of user-drawn points is completely filled in.
+This is enforced using Shiny's message-passing interface and a JavaScript hook that notifies Shiny when the array of user-drawn points is completely filled in.
 Once the user is done drawing the line, we save the results of the drawn line to a SQLite database, using Shiny to pass data from JavaScript to R. 
@@
 Researchers conducted the study with a latin square experimental design to test for a practice effect on the task.
-Results indicated participants tended to fit the slope of the first principal component (minimizes Euclidean distance) as opposed to the slope from the ordinary least squares regression (minimizes vertical distance) [@fig-pca-plot] and found no effect of order.
+Qualitative analysis suggested that participants tended to fit the slope of the first principal component (minimizes Euclidean distance) as opposed to the slope from the ordinary least squares regression (minimizes vertical distance) [@fig-pca-plot] and found no effect of order.
 
@@
 \end{align} 
-For each participant, four unique data sets were randomly and independently generated from the underlying parameters at the beginning of the study and mapped to a scatter-plot graphic.
+For each participant, four unique data sets were randomly and independently generated from the underlying parameters at the beginning of the study and mapped to a scatterplot graphic.
 <!-- @fig-eyefitting-simplot illustrates an example of simulated data for all four parameter choices intended to reflect the trends in the original study. -->
 When participants started the study, they were first asked to complete two practice plots - accompanied by instructions and a .gif demonstrating the task - in order to train them in the skills associated with executing the task.
-The four 'You Draw It' task plots associated with the selected parameters followed the practice plots in a random order for each individual.
+The four 'You Draw It' task plots associated with the selected parameters followed the practice plots in a random order for each individual; these plots were interspersed with plots from a different experiment.
 
-Participants were recruited via Prolific in March 2022; while this study does utilize a convenience sample, as this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context \ear{(CITE SPATIAL PAPER)}.
+Participants were recruited via Prolific in March 2022; while this study does utilize a convenience sample, as this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
 The study was conducted and distributed via a Shiny application and participants completed the tasks on their own computer, with a computer mouse, and in an environment of their choosing.
@@
 As a part of the validation study, we collected data for each participant drawn trend line, allowing us to conduct formal analyses which compared the intuitive visually fit lines to statistical regression results - the original study lacked these formal comparisons, providing only summaries of averages, variances, and actual (least squares) values for the slope and intercept of each data set.
-The feedback data, stored in a SQLite database, contained the simulated data points} - $(x_{ijk}, y_{ijk})$ -, the predicted values from the statistical regression models - $\hat y_{ijk,OLS}$, and $\hat y_{ijk,PCA}$ -, and the predicted values from the user drawn line - $y_{ijk,drawn}$ for parameter choice $i = 1,2,3,4$, $j = 1,...N_\text{participant}$, and $x_{ijk}$ value $k = 1, ...,4 x_{max} + 1$.
+The feedback data, stored in a SQLite database, contained the simulated data points} - $(x_{ijk}, y_{ijk})$ -, the predicted values from the statistical regression models - $\hat y_{ijk,OLS}$, and $\hat y_{ijk,PCA}$ -, and the predicted values from the user drawn line - $y_{ijk,drawn}$ for parameter choice $i = 1,2,3,4$, $j = 1,...,N_\text{participant}$, and $x_{ijk}$ value $k = 1, ...,4 x_{max} + 1$.
 @fig-eyefitting-plots displays an example of all three fitted trend lines for parameter choice F.
@@
 Eliminating the linear trend constraint, we can extend the method and analysis beyond ordinary least squares by using generalized additive mixed models (GAMM) to estimate smoothing splines and allow for a more flexible residual trend.
-The `bam` function in the `mgcv` package was used to fit a GAMM separately to the OLS and PCA.
+The `bam` function in the `mgcv` [@mgcv_pkg] package was used to fit a GAMM separately to the OLS and PCA.
 Parameter choice was treated as a fixed effect with no estimated intercept and a separate smoothing spline for $x$ was estimated for each parameter choice. 
@@
 
+This method allows us to quantitatively compare participant lines to the OLS and principal component regression lines, which is an improvement over the qualitative approach in @mosteller1981eye.
 While the statistical models were fit to all four parameter choices, @fig-eyefitting-plots displays the estimated trends of residuals (vertical deviation of participant drawn points from both the OLS and PCA fitted points) as modeled by a LMM and GAMM respectively for parameter choice set F.
@@
 
-After validating the 'You Draw It' method, we conducted a study utilizing the new method to evaluate participants ability to make forecasts for exponentially increasing data on log and linear scales.
+The analysis method presented above extends nicely beyond the linear regressions tested in @mosteller1981eye.
+Here we briefly demonstrate this with an example from a study examining participants ability to make forecasts for exponentially increasing data on log and linear scales.
 Along with analyzing the feedback data with the GAMM method for flexibility due to nonlinear data, we used spaghetti plots to conduct a visual analysis of participant forecasts compared to the nonlinear least squares statistical model and make comparisons between two chart design features [@fig-exponential-spaghetti-plot].
-This study demonstrated the advantage of using the 'You Draw It' method paired with GAMM and visual analysis methods for measuring perception and visual patterns of nonlinear data.
+The combination of the GAMM analysis and visual display demonstrates the strength of the 'You Draw It' method for testing statistical graphics.
 
@@
 
-\svp{XXX add references to the anchoring bit XXX}
+<!-- \svp{XXX add references to the anchoring bit XXX} -->
+
+# Supplementary Material {-}
+
++ **'You Draw It' Demonstration Applet:** The shiny app used to demonstrate the 'You Draw It' method can be accessed at  [emily-robinson.shinyapps.io/you-draw-it-validation-applet](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
++ **'You Draw It' Code:** The 'You Draw It' JavaScript code can be accessed at  [github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js](https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js).
++ **Study Applet:** The shiny app used to conduct the study can be accessed at  [shiny.srvanderplas.com/perception-of-statistical-graphics/](https://shiny.srvanderplas.com/perception-of-statistical-graphics/).
++ **Study Applet Code:** The code used to create the RShiny Applet for data collection can be found at [github.com/earobinson95/log-perception-prolific/tree/main/perception-of-statistical-graphics](https://github.com/earobinson95/log-perception-prolific/tree/main/perception-of-statistical-graphics).
++ **Participant Data (Linear):** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/log-perception-prolific/raw/main/analyses/02b-eyefitting/data/eyefitting-model-data.csv](https://github.com/earobinson95/log-perception-prolific/raw/main/analyses/02b-eyefitting/data/eyefitting-model-data.csv).
++ **Data Analysis Code (Linear):** The code used to replicate the linear study analysis in this paper can be found at [github.com/earobinson95/log-perception-prolific/blob/main/analyses/02b-eyefitting/eyefitting-analysis.Rmd](https://github.com/earobinson95/log-perception-prolific/blob/main/analyses/02b-eyefitting/eyefitting-analysis.Rmd).
++ **Participant Data (Non-linear):** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/log-perception-prolific/raw/main/analyses/02a-you-draw-it/data/youdrawit-model-data.csv](https://github.com/earobinson95/log-perception-prolific/raw/main/analyses/02a-you-draw-it/data/youdrawit-model-data.csv).
++ **Data Analysis Code (Non-linear):** The code used to replicate the nonlinear study analysis in this paper can be found at [github.com/earobinson95/log-perception-prolific/blob/main/analyses/02a-you-draw-it/you-draw-it-analysis.Rmd](https://github.com/earobinson95/log-perception-prolific/blob/main/analyses/02a-you-draw-it/you-draw-it-analysis.Rmd).
 

### 2022-07-27 2022-sdss-you-draw-it-manuscript.tex
edits. can't get author names added to .qmd

@@
 \documentclass[
-  letterpaper,
-  DIV=11,
-  numbers=noendperiod]{scrartcl}
+]{jds}
 
@@
 \newcommand{\rh}[1]{{\textcolor{Green}{#1}}}
-% \usepackage[capitalise]{cleveref}
-% \newcommand\pcref[1]{(\cref{#1})}
-
-\usepackage{natbib}
-\bibliographystyle{unsrtnat}
-\KOMAoption{captions}{tableheading}
 \makeatletter
@@
   pdftitle={`You Draw It': Implementation of visually fitted trends with r2d3},
+  pdfauthor={Emily A. Robinson},
   pdfkeywords={graphics, user interaction, regression},
@@
 \texttt{r2d3}}
-\author{}
+\author{Emily A. Robinson}
 \date{}
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, sharp corners, breakable]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-One of the most common charts created is a scatter plot of points over
+One of the most common charts created is a scatterplot of points over
 time; these charts show up regularly in news articles and in scientific
@@
 States, 2019-June 2022. Additional mental operations a viewer might
-perform while looking at the plot are annotated in grey. US Energy
-Information Administration (2022)}
+perform while looking at the plot are annotated in grey (US Energy
+Information Administration 2022).}
 
@@
 Initial studies in the 20th century explored the use of fitting lines by
-eye through a set of points Mosteller et al. (1981). Common methods of
-fitting trends by eye involved maneuvering a string, black thread, or
-ruler until the fit was suitable, then drawing the line through the set
-of points.
-\ear{Results from these early studies provided groundwork for visually judging slopes and selecting accurate starting values for common iterative calculations.}
+eye through a set of points (Finney 1951; Mosteller et al. 1981; Unwin
+and Wills 1988). Common methods of fitting trends by eye involved
+maneuvering a string, black thread, or ruler until the fit was suitable,
+then drawing the line through the set of points. Results from these
+early studies provided groundwork for visually judging slopes and
+selecting accurate starting values for common iterative calculations.
 Recently, Ciccione and Dehaene (2021) conducted a comprehensive set of
 studies investigating human ability to detect trends in graphical
-representations from a psychophysical approach.
-\ear{This set of studies asked participants to judge trends, estimate slopes, and conduct extrapolation by using a track-pad to adjust the tilt of a line on the screen. 
-Results indicated the slopes participants reported were always in excess of the ideal slopes, both in the positive and in the negative direction, and those biases increased with noise and with number of points.
-This supported the results found in @mosteller1981eye and suggested that participants might use Deming regression [@deming1943statistical], which minimizes the Euclidean distance of points from the line, when fitting a line to a noisy scatter-plot.}
+representations from a psychophysical approach. This set of studies
+asked participants to judge trends, estimate slopes, and conduct
+extrapolation by using a track-pad to adjust the tilt of a line on the
+screen. Results indicated the slopes participants reported were always
+in excess of the ideal slopes, both in the positive and in the negative
+direction, and those biases increased with noise and with number of
+points. This supported the results found in Mosteller et al. (1981) and
+suggested that participants might use Deming regression (Deming 1943),
+which minimizes the Euclidean distance of points from the line, when
+fitting a line to a noisy scatterplot.
 
@@
 draw readers in and demonstrate the difference between readers'
-assumptions and reality. In 2015, the New York Times introduced an
+assumptions and reality. In 2015, the New York Times (NYT) introduced an
 interactive feature, called `You Draw It' (Aisch, Cox, and Quealy 2015),
 where readers input their own assumptions about various metrics of news
-interest and compare these assumptions to reality. The New York Times
-team used Data Driven Documents (D3), a JavaScript library that allows
-readers to interact with a chart directly by drawing a line on their
-computer screen with a mouse. Despite the somewhat different purpose
-behind this feature, the D3 driven method used by the NYT is wonderfully
-intuitive, and does not require the assumption of linearity, making it
-much more adaptable to testing how viewers perceive and predict when
-presented with non-linear data.
+interest and compare these assumptions to reality. The NYT team used
+Data Driven Documents (D3), a JavaScript library that allows readers to
+interact with a chart directly by drawing a line on their computer
+screen with a mouse. Despite the somewhat different purpose behind this
+feature, the D3 driven method used by the NYT is wonderfully intuitive,
+and does not require the assumption of linearity, making it much more
+adaptable to testing how viewers perceive and predict when presented
+with non-linear data.
 
-We set out to implement `You Draw It', adapted from the New York Times
-feature, as a way to experimentally assess the patterns we see in data.
-Here, we provide technical details of the software development,
-utilizing interactive graphics in R (R Core Team 2022). We then share
-results from our study which validates `You Draw It' as a method for
-graphical testing on linear trends and apply an appropriate data
-analysis method to the participant data. We also briefly demonstrate the
-use of the `You Draw It' method and analysis on nonlinear data.
+We set out to implement `You Draw It', adapted from the NYT feature, as
+a way to experimentally assess the patterns we see in data. Here, we
+provide technical details of the software development, utilizing
+interactive graphics in R (R Core Team 2022). We then share results from
+our study which validates `You Draw It' as a method for graphical
+testing on linear trends and apply an appropriate data analysis method
+to the participant data. We also briefly demonstrate the use of the `You
+Draw It' method and analysis on nonlinear data.
 
@@
 \hypertarget{you-draw-it-v0}{%
-\subsection{You Draw It v0}\label{you-draw-it-v0}}
+\subsection{`You Draw It' v0}\label{you-draw-it-v0}}
 
-\svp{Describe the NYT code - what it does, how it works (for loop to connect mouse coordinates to closest point via inverting the scales)}
+The NYT uses D3, Data Driven Documents (Bostock, Ogievetsky, and Heer
+2011), to create many of the interactive data graphics that users rely
+on, from the famous ``election needle'' in 2016 to their COVID
+dashboards. Naturally, this same framework was used to create the `You
+Draw It' feature. The NYT code used to create the initial D3 plot and
+`You Draw It' interactive feature includes multiple working parts.
+First, the relationship between coordinate mappings and pixels sets up
+the chart and the initial data is plotted. Next, the \texttt{d3.drag}
+function adds user interactivity and observes the movement of the user's
+mouse. Additionally, the \texttt{forEach} loop finds the closest point
+in the coordinate system to the user's mouse. Finally, the code detects
+if all points are filled by the user and updates the status of the task.
 
-The New York Times uses D3, Data Driven Documents (Bostock, Ogievetsky,
-and Heer 2011), to create many of the interactive data graphics that
-users rely on, from the famous ``election needle'' in 2016 to their
-COVID dashboards. Naturally, this same framework was used to create the
-`You Draw It' feature.
-
-A challenge of working with D3 is the environment necessary to display
-the graphics and images.
-\ear{Typically, a server setup is needed in order to render D3 visuals and show output from JavaScript on the webpage.
-This can be a barrier to getting started with D3 for someone unfamiliar with server setup and minimal JavaScript coding background.
-Resources, such as @wattenberger-fullstack, help novice users overcome this challenge by walking them step by step through setting up a local server and creating their first plot in D3.
-In the next section, we will see how the `r2d3` package [@r2d3pkg] in R can be used to display D3 visuals in familiar R HTML formats such as Rmarkdown and Shiny.}
+When working with a new programming language, setting up a new
+development environment can be challenging. As R users, we found the
+software overhead required to develop straight JavaScript code to be
+intimidating (Wattenberger 2019), but luckily, the R community
+anticipated this barrier. In the next section, we will see how the
+\texttt{r2d3} package (Strayer, Luraschi, and Allaire 2020) in R can be
+used to render the JavaScript code and display D3 visuals in familiar R
+HTML formats such as Rmarkdown and Shiny.
 
@@
 
-\svp{Describe the process of using `r2d3`}
-
 We leveraged the \texttt{r2d3} package to take data randomly generated
-in R and create a D3 plot that could be modified with the You Draw It
+in R and create a D3 plot that could be modified with the `You Draw It'
 script. This step allowed us to more easily connect the data generation
@@
 We conducted all data simulation and processing in R and output two data
-sets - \emph{point data} and \emph{line data} - containing (x, y)
+sets - \emph{point data} and \emph{line data} - containing \((x, y)\)
 coordinates corresponding to either a simulated point or fitted value
@@
 
-\hypertarget{modifying-you-draw-it}{%
-\subsection{Modifying You Draw It}\label{modifying-you-draw-it}}
+\hypertarget{adding-you-draw-it-functionality}{%
+\subsection{Adding `You Draw It'
+Functionality}\label{adding-you-draw-it-functionality}}
 
@@
 portions of a trend line. This requires that we modify the original
-JavaScript \ear{source} code to accommodate these additional elements,
-which we generate using \texttt{r2d3}, as described above.
+JavaScript source code to accommodate these additional elements, which
+we generate using \texttt{r2d3}, as described above.
 
-We defined JavaScript functions to draw the initial plot and set up
-drawable points for the user drawn line. Drag events in \texttt{D3.js}
-are utilized to observe and react to user input, as in the original `You
-Draw It' script from the NYT. \footnote{The full source code we
-  developed for the YouDrawIt task is available at
-  \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/main-d3v5.js}.}
+As previously discussed, we modified \texttt{r2d3} code to draw the
+initial plot and data. Drag events in \texttt{D3.js} are utilized to
+observe and react to user input, as in the original `You Draw It' script
+from the NYT. \footnote{The full source code we developed for the `You
+  Draw It' task is available at
+  \url{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js}.}
+We adapted initial code from Pearce (2019) and Furic (2017).
 
@@
 
-During testing of our modifications to the You Draw It' script, we
+During testing of our modifications to the `You Draw It' script, we
 discovered that frequently the JavaScript code would ``skip'' from point
@@
 compromise user results by introducing interpolation artifacts, so we
-opted instead for a visual cue that fixed the problem from a
-psychological angle.
+opted instead for a visual cue that fixed the problem by modifying the
+human behavior.
 
@@
 layer meant that we had to ensure that not only the grids, lines, and
-points of the default scatter plot and trend line rendered in the
-correct order, but that the yellow box rendered behind all of these
-points as well so that no information was masked.
+points of the default scatterplot and trend line rendered in the correct
+order, but that the yellow box rendered behind all of these points as
+well so that no information was masked; and that this layer order
+updated in real time.
 
@@
 response, so long as they have filled in all of the points. This is
-enforced using Shiny's message-passing interface and a hook in the
-JavaScript code that notifies Shiny when the array of user-drawn points
-is completely filled in. Once the user is done drawing the line, we save
-the results of the drawn line to a SQLite database, using Shiny to pass
-data from JavaScript to R. The connections between each portion of the
-app are shown in Figure~\ref{fig-you-draw-it-code-sketch}.
+enforced using Shiny's message-passing interface and a JavaScript hook
+that notifies Shiny when the array of user-drawn points is completely
+filled in. Once the user is done drawing the line, we save the results
+of the drawn line to a SQLite database, using Shiny to pass data from
+JavaScript to R. The connections between each portion of the app are
+shown in Figure~\ref{fig-you-draw-it-code-sketch}.
 
@@
 middle. Researchers conducted the study with a latin square experimental
-design to test for a practice effect on the task. Results indicated
-participants tended to fit the slope of the first principal component
-(minimizes orthogonal residuals) as opposed to the slope from the
-ordinary least squares regression (minimizes vertical residuals)
-Figure~\ref{fig-pca-plot} and found no effect of order.
+design to test for a practice effect on the task. Qualitative analysis
+suggested that participants tended to fit the slope of the first
+principal component (minimizes Euclidean distance) as opposed to the
+slope from the ordinary least squares regression (minimizes vertical
+distance) Figure~\ref{fig-pca-plot} and found no effect of order.
 
@@
 
-{\centering \includegraphics{./images/fig-pca-plot-1.pdf}
+{\centering \includegraphics[width=0.7\textwidth,height=\textheight]{./images/fig-pca-plot-1.pdf}
 
@@
 \text{with } e_i & \sim N(0, \sigma^2). \nonumber
-\end{align} For each participant, a unique data set was randomly and
-independently generated from the underlying parameters at the beginning
-of the study and mapped to a scatter-plot graphic.
-Figure~\ref{fig-eyefitting-simplot} illustrates an example of simulated
-data for all four parameter choices intended to reflect the trends in
-the original study. When participants started the study, they were first
-asked to complete two practice plots - accompanied by instructions and a
-.gif demonstrating the task - in order to train them in the skills
-associated with executing the task. The four `You Draw It' task plots
-associated with the selected parameters followed the practice plots in a
-random order for each individual.
-
-\begin{figure}
-
-{\centering \includegraphics{./images/fig-eyefitting-simplot-1.pdf}
-
-}
-
-\caption{\label{fig-eyefitting-simplot}Example of simulated data points
-displayed in a scatter-plot illustrating the trends associated with the
-four selected parameter choices.}
-
-\end{figure}
+\end{align} For each participant, four unique data sets were randomly
+and independently generated from the underlying parameters at the
+beginning of the study and mapped to a scatterplot graphic. When
+participants started the study, they were first asked to complete two
+practice plots - accompanied by instructions and a .gif demonstrating
+the task - in order to train them in the skills associated with
+executing the task. The four `You Draw It' task plots associated with
+the selected parameters followed the practice plots in a random order
+for each individual; these plots were interspersed with plots from a
+different experiment.
 
@@
 task, previous results have found few differences between expert and
-non-expert participants in this context {[}CITE SPATIAL PAPER{]}. The
-study was conducted and distributed via a Shiny application and
+non-expert participants in this context (VanderPlas and Hofmann 2015).
+The study was conducted and distributed via a Shiny application and
 participants completed the tasks on their own computer, with a computer
@@
 line - \(y_{ijk,drawn}\) for parameter choice \(i = 1,2,3,4\),
-\(j = 1,...N_\text{participant}\), and \(x_{ijk}\) value
-\(k = 1, ...,4 x_{max} + 1\). Figure~\ref{fig-eyefitting-trial-plot}
-displays an example of all three fitted trend lines for parameter choice
-F.
-
-\begin{figure}
-
-{\centering \includegraphics{./images/fig-eyefitting-trial-plot-1.pdf}
-
-}
-
-\caption{\label{fig-eyefitting-trial-plot}Example of validation feedback
-data where three trend lines show the the OLS fitted, PCA fitted, and
-participant drawn values overlaid on the simulated data points.}
-
-\end{figure}
+\(j = 1,...,N_\text{participant}\), and \(x_{ijk}\) value
+\(k = 1, ...,4 x_{max} + 1\). Figure~\ref{fig-eyefitting-plots} displays
+an example of all three fitted trend lines for parameter choice F.
 
@@
 participant differences.
+
+\begin{figure}
+
+{\centering \includegraphics{./images/fig-eyefitting-plots-1.pdf}
+
+}
+
+\caption{\label{fig-eyefitting-plots}Visual analysis results for
+parameter choice set F. \linebreak \textbf{A.} Example of validation
+feedback data where three trend lines show the the OLS fitted, PCA
+fitted, and participant drawn values overlaid on the simulated data
+points. \linebreak \textbf{B \& C.} Estimated trends of residuals
+(vertical deviation of participant drawn points from both the OLS (blue)
+and PCA (orange) fitted points) as fit by the (B) linear mixed model and
+(C) generalized additive mixed model. A random sample of 75 participants
+was selected to display the individual participant residuals behind the
+overall trend.}
+
+\end{figure}
 
@@
 
-\begin{figure}
-
-{\centering \includegraphics{./images/fig-eyefitting-lmer-residualplots-1.pdf}
-
-}
-
-\caption{\label{fig-eyefitting-lmer-residualplots}Estimated trends of
-residuals (vertical deviation of participant drawn points from both the
-OLS (blue) and PCA (orange) fitted points) as fit by the linear mixed
-model. A random sample of 75 participants was selected to display the
-individual participant residuals behind the overall trend.}
-
-\end{figure}
-
 Eliminating the linear trend constraint, we can extend the method and
@@
 flexible residual trend. The \texttt{bam} function in the \texttt{mgcv}
-package was used to fit a GAMM separately to the OLS and PCA. Parameter
-choice was treated as a fixed effect with no estimated intercept and a
-separate smoothing spline for \(x\) was estimated for each parameter
-choice. A random participant effect accounting for variation due to
-participant and a random spline for each participant accounted for
-variation in spline for each participant. The GAMM equation for each fit
-(OLS and PCA) residuals is given by: \begin{equation}
+(Wood 2017) package was used to fit a GAMM separately to the OLS and
+PCA. Parameter choice was treated as a fixed effect with no estimated
+intercept and a separate smoothing spline for \(x\) was estimated for
+each parameter choice. A random participant effect accounting for
+variation due to participant and a random spline for each participant
+accounted for variation in spline for each participant. The GAMM
+equation for each fit (OLS and PCA) residuals is given by:
+\begin{equation}
 y_{ijk, drawn} - \hat y_{ijk, fit} = e_{ijk,fit} = \alpha_i + s_{i}(x_{ijk}) + p_{j} + s_{j}(x_{ijk})
@@
 
-\begin{figure}
-
-{\centering \includegraphics{./images/fig-eyefitting-gamm-residualplots-1.pdf}
-
-}
-
-\caption{\label{fig-eyefitting-gamm-residualplots}Estimated trends of
+This method allows us to quantitatively compare participant lines to the
+OLS and principal component regression lines, which is an improvement
+over the qualitative approach in Mosteller et al. (1981). While the
+statistical models were fit to all four parameter choices,
+Figure~\ref{fig-eyefitting-plots} displays the estimated trends of
 residuals (vertical deviation of participant drawn points from both the
-OLS (blue) and PCA (orange) fitted points) as fit by the generalized
-additive mixed model. A random sample of 75 participants was selected to
-display the individual participant residuals behind the overall trend.}
-
-\end{figure}
-
-Figure~\ref{fig-eyefitting-lmer-residualplots} and
-Figure~\ref{fig-eyefitting-gamm-residualplots} show a visual analysis of
-the estimated trends of residuals (vertical deviation of participant
-drawn points from both the OLS and PCA fitted points) as modeled by a
-LMM and GAMM respectively. A random sample of 75 participants was
-selected to display individual participant residuals behind the overall
-residual trend. Examining the plots, the estimated trends of PCA
-residuals (orange) appear to align more parallel and closer to the
-\(y=0\) horizontal (dashed) line than the OLS residuals (blue). In
-particular, this trend is more prominent in parameter choices with large
-variances (F and N). Results from our study were consistent with those
-found in the original study; when shown points following a linear trend,
-participants tended to fit the slope of the first principal component
-Our study established `You Draw It' as a method for graphical testing
-and reinforced the differences between intuitive visual model fitting
-and statistical model fitting, providing information about human
-perception as it relates to the use of statistical graphics.
+OLS and PCA fitted points) as modeled by a LMM and GAMM respectively for
+parameter choice set F. A random sample of 75 participants was selected
+to display individual participant residuals behind the overall residual
+trend. Examining the plots, the estimated trends of PCA residuals
+(orange) appear to align more parallel and closer to the \(y=0\)
+horizontal (dashed) line than the OLS residuals (blue). In particular,
+this trend was more prominent in parameter choices with large variances.
+Results from our study were consistent with those found in the original
+study; when shown points following a linear trend, participants tended
+to fit the slope of the first principal component Our study established
+`You Draw It' as a method for graphical testing and reinforced the
+differences between intuitive visual model fitting and statistical model
+fitting, providing information about human perception as it relates to
+the use of statistical graphics.
 
@@
 
-After validating the `You Draw It' method, we conducted a study
-utilizing the new method to evaluate participants ability to make
-forecasts for exponentially increasing data on log and linear scales.
-Along with analyzing the feedback data with the GAMM method for
-flexibility due to nonlinear data, we used spaghetti plots to conduct a
-visual analysis of participant forecasts compared to the nonlinear least
-squares statistical model and make comparisons between two chart design
-features Figure~\ref{fig-exponential-yloess-spaghetti-plot-2-1}. This
-study demonstrated the advantage of using the `You Draw It' method
-paired with GAMM and visual analysis methods for measuring perception
-and visual patterns of nonlinear data.
+The analysis method presented above extends nicely beyond the linear
+regressions tested in Mosteller et al. (1981). Here we briefly
+demonstrate this with an example from a study examining participants
+ability to make forecasts for exponentially increasing data on log and
+linear scales. Along with analyzing the feedback data with the GAMM
+method for flexibility due to nonlinear data, we used spaghetti plots to
+conduct a visual analysis of participant forecasts compared to the
+nonlinear least squares statistical model and make comparisons between
+two chart design features Figure~\ref{fig-exponential-spaghetti-plot}.
+The combination of the GAMM analysis and visual display demonstrates the
+strength of the `You Draw It' method for testing statistical graphics.
 
@@
 
-{\centering \includegraphics{images/exponential-yloess-spaghetti-plot-2-1.png}
+{\centering \includegraphics{./images/fig-exponential-spaghetti-plot-1.pdf}
 
@@
 
-\caption{\label{fig-exponential-yloess-spaghetti-plot-2-1}Spaghetti plot
-of results from a study which asked participants to forcast trends of
+\caption{\label{fig-exponential-spaghetti-plot}Spaghetti plot of results
+from a study which asked participants to forecast trends of
 exponentially increasing data. Participants drawn lines on the linear
 scale are shown in blue and the log scale are shown in orange.
-Variability in the statistically fitted regression lines occured due to
+Variability in the statistically fitted regression lines occurred due to
 a unique data set being simulated for each individual; the gray band
@@
 
-\svp{XXX add references to the anchoring bit XXX}
+\hypertarget{supplementary-material}{%
+\section*{Supplementary Material}\label{supplementary-material}}
+\addcontentsline{toc}{section}{Supplementary Material}
+
+\begin{itemize}
+\tightlist
+\item
+  \textbf{`You Draw It' Demonstration Applet:} The shiny app used to
+  demonstrate the `You Draw It' method can be accessed at
+  \href{https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/}{emily-robinson.shinyapps.io/you-draw-it-validation-applet}.
+\item
+  \textbf{`You Draw It' Code:} The `You Draw It' JavaScript code can be
+  accessed at
+  \href{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js}{github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js}.
+\item
+  \textbf{Study Applet:} The shiny app used to conduct the study can be
+  accessed at
+  \href{https://shiny.srvanderplas.com/perception-of-statistical-graphics/}{shiny.srvanderplas.com/perception-of-statistical-graphics/}.
+\item
+  \textbf{Study Applet Code:} The code used to create the RShiny Applet
+  for data collection can be found at
+  \href{https://github.com/earobinson95/log-perception-prolific/tree/main/perception-of-statistical-graphics}{github.com/earobinson95/log-perception-prolific/tree/main/perception-of-statistical-graphics}.
+\item
+  \textbf{Participant Data (Linear):} De-identified participant data
+  collected in the study and used for analyses are available to be
+  downloaded from GitHub at
+  \href{https://github.com/earobinson95/log-perception-prolific/raw/main/analyses/02b-eyefitting/data/eyefitting-model-data.csv}{github.com/earobinson95/log-perception-prolific/raw/main/analyses/02b-eyefitting/data/eyefitting-model-data.csv}.
+\item
+  \textbf{Data Analysis Code (Linear):} The code used to replicate the
+  linear study analysis in this paper can be found at
+  \href{https://github.com/earobinson95/log-perception-prolific/blob/main/analyses/02b-eyefitting/eyefitting-analysis.Rmd}{github.com/earobinson95/log-perception-prolific/blob/main/analyses/02b-eyefitting/eyefitting-analysis.Rmd}.
+\item
+  \textbf{Participant Data (Non-linear):} De-identified participant data
+  collected in the study and used for analyses are available to be
+  downloaded from GitHub at
+  \href{https://github.com/earobinson95/log-perception-prolific/raw/main/analyses/02a-you-draw-it/data/youdrawit-model-data.csv}{github.com/earobinson95/log-perception-prolific/raw/main/analyses/02a-you-draw-it/data/youdrawit-model-data.csv}.
+\item
+  \textbf{Data Analysis Code (Non-linear):} The code used to replicate
+  the nonlinear study analysis in this paper can be found at
+  \href{https://github.com/earobinson95/log-perception-prolific/blob/main/analyses/02a-you-draw-it/you-draw-it-analysis.Rmd}{github.com/earobinson95/log-perception-prolific/blob/main/analyses/02a-you-draw-it/you-draw-it-analysis.Rmd}.
+\end{itemize}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-deming1943statistical}{}}%
+Deming, William Edwards. 1943. {``Statistical Adjustment of Data.''}
+
 \leavevmode\vadjust pre{\hypertarget{ref-finney1951subjective}{}}%
@@
 Series B (Methodological)} 13 (2): 284--97.
+
+\leavevmode\vadjust pre{\hypertarget{ref-furic_2017}{}}%
+Furic, Benoit. 2017. {``You-Draw-It-Graph: The User Can Draw His Own
+Graph Line and Compare with the Reality.''} \emph{GitHub}.
+\url{https://github.com/BenoitFuric/you-draw-it-graph}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-pearce_2019}{}}%
+Pearce, Adam. 2019. {``You-Draw-It.''} \emph{Observable}.
+\url{https://bl.ocks.org/1wheel/07d9040c3422dac16bd5be741433ff1e}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-r-software}{}}%
@@
 \url{https://www.R-project.org/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-r2d3pkg}{}}%
+Strayer, Nick, Javier Luraschi, and JJ Allaire. 2020. \emph{R2d3:
+Interface to 'D3' Visualizations}.
+\url{https://CRAN.R-project.org/package=r2d3}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-unwin1988eyeballing}{}}%
+Unwin, Antony R, and Graham Wills. 1988. {``Eyeballing Time Series.''}
+In \emph{Proceedings of the 1988 ASA Statistical Computing Section},
+263--68.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-vanderplas2015spatial}{}}%
+VanderPlas, Susan, and Heike Hofmann. 2015. {``Spatial Reasoning and
+Data Displays.''} \emph{IEEE Transactions on Visualization and Computer
+Graphics} 22 (1): 459--68.
+
+\leavevmode\vadjust pre{\hypertarget{ref-wattenberger-fullstack}{}}%
+Wattenberger, Amelia. 2019. \emph{Fullstack D3 and Data Visualization:
+Build Beautiful Data Visualizations and Dashboards with D3}.
+Fullstack.io.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mgcv_pkg}{}}%
+Wood, S. N. 2017. \emph{Generalized Additive Models: An Introduction
+with r}. 2nd ed. Chapman; Hall/CRC.
+
 \end{CSLReferences}

### 2022-07-27 2022-sdss-you-draw-it-manuscript.qmd
added analysis

@@
   - Emily A. Robinson
+  - Reka Howard
+  - Susan VanderPlas
 keywords: 'graphics, user interaction, regression'
@@
 + **'You Draw It' Demonstration Applet:** The shiny app used to demonstrate the 'You Draw It' method can be accessed at  [emily-robinson.shinyapps.io/you-draw-it-validation-applet](https://emily-robinson.shinyapps.io/you-draw-it-validation-applet/).
-+ **'You Draw It' Code:** The 'You Draw It' JavaScript code can be accessed at  [github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js](https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js).
++ **'You Draw It' Script:** The 'You Draw It' JavaScript code can be accessed at  [github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js](https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js).
 + **Study Applet:** The shiny app used to conduct the study can be accessed at  [shiny.srvanderplas.com/perception-of-statistical-graphics/](https://shiny.srvanderplas.com/perception-of-statistical-graphics/).
 + **Study Applet Code:** The code used to create the RShiny Applet for data collection can be found at [github.com/earobinson95/log-perception-prolific/tree/main/perception-of-statistical-graphics](https://github.com/earobinson95/log-perception-prolific/tree/main/perception-of-statistical-graphics).
-+ **Participant Data (Linear):** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/log-perception-prolific/raw/main/analyses/02b-eyefitting/data/eyefitting-model-data.csv](https://github.com/earobinson95/log-perception-prolific/raw/main/analyses/02b-eyefitting/data/eyefitting-model-data.csv).
-+ **Data Analysis Code (Linear):** The code used to replicate the linear study analysis in this paper can be found at [github.com/earobinson95/log-perception-prolific/blob/main/analyses/02b-eyefitting/eyefitting-analysis.Rmd](https://github.com/earobinson95/log-perception-prolific/blob/main/analyses/02b-eyefitting/eyefitting-analysis.Rmd).
-+ **Participant Data (Non-linear):** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/log-perception-prolific/raw/main/analyses/02a-you-draw-it/data/youdrawit-model-data.csv](https://github.com/earobinson95/log-perception-prolific/raw/main/analyses/02a-you-draw-it/data/youdrawit-model-data.csv).
-+ **Data Analysis Code (Non-linear):** The code used to replicate the nonlinear study analysis in this paper can be found at [github.com/earobinson95/log-perception-prolific/blob/main/analyses/02a-you-draw-it/you-draw-it-analysis.Rmd](https://github.com/earobinson95/log-perception-prolific/blob/main/analyses/02a-you-draw-it/you-draw-it-analysis.Rmd).
++ **Participant Data (Linear):** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv](https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv).
++ **Participant Data (Nonlinear):** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv](https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv).
++ **Data Analysis Code:** The code used to replicate the nonlinear study analysis in this paper can be found at [earobinson95.github.io/sdss-2022-you-draw-it/analysis/data-analysis.html](earobinson95.github.io/sdss-2022-you-draw-it/analysis/data-analysis.html).
 

### 2022-07-27 analysis/data-analysis.Rmd
added analysis

---
title: "Analysis"
subtitle: "'You Draw It': Implementation of visually fitted trends with `r2d3`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE,
	cache = TRUE
)
library(tidyverse)
library(digest)
library(pls)
library(mcr)

library(lme4)
library(emmeans)

library(openssl)
library(mgcv)
library(lme4)
library(tictoc)
```

# Linear Study

Read in data for linear study.

```{r}
# read in eyefitting_model_data and eyefitting_simulated_data
eyefitting_model_data     <- read_csv("https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv")
eyefitting_simulated_data <- read_csv("https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-simulated-data.csv")
```

There are `r length(unique(eyefitting_model_data$prolific_id)) %>% as.numeric` participants and `r length(unique(eyefitting_model_data$plot_id)) %>% as.numeric()` 'You Draw It' Eyefitting task plots complete.

```{r}
eyefitting_model_data %>%
  select(participant_id, parm_id, x, yols, ypca, ydrawn, yloess, residual_ols_drawn, residual_pca_drawn) %>%
  head() %>%
  knitr::kable(digits = 3, caption = "Eyefitting Model Data (head)")
```

## Example

Look at example from linear data.

```{r}
# example
eyefitting_example <- eyefitting_model_data %>%
  # filter(participant_id == "65c10d498eae365e108efcd3dcb75287", parm_id == "N") %>%
  filter(participant_id == "60b16b9bd5a122c1457d31055df51a45", parm_id == "F") %>%
  ggplot(aes(x = x)) +
  geom_line(aes(y = yols, color = "OLS", linetype = "OLS")) +
  geom_line(aes(y = ypca, color = "PCA", linetype = "PCA")) +
  geom_line(aes(y = ydrawn, color = "Drawn", linetype = "Drawn")) +
  geom_point(data = eyefitting_simulated_data %>%
               filter(dataset == "point_data", participant_id == "60b16b9bd5a122c1457d31055df51a45", parm_id == "F"),
             aes(x = x, y = y)) +
  facet_wrap(~parm_id, labeller = labeller(parm_id = label_both)) +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1,
        legend.position = "bottom",
        plot.title   = element_text(size = 14, hjust = 0),
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text  = element_text(size = 12),
        # strip.text = element_text(size = 8, margin = margin(0.1,0,0.1,0, "cm")),
        # strip.background = element_rect(size = 0.8),
        legend.key.size = unit(1, "line")
        ) +
  scale_x_continuous(limits = c(0,20)) +
  scale_color_manual("", values = c("black", "steelblue", "orange")) +
  scale_linetype_manual("", values = c("dashed", "solid", "solid")) +
  scale_y_continuous("y")
eyefitting_example
```

# Linear Constraint (LMER)

Run lmer model on linear data and obtain predictions.

```{r}
# OLS
# tic()
eyefitting.ols.lmer <- lmer(residual_ols_drawn ~ -1 + parm_id + x:parm_id + (1|participant_id),
                            data = eyefitting_model_data %>%
                             mutate(parm_id = factor(parm_id),
                                    participant_id = factor(participant_id)))
# toc()
# summary(eyefitting.ols.lmer)
# anova(eyefitting.ols.lmer)

# pca
# tic()
eyefitting.pca.lmer <- lmer(residual_pca_drawn ~ -1 + parm_id + x:parm_id + (1|participant_id),
                            data = eyefitting_model_data %>%
                             mutate(parm_id = factor(parm_id),
                                    participant_id = factor(participant_id)))
# toc()
# summary(eyefitting.pca.lmer)
# anova(eyefitting.pca.lmer)

# Obtain Predictions
eyefitting.ols.grid.lmer  <- ref_grid(eyefitting.ols.lmer, at = list(x = seq(1,20,0.5)))
eyefitting.ols.preds.lmer <- emmeans(eyefitting.ols.grid.lmer, ~ parm_id:x) %>%
  as_tibble()

eyefitting.pca.grid.lmer  <- ref_grid(eyefitting.pca.lmer, at = list(x = seq(1,20,0.5)))
eyefitting.pca.preds.lmer <- emmeans(eyefitting.pca.grid.lmer, ~ parm_id:x) %>%
  as_tibble()

eyefitting_lmer_preds <- eyefitting.ols.preds.lmer %>%
  full_join(eyefitting.pca.preds.lmer, by = c("x", "parm_id"), suffix = c(".ols", ".pca"))

# write.csv(eyefitting.preds.lmer, file = here("analyses/02b-eyefitting/data/eyefitting-lmer-preds.csv"), row.names = F, na = "")
```

Create plot for lmer model for linear data.

```{r}
set.seed(68505)
participant_sample <- sample(unique(eyefitting_model_data$prolific_id), 75)

# Plot Predictions
eyefitting_lmer_plot <- eyefitting_lmer_preds %>%
  filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
  filter(parm_id == "F") %>%
  ggplot(aes(x = x)) +
  geom_line(data = eyefitting_model_data %>% 
              filter(prolific_id %in% participant_sample)  %>% filter(parm_id == "F"), 
            aes(x = x, y = residual_ols_loess, group = plot_id, color = "OLS"), alpha = 0.1) +
  geom_line(data = eyefitting_model_data %>% 
              filter(prolific_id %in% participant_sample) %>% filter(parm_id == "F"), 
            aes(x = x, y = residual_pca_loess, group = plot_id, color = "PCA"), alpha = 0.1) +
  geom_ribbon(aes(ymin = asymp.LCL.ols, ymax = asymp.UCL.ols, fill = "OLS"), color = NA, alpha = 0.4) +
  geom_line(aes(y = emmean.ols, color = "OLS")) +
  geom_ribbon(aes(ymin = asymp.LCL.pca, ymax = asymp.UCL.pca, fill = "PCA"), color = NA, alpha = 0.4) +
  geom_line(aes(y = emmean.pca, color = "PCA")) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~parm_id, labeller = labeller(parm_id = label_both), scales = "free") +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1,
        legend.position = "bottom",
        plot.title   = element_text(size = 14, hjust = 0),
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text  = element_text(size = 12),
        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
        # strip.background = element_rect(size = 0.5),
        legend.key.size = unit(1, "line"), legend.box="vertical", legend.margin=margin()
        ) +
  scale_y_continuous("Residual", limits = c(-5,5), breaks = seq(-5,5,2.5)) +
  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
  scale_fill_manual("LMER fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA"))  +
  guides(color = guide_legend(override.aes = list(alpha = 1)),
         fill = guide_legend(override.aes = list(alpha = 1)))

eyefitting_lmer_plot
```

# Smoothing Splines (GAMM)

Run gamm model on linear data.

```{r}
# OLS
# tic()
eyefitting.ols.gamm <- bam(residual_ols_drawn ~ -1 + parm_id + 
                             s(x, by = parm_id) +
                             s(participant_id, bs = "re") +
                             s(x,participant_id, bs = "re"),
                           method = "REML",
                           data = eyefitting_model_data %>%
                             mutate(parm_id = factor(parm_id),
                                    participant_id = factor(participant_id)))
# toc()
# summary(eyefitting.ols.gamm)
# anova(eyefitting.ols.gamm)

# pca
# tic()
eyefitting.pca.gamm <- bam(residual_pca_drawn ~ -1 + parm_id + 
                             s(x, by = parm_id) +
                             s(participant_id, bs = "re") +
                             s(x,participant_id, bs = "re"),
                           method = "REML",
                           data = eyefitting_model_data %>%
                             mutate(parm_id = factor(parm_id),
                                    participant_id = factor(participant_id)))
# toc()
# summary(eyefitting.pca.gamm)
# anova(eyefitting.pca.gamm)
```

Prediction function for GAMM model.

```{r}
#' library(lme4)
#' library(mgcv)
#' lmer_model <- lmer(Reaction ~ Days + (Days || Subject), data = sleepstudy)
#' ga_model <- gam(Reaction ~ Days + s(Subject, bs = "re") + s(Days, Subject, bs = "re"),
#'   data = sleepstudy,
#'   method = "REML"
#' )
#'
#' head(
#'   data.frame(
#'     lmer = predict(lmer_model),
#'     gam = predict_gamm(ga_model)
#'   )
#' )
#'
#' head(
#'   cbind(
#'     lmer = predict(lmer_model, re.form = NA),
#'     gam1 = predict_gamm(ga_model, re_form = NA),
#'     gam2 = predict_gamm(ga_model,
#'       exclude = c("s(Subject)", "s(Days,Subject)")
#'     )
#'   )
#' )
#'
#' head(predict_gamm(ga_model, se = TRUE))
#' @export
predict_gamm <- function(
  model,
  newdata,
  re_form = NULL,
  se = FALSE,
  include = NULL,
  exclude = NULL,
  keep_prediction_data = FALSE,
  ...) {

  # Note because predict doesn't use NULL, can't use NULL for new_data arg or
  # even a differently named arg, and I'm not going into the weeds of rlang to
  # find a hack.

  # basic checks
  if (!inherits(model, "gam")) stop("Need a gam object.")

  if (!rlang::is_null(include) && !rlang::is_character(include)) {
    stop("include must be NULL or character.")
  }

  if (!rlang::is_null(exclude) && !rlang::is_character(exclude)) {
    stop("exclude must be NULL or character.")
  }

  if (!rlang::is_null(re_form) &&
      !rlang::is_na(re_form) &
      !rlang::is_character(re_form)) {
    stop("re_form must be NULL, NA, or character.")
  }

  if (any(include %in% exclude)) {
    stop("You can't include and exclude the same thing.")
  }

  if (!rlang::is_logical(se)) {
    stop("se must be TRUE or FALSE")
  }

  if (!rlang::is_logical(keep_prediction_data)) {
    stop("keep_prediction_data must be TRUE or FALSE")
  }

  # standard prediction would simply call predict.gam
  if (rlang::is_null(re_form) | rlang::is_character(re_form)) {
    if (rlang::is_null(re_form)) {
      preds <- predict(model,
                       newdata,
                       se = se,
                       terms = include,
                       exclude = exclude,
                       ...
      )
    } else {
      preds <- predict(model,
                       newdata,
                       se = se,
                       terms = c(include, re_form),
                       exclude = exclude,
                       ...
      )
    }
  } else if (rlang::is_na(re_form)) {

    # FE only
    re_terms <- sapply(model$smooth, function(x) inherits(x, "random.effect"))
    re_terms <- sapply(model$smooth[re_terms], function(x) x$label)

    preds <- predict(model,
                     newdata,
                     se = se,
                     terms = include,
                     exclude = c(re_terms, exclude),
                     ...
    )
  }

  if (se) {
    preds <- data.frame(prediction = preds$fit, se = preds$se)
  } else {
    preds <- data.frame(prediction = preds)
  }

  if (keep_prediction_data) {
    if (missing(newdata)) {
      base <- model$model
    } else {
      base <- newdata
    }
    preds <- data.frame(base, preds)
  }

  preds
}

```

Obtain predictions for linear data from gamm model.

```{r}
# Obtain Predictions
eyefitting.grid.gamm <- expand_grid(parm_id = c("S", "V", "F", "N"),
                                    x = seq(0,20, 0.5),
                                    participant_id = eyefitting_model_data$participant_id[1])

# OLS
eyefitting.ols.preds <- predict_gamm(eyefitting.ols.gamm, newdata = eyefitting.grid.gamm, se = T, re_form = NA)
eyefitting.grid.gamm$ols.pred <- eyefitting.ols.preds$prediction
eyefitting.grid.gamm$ols.lower <- eyefitting.ols.preds$prediction - (1.96 * eyefitting.ols.preds$se)
eyefitting.grid.gamm$ols.upper <- eyefitting.ols.preds$prediction + (1.96 * eyefitting.ols.preds$se)

# pca
eyefitting.pca.preds <- predict_gamm(eyefitting.pca.gamm, newdata = eyefitting.grid.gamm, se = T, re_form = NA)
eyefitting.grid.gamm$pca.pred <- eyefitting.pca.preds$prediction
eyefitting.grid.gamm$pca.lower <- eyefitting.pca.preds$prediction - (1.96 * eyefitting.pca.preds$se)
eyefitting.grid.gamm$pca.upper <- eyefitting.pca.preds$prediction + (1.96 * eyefitting.pca.preds$se)

eyefitting_gamm_preds <- eyefitting.grid.gamm

# write.csv(eyefitting.grid.gamm, file = here("analyses/02b-eyefitting/data/eyefitting-gamm-preds.csv"), row.names = F, na = "")
```

Create gamm plot for linear data.

```{r}
eyefitting_gamm_plot <- eyefitting_gamm_preds %>%
  filter((parm_id %in% c("F", "N", "S") | (x <= 16 & x >= 4))) %>%
  mutate(parm_id = factor(parm_id, levels = c("S", "F", "V", "N"))) %>%
  filter(parm_id == "F") %>%
  ggplot(aes(x = x)) +
  geom_line(data = eyefitting_model_data %>% 
              filter(prolific_id %in% participant_sample)  %>% filter(parm_id == "F"), 
            aes(x = x, y = residual_ols_loess, group = plot_id, color = "OLS"), alpha = 0.1) +
  geom_line(data = eyefitting_model_data %>% 
              filter(prolific_id %in% participant_sample)  %>% filter(parm_id == "F"), 
            aes(x = x, y = residual_pca_loess, group = plot_id, color = "PCA"), alpha = 0.1) +
  geom_ribbon(aes(ymin = ols.lower, ymax = ols.upper, fill = "OLS"), color = NA, alpha = 0.4) +
  geom_line(aes(y = ols.pred, color = "OLS")) +
  geom_ribbon(aes(ymin = pca.lower, ymax = pca.upper, fill = "PCA"), color = NA, alpha = 0.4) +
  geom_line(aes(y = pca.pred, color = "PCA")) +
  geom_hline(yintercept = 0, linetype = "dashed", alpha = 0.5) +
  facet_wrap(~parm_id, labeller = labeller(parm_id = label_both), scales = "free") +
  theme_bw(base_size = 14) +
  theme(aspect.ratio = 1,
        legend.position = "bottom",
        plot.title   = element_text(size = 14, hjust = 0),
        axis.text    = element_text(size = 12),
        axis.title   = element_text(size = 12),
        legend.title = element_text(size = 12),
        legend.text  = element_text(size = 12),
        # strip.text = element_text(size = 5, margin = margin(0.05,0,0.05,0, "cm")),
        # strip.background = element_rect(size = 0.5),
        legend.key.size = unit(1, "line"), legend.box="vertical", legend.margin=margin()
        ) +
  scale_y_continuous("Residual", limits = c(-5,5), breaks = seq(-5,5,2.5)) +
  scale_color_manual("Individual participant \nresiduals", values = c("steelblue", "orange"), labels = c("OLS", "PCA")) +
  scale_fill_manual("GAMM fitted trend", values = c("steelblue", "orange"), labels = c("OLS", "PCA"))   +
  guides(color = guide_legend(override.aes = list(alpha = 1)),
         fill = guide_legend(override.aes = list(alpha = 1)))

eyefitting_gamm_plot
```

# Nonlinear Study

Read in youdrawit_model_data and eyefitting_simulated_data.

```{r}
youdrawit_model_data <- read.csv("https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv")

factorCols <- c("nick_name", "ip_address", "prolific_id", "age", "gender", "academic_study", "computer_mouse", "recruitment", "participant_id", "plot_id", "beta", "points_truncated", "scale")
youdrawit_model_data[,factorCols] <- lapply(youdrawit_model_data[,factorCols], factor)

youdrawit_model_data %>%
  select(participant_id, beta, points_truncated, scale, x, ynls, ydrawn, yloess, residual_nls_drawn) %>%
  head() %>%
  knitr::kable()

youdrawit_model_data      <- youdrawit_model_data  %>%
  mutate(`Points Truncated` = ifelse(points_truncated == 10, "50%", "75%"),
         `Growth Rate` = ifelse(beta == "beta0.1", "Low", "High")) %>%
  mutate(`Growth Rate` = factor(`Growth Rate`, levels = c("Low", "High"))) %>%
  filter(`Points Truncated` == "50%", `Growth Rate` == "High")

youdrawit_simulated_band <- youdrawit_model_data %>%
  group_by(`Growth Rate`, `Points Truncated`, x) %>%
  summarize(min_ynls = min(ynls),
            max_ynls = max(ynls))
```

There are `r length(unique(youdrawit_model_data$participant_id)) %>% as.numeric` participants and `r length(unique(youdrawit_model_data$plot_id)) %>% as.numeric()` 'You Draw It' task plots complete.

Create spaghetti plot for visual analysis.

```{r}

spaghetti_plot <- youdrawit_model_data %>%
  ggplot(aes(x = x)) +
  geom_line(aes(y = yloess, group = plot_id, color = scale), alpha = 0.2) +
  geom_ribbon(data = youdrawit_simulated_band, aes(ymin = min_ynls, ymax = max_ynls, fill = "Fitted NLS", group = NA), color = NA, alpha = 0.35) +
  # facet_grid(`Growth Rate` ~ `Points Truncated`, scales = "free", labeller = labeller(`Growth Rate` = label_both, `Points Truncated` = label_both)) +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "right"
        ) +
  scale_color_manual("", values = c("steelblue", "orange"), labels = c("Visual fit, linear scale \n (drawn, loess)", "Visual fit, log scale \n (drawn, loess)")) +
  scale_fill_manual("", values = c("black"), labels = c("Range of fitted values \n from statistically fitted regression \n lines (NLS)")) +
  scale_x_continuous(limits = c(10, 20)) +
  scale_y_continuous("y", limits = c(0, 300), breaks = seq(0,300,50)) +
  guides(color = guide_legend(override.aes = list(alpha = 1)),
         fill = guide_legend(override.aes = list(alpha = 0.3)))
spaghetti_plot
```

### 2022-07-27 2022-sdss-you-draw-it-manuscript.qmd
updated analysis link

@@
 This research is intended to implement 'You Draw It', adapted from the New York Times, as a way to measure the patterns we see in data."
-author: 
-  - Emily A. Robinson
-  - Reka Howard
-  - Susan VanderPlas
 keywords: 'graphics, user interaction, regression'
@@
 + **Participant Data (Nonlinear):** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv](https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv).
-+ **Data Analysis Code:** The code used to replicate the nonlinear study analysis in this paper can be found at [earobinson95.github.io/sdss-2022-you-draw-it/analysis/data-analysis.html](earobinson95.github.io/sdss-2022-you-draw-it/analysis/data-analysis.html).
++ **Data Analysis Code:** The code used to replicate the nonlinear study analysis in this paper can be found at [earobinson95.github.io/sdss-2022-you-draw-it-manuscript/analysis/data-analysis.htmll](https://earobinson95.github.io/sdss-2022-you-draw-it-manuscript/analysis/data-analysis.html).
 

### 2022-07-27 2022-sdss-you-draw-it-manuscript.tex
updated analysis link

@@
   pdftitle={`You Draw It': Implementation of visually fitted trends with r2d3},
-  pdfauthor={Emily A. Robinson},
   pdfkeywords={graphics, user interaction, regression},
@@
 \texttt{r2d3}}
-\author{Emily A. Robinson}
+\author{}
 \date{}
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[enhanced, borderline west={3pt}{0pt}{shadecolor}, frame hidden, breakable, sharp corners, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi
 
@@
 \item
-  \textbf{`You Draw It' Code:} The `You Draw It' JavaScript code can be
-  accessed at
+  \textbf{`You Draw It' Script:} The `You Draw It' JavaScript code can
+  be accessed at
   \href{https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js}{github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js}.
@@
   downloaded from GitHub at
-  \href{https://github.com/earobinson95/log-perception-prolific/raw/main/analyses/02b-eyefitting/data/eyefitting-model-data.csv}{github.com/earobinson95/log-perception-prolific/raw/main/analyses/02b-eyefitting/data/eyefitting-model-data.csv}.
+  \href{https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv}{github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv}.
 \item
-  \textbf{Data Analysis Code (Linear):} The code used to replicate the
-  linear study analysis in this paper can be found at
-  \href{https://github.com/earobinson95/log-perception-prolific/blob/main/analyses/02b-eyefitting/eyefitting-analysis.Rmd}{github.com/earobinson95/log-perception-prolific/blob/main/analyses/02b-eyefitting/eyefitting-analysis.Rmd}.
-\item
-  \textbf{Participant Data (Non-linear):} De-identified participant data
+  \textbf{Participant Data (Nonlinear):} De-identified participant data
   collected in the study and used for analyses are available to be
   downloaded from GitHub at
-  \href{https://github.com/earobinson95/log-perception-prolific/raw/main/analyses/02a-you-draw-it/data/youdrawit-model-data.csv}{github.com/earobinson95/log-perception-prolific/raw/main/analyses/02a-you-draw-it/data/youdrawit-model-data.csv}.
+  \href{https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv}{github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv}.
 \item
-  \textbf{Data Analysis Code (Non-linear):} The code used to replicate
-  the nonlinear study analysis in this paper can be found at
-  \href{https://github.com/earobinson95/log-perception-prolific/blob/main/analyses/02a-you-draw-it/you-draw-it-analysis.Rmd}{github.com/earobinson95/log-perception-prolific/blob/main/analyses/02a-you-draw-it/you-draw-it-analysis.Rmd}.
+  \textbf{Data Analysis Code:} The code used to replicate the nonlinear
+  study analysis in this paper can be found at
+  \href{https://earobinson95.github.io/sdss-2022-you-draw-it-manuscript/analysis/data-analysis.html}{earobinson95.github.io/sdss-2022-you-draw-it-manuscript/analysis/data-analysis.htmll}.
 \end{itemize}

### 2022-07-27 2022-sdss-you-draw-it-manuscript.tex
added author information

@@
 
-\usepackage[dvipsnames]{xcolor} % colors
-\newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
-\newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
-\newcommand{\rh}[1]{{\textcolor{Green}{#1}}}
+\author[1]{Emily A. Robinson\thanks{Corresponding author. Email: erobin17@calpoly.edu}}
+\author[2]{Reka Howard\footnote{Email: rekahoward@unl.edu}}
+\author[2]{Susan VanderPlas\footnote{Email: susan.vanderplas@unl.edu}}
+\affil[1]{Department of Statistics, CalPoly-San Luis Obispo}
+\affil[2]{Department of Statistics, University of Nebraska-Lincoln}
 \makeatletter
@@
 \texttt{r2d3}}
-\author{}
+% \author{}
 \date{}
@@
 research is intended to implement `You Draw It', adapted from the New
-York Times, as a way to measure the patterns we see in data.
+York Times, as a way to measure the patterns we see in data. In this
+paper, we describe the adaptation of an old tool for graphical testing
+and evaluation, eye-fitting, for use in modern web-applications suitable
+for testing statistical graphics. We present an empirical evaluation of
+this testing method for linear regression, and briefly discuss an
+extension of this method to non-linear applications.
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, breakable, frame hidden, borderline west={3pt}{0pt}{shadecolor}, enhanced, boxrule=0pt, sharp corners]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, breakable, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 to the participant data. We also briefly demonstrate the use of the `You
-Draw It' method and analysis on nonlinear data.
+Draw It' method and analysis on non-linear data.
 
@@
 
-\hypertarget{extension-to-nonlinear-data}{%
-\subsection{Extension to Nonlinear
-Data}\label{extension-to-nonlinear-data}}
+\hypertarget{extension-to-non-linear-data}{%
+\subsection{Extension to Non-linear
+Data}\label{extension-to-non-linear-data}}
 
@@
 linear scales. Along with analyzing the feedback data with the GAMM
-method for flexibility due to nonlinear data, we used spaghetti plots to
-conduct a visual analysis of participant forecasts compared to the
-nonlinear least squares statistical model and make comparisons between
+method for flexibility due to non-linear data, we used spaghetti plots
+to conduct a visual analysis of participant forecasts compared to the
+non-linear least squares statistical model and make comparisons between
 two chart design features Figure~\ref{fig-exponential-spaghetti-plot}.
@@
 \item
-  \textbf{Study Applet:} The shiny app used to conduct the study can be
-  accessed at
+  \textbf{Study Applet:} The RShiny applet used to conduct the linear
+  and non-linear studies can be accessed at
   \href{https://shiny.srvanderplas.com/perception-of-statistical-graphics/}{shiny.srvanderplas.com/perception-of-statistical-graphics/}.
 \item
-  \textbf{Study Applet Code:} The code used to create the RShiny Applet
+  \textbf{Study Applet Code:} The code used to create the RShiny applet
   for data collection can be found at
@@
   \textbf{Participant Data (Linear):} De-identified participant data
-  collected in the study and used for analyses are available to be
-  downloaded from GitHub at
+  collected in the linear study and used for analyses are available to
+  be downloaded from GitHub at
   \href{https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv}{github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv}.
 \item
-  \textbf{Participant Data (Nonlinear):} De-identified participant data
-  collected in the study and used for analyses are available to be
-  downloaded from GitHub at
+  \textbf{Participant Data (Non-linear):} De-identified participant data
+  collected in the non-linear study and used for analyses are available
+  to be downloaded from GitHub at
   \href{https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv}{github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv}.
 \item
-  \textbf{Data Analysis Code:} The code used to replicate the nonlinear
-  study analysis in this paper can be found at
+  \textbf{Data Analysis Code:} The code used to replicate the linear and
+  non-linear study analysis in this paper can be found at
   \href{https://earobinson95.github.io/sdss-2022-you-draw-it-manuscript/analysis/data-analysis.html}{earobinson95.github.io/sdss-2022-you-draw-it-manuscript/analysis/data-analysis.htmll}.

### 2022-11-01 revisions/response-to-reviewers.qmd
working on revisions

---
title: "Response to reviewers of JDS2207-018 "
subtitle: "'You Draw It': Implementation of visually fitted trends with `r2d3`"
author: "Emily A. Robinson, Reka Howard, and Susan VanderPlas"
date: "2022"
# output: pdf_document
format: 
  pdf:
    include-in-header:
      - revision-edits.tex
---

## Reviewer(s)' Comments to Author

All responses to reviewers' comments are written in \response{blue}.

<!-- \todo{Need to address yet.} \dubcheck{Double check / review.} -->

### Review: Associate Editor

This is a nicely written paper about an implementation of a tool to assess visually fitted trends. This tool was inspired by a New York Times feature "You Draw It" and was developed using the r2d3 R package. The authors explain the motivation for developing the tool, provide computing details, and present results from an experiment studying how people fit trends by visualizing them on a scatter plot.

\todo{The primary element missing from the paper is how this tool and insights from the experiment might be useful for the reader. One context that came to mind was classroom use, where instructors could use the tool in lessons about multivariable relationships and regression. It would be beneficial for the authors to include more detail about how this might be used in practice.}

Some minor edits:

+ Figure 1: The gray is hard to see, particularly for the fitted line. It would also be helpful to add more details to the to the labels on the plot or to the caption to clarify what the plot is showing. Right now it is unclear unless without reading the paragraph description in the body of the text.
  + \response{Figure 1's readability was increased by changing the text on the graphic to black, darkening the gray and alpha blending of the shaded rectangle regions, making the points open circles rather than closed, and darking the gray and increasing the alpha blending of the fitted line.}
  + \response{We elaborated in the caption for figure 1 to clarify what the plot is showing: "Demonstration of a viewer's engagement with a scatterplot showing the weekly average gas prices in the United States, from 2019-June 2022. The veiwer might perform several cognitive operations such as assessing points to determine outliers, fitting a rough mental smooth/trend line (overlaying solid gray line), and using additional contextual information from long term memory to explain the variation (shaded rectangular gray regions annotated in black text)."}

+ pg. 11 - Please provide a brief description of Prolific for readers unfamiliar with company.
  + \response{A breif description of prolific, "a crowd sourcing platform for researchers to recruit participants for their online studies," was added to the Validation Study section.}

### Review: Guest Editor for the SDSS Special Issue of the JDS

I agree with the AE’s decision of “Requires Major Revision”.

+ \todo{In addition to the AE’s comments, another possible application is using the tool for staff training. There is a growing demand to train up staff given the current job market.}

+ Another comment is that I recommend the authors update figure 2 too (AE commented about figure 1), because the gray text is hard to see. 
  + \response{Figure 2 is a screenshot directly from the New York Times website used to demonstrate the use of the 'You Draw It' feature for interactive reading. While we could try to recreate this image or modify the colors, this changes the actual figure shown to readers in the New York Times.}

+ Also, figures 3 and 6 are difficult to read given there are three subplots. If there is space, perhaps the authors can reconfigure the plots, such as creating two rows (i.e., one plot in one row, and two plots in one row). 
  + \response{Figure 3 was reconstructed to provide the three subplots in vertical order spanning an entire page for better visability. As these subplots have a natural flow, we did not want to interupt this by creating a 2 + 1 subplot figure structure.}
  + \response{Figure 6 was reconstructed to provide a 1 + 2 subplot figure structure, keeping the two predicted plots from different models next to each other in the second row.}

+ Further, please note that all figures should be generated as vector images to ensure they’re high quality for final journal publication.
  + \response{Figure 1, Figure 5, and Figure 6 were generated within R and saved directly to pdf files from within the R environment.}
  + \response{Figure 2 and Figure 3 were screenshots taken from websites. We zoomed in on a high quality screen and took a screenshot; saved as high resolution png files.}
  + \response{Figure 4 was sketched using an online tool; saving this as an svg changes the appearance of the font type and we would like to keep the sketched font. We saved this at the highest resoultion the tool allowed as a png file.}

### 2022-11-01 original-2022-sdss-you-draw-it-manuscript.qmd
working on revisions

--- 2022-07-27 2022-sdss-you-draw-it-manuscript.qmd
@@
 In 2015, the New York Times introduced an interactive feature, called 'You Draw It,' where readers are asked to input their own assumptions about various metrics and compare how these assumptions relate to reality.
-This research is intended to implement 'You Draw It', adapted from the New York Times, as a way to measure the patterns we see in data."
+This research is intended to implement 'You Draw It', adapted from the New York Times, as a way to measure the patterns we see in data.
+In this paper, we describe the adaptation of an old tool for graphical testing and evaluation, eye-fitting, for use in modern web-applications suitable for testing statistical graphics.
+We present an empirical evaluation of this testing method for linear regression, and briefly discuss an extension of this method to non-linear applications."
 keywords: 'graphics, user interaction, regression'
@@
       - edits.tex
+      - authors.tex
       # - refs.tex
@@
 We then share results from our study which validates 'You Draw It' as a method for graphical testing on linear trends and apply an appropriate data analysis method to the participant data.
-We also briefly demonstrate the use of the 'You Draw It' method and analysis on nonlinear data.
+We also briefly demonstrate the use of the 'You Draw It' method and analysis on non-linear data.
 
@@
 Don't be afraid to quote Mosteller's paper if you think it's necessary.
-Focus on one set of plots from the Mosteller paper and one set of plots from our exponential stuff, so you're demonstrating linear and nonlinear analysis.
+Focus on one set of plots from the Mosteller paper and one set of plots from our exponential stuff, so you're demonstrating linear and non-linear analysis.
 Include basic demographics, and compare to Mosteller's paper - we also had a sample of highly educated individuals.
@@
 
-## Extension to Nonlinear Data
+## Extension to Non-linear Data
 
@@
 Here we briefly demonstrate this with an example from a study examining participants ability to make forecasts for exponentially increasing data on log and linear scales.
-Along with analyzing the feedback data with the GAMM method for flexibility due to nonlinear data, we used spaghetti plots to conduct a visual analysis of participant forecasts compared to the nonlinear least squares statistical model and make comparisons between two chart design features [@fig-exponential-spaghetti-plot].
+Along with analyzing the feedback data with the GAMM method for flexibility due to non-linear data, we used spaghetti plots to conduct a visual analysis of participant forecasts compared to the non-linear least squares statistical model and make comparisons between two chart design features [@fig-exponential-spaghetti-plot].
 The combination of the GAMM analysis and visual display demonstrates the strength of the 'You Draw It' method for testing statistical graphics.
@@
 + **'You Draw It' Script:** The 'You Draw It' JavaScript code can be accessed at  [github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js](https://github.com/earobinson95/presentations/blob/master/can-you-draw-it/www/you-draw-it.js).
-+ **Study Applet:** The shiny app used to conduct the study can be accessed at  [shiny.srvanderplas.com/perception-of-statistical-graphics/](https://shiny.srvanderplas.com/perception-of-statistical-graphics/).
-+ **Study Applet Code:** The code used to create the RShiny Applet for data collection can be found at [github.com/earobinson95/log-perception-prolific/tree/main/perception-of-statistical-graphics](https://github.com/earobinson95/log-perception-prolific/tree/main/perception-of-statistical-graphics).
-+ **Participant Data (Linear):** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv](https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv).
-+ **Participant Data (Nonlinear):** De-identified participant data collected in the study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv](https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv).
-+ **Data Analysis Code:** The code used to replicate the nonlinear study analysis in this paper can be found at [earobinson95.github.io/sdss-2022-you-draw-it-manuscript/analysis/data-analysis.htmll](https://earobinson95.github.io/sdss-2022-you-draw-it-manuscript/analysis/data-analysis.html).
++ **Study Applet:** The RShiny applet used to conduct the linear and non-linear studies can be accessed at  [shiny.srvanderplas.com/perception-of-statistical-graphics/](https://shiny.srvanderplas.com/perception-of-statistical-graphics/).
++ **Study Applet Code:** The code used to create the RShiny applet for data collection can be found at [github.com/earobinson95/log-perception-prolific/tree/main/perception-of-statistical-graphics](https://github.com/earobinson95/log-perception-prolific/tree/main/perception-of-statistical-graphics).
++ **Participant Data (Linear):** De-identified participant data collected in the linear study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv](https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/eyefitting-model-data.csv).
++ **Participant Data (Non-linear):** De-identified participant data collected in the non-linear study and used for analyses are available to be downloaded from GitHub at  [github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv](https://github.com/earobinson95/sdss-2022-you-draw-it-manuscript/raw/master/data/youdrawit-model-data.csv).
++ **Data Analysis Code:** The code used to replicate the linear and non-linear study analysis in this paper can be found at [earobinson95.github.io/sdss-2022-you-draw-it-manuscript/analysis/data-analysis.htmll](https://earobinson95.github.io/sdss-2022-you-draw-it-manuscript/analysis/data-analysis.html).
 

### 2022-11-01 original-2022-sdss-you-draw-it-manuscript.tex
working on revisions

--- 2022-07-27 2022-sdss-you-draw-it-manuscript.tex
@@
 
+\usepackage[dvipsnames]{xcolor} % colors
+\newcommand{\ear}[1]{{\textcolor{blue}{#1}}}
+\newcommand{\svp}[1]{{\textcolor{RedOrange}{#1}}}
+\newcommand{\rh}[1]{{\textcolor{Green}{#1}}}
 \author[1]{Emily A. Robinson\thanks{Corresponding author. Email: erobin17@calpoly.edu}}
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, breakable, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[borderline west={3pt}{0pt}{shadecolor}, frame hidden, enhanced, breakable, sharp corners, interior hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 

### 2022-11-01 revisions/revision-diffs.R

@@ -1,0 +1,34 @@
+library(latexdiffr)
+
+# first revision comparison -----------------------
+
+latexdiff(
+  "original-submission-Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex",
+  "intermediate-revisions-Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex",
+  output = "diff-1",
+  open = TRUE,
+  clean = TRUE,
+  quiet = TRUE,
+  output_format = NULL,
+  ld_opts = "--replace-context2cmd=\"none\""
+)
+
+# second revision comparison -------------------------
+
+latexdiff(
+  "intermediate-revisions-Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex",
+  "Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex",
+  output = "manuscript-revisions/diff-2",
+  open = TRUE,
+  clean = TRUE,
+  quiet = TRUE,
+  output_format = NULL,
+  ld_opts = "--replace-context2cmd=\"none\""
+)
+
+# in terminal
+
+# latexdiff intermediate-revisions-Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex Eye-Fitting-Straight-Lines-in-the-Modern-Era.tex > diff-2.tex
+
+
+

### 2022-11-12 revisions/response-to-reviewers.qmd
added discussion section. needs to be reviewed/edited.

@@
 
-\todo{The primary element missing from the paper is how this tool and insights from the experiment might be useful for the reader.
+The primary element missing from the paper is how this tool and insights from the experiment might be useful for the reader.
 One context that came to mind was classroom use, where instructors could use the tool in lessons about multivariable relationships and regression.
-It would be beneficial for the authors to include more detail about how this might be used in practice.}
+It would be beneficial for the authors to include more detail about how this might be used in practice.
+
++ \response{We added a discussion section to address the appllicaiton of teaching and training by connecting the idea of the tool being used for interactive reading (NYT), interactive testing (our study), and interactive teaching/training (future applicaiton).}
 
@@
 
-+ \todo{In addition to the AE’s comments, another possible application is using the tool for staff training. There is a growing demand to train up staff given the current job market.}
++ In addition to the AE’s comments, another possible application is using the tool for staff training. There is a growing demand to train up staff given the current job market.
+  + \response{We added a discussion section to address the appllicaiton of teaching and training by connecting the idea of the tool being used for interactive reading (NYT), interactive testing (our study), and interactive teaching/training (future applicaiton).}
 
@@
 + Further, please note that all figures should be generated as vector images to ensure they’re high quality for final journal publication.
-  + \response{Figure 1, Figure 5, and Figure 6 were generated within R and saved directly to pdf files from within the R environment.}
+  + \response{Figure 1, Figure 5, and Figure 6 were generated within R and saved directly to pdf files from within the R environment.
+We expect these are all vector compatible; please let us know if this isn't sufficient or if we should use a different graphics device.}
   + \response{Figure 2 and Figure 3 were screenshots taken from websites. We zoomed in on a high quality screen and took a screenshot; saved as high resolution png files.}

### 2022-11-12 2022-sdss-you-draw-it-manuscript.qmd
added discussion section. needs to be reviewed/edited.

@@
   theme(aspect.ratio = 1,
-        legend.position = "bottom",
+        legend.position = "right",
         plot.title   = element_text(size = 14, hjust = 0),
@@
 ```{r gas-prices, echo = F, message = F, warning = F}
-#| fig-cap: "Weekly average gas prices in the United States, 2019-June 2022.
-Additional mental operations a viewer might perform while looking at the plot are annotated in grey [@usenergyinformationadministrationWeeklyAllGrades2022]."
+#| fig-cap: "Demonstration of a viewer's engagement with a scatterplot showing the weekly average gas prices in the United States, from 2019-June 2022.
+The veiwer might perform several cognitive operations such as assessing points to determine outliers, fitting a rough mental smooth/trend line (overlaying solid gray line), and using additional contextual information from long term memory to explain the variation (shaded rectangular gray regions annotated in black text) [@usenergyinformationadministrationWeeklyAllGrades2022].
+"
 #| out-width: \columnwidth
@@
 ggplot(gas, aes(x = Date, y = Price)) + 
-  geom_rect(data = annotations, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), alpha= 0.2, fill = "grey", inherit.aes = F) + 
-  geom_text(data = annotations, aes(x = xmin, y = 1, label = label), color = "grey") + 
+  geom_rect(data = annotations, aes(xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax), alpha= 0.3, fill = "gray30", inherit.aes = F) + 
+  geom_text(data = annotations, aes(x = xmin, y = 1, label = label), color = "black") + 
   ylab("Price per Gallon ($)") + 
   ggtitle("Weekly Average Gas Prices in the United States, 2019 - 2022", subtitle = "US Energy Information Administration") + 
-  geom_point() + 
+  geom_point(shape = 1) + 
   scale_x_datetime(limits = as.POSIXct(ymd("2019-01-01", "2022-06-29"))) + 
-  geom_line(se = F, color = "grey", method = "loess", span = .1, alpha = .5, stat = "smooth", size = 1.5)
+  geom_line(se = F, color = "grey30", method = "loess", span = .1, alpha = .6, stat = "smooth", size = 1.5)
 ```
@@
 
-![You Draw It task plot as shown to user.    
-**left:** Initial state, with instructions *Use your mouse to fill in the trend in the yellow box region.*    
-**middle:** User view during task completion.    
-**right:** Finished state](images/ydi-stimuli.png){#fig-you-draw-it-task-plot}
+<!-- ![You Draw It task plot as shown to user.     -->
+<!-- **left:** Initial state, with instructions *Use your mouse to fill in the trend in the yellow box region.*     -->
+<!-- **middle:** User view during task completion.     -->
+<!-- **right:** Finished state](images/ydi-stimuli.png){#fig-you-draw-it-task-plot} -->
+
+```{r ydi-stimuli}
+#| fig-cap: "You Draw It task plot as shown to user.
+\\textbf{top:} Initial state, with instructions, \\textit{Use your mouse to fill in the trend in the yellow box region.} \\textbf{middle:} User view during task completion.
+\\textbf{bottom:} Finished state."
+#| out-width: "70%"
+#| label: fig-you-draw-it-task-plot
+#| fig-res: 300
+#| fig-width: 3.3
+#| fig-height: 10
+#| fig-env: figure
+#| warning: false
+#| message: false
+
+library(cowplot)
+frame1 <- ggdraw() + draw_image("images/ydi-stimuli-frame1.PNG")
+frame2 <- ggdraw() + draw_image("images/ydi-stimuli-frame2.PNG")
+frame3 <- ggdraw() + draw_image("images/ydi-stimuli-frame3.PNG")
+
+frame1 + frame2 + frame3 + plot_layout(ncol = 1)
+```
 
@@
 
-![Sketch of underlying code for 'You Draw It', illustrating the data simulation conducted in R, the initial setup of the visual stimuli with D3 source code, along with the iterative process between the user interaction and plotting in Shiny.
-Once the user is done drawing the line, we saved the results of the drawn line to a database for analysis.](images/code-sketch-2.png){#fig-you-draw-it-code-sketch}
+```{r, code sketch}
+#| fig-cap: "Sketch of underlying code for 'You Draw It', illustrating the data simulation conducted in R, the initial setup of the visual stimuli with D3 source code, along with the iterative process between the user interaction and plotting in Shiny.
+Once the user is done drawing the line, we saved the results of the drawn line to a database for analysis."
+#| out-width: \columnwidth
+#| label: fig-you-draw-it-code-sketch
+#| fig-res: 300
+#| fig-width: 12
+#| fig-height: 6
+#| fig-env: figure
+knitr::include_graphics("images/code-sketch.png")
+# knitr::include_graphics("images/code-sketch.svg")
+```
 
@@
 
-Participants were recruited via Prolific in March 2022; while this study does utilize a convenience sample, as this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
+Participants were recruited via Prolific, a crowd sourcing platform for researchers to recruit participants for their online studies, in March 2022; while this study does utilize a convenience sample, as this is primarily a perceptual task, previous results have found few differences between expert and non-expert participants in this context [@vanderplas2015spatial].
 The study was conducted and distributed via a Shiny application and participants completed the tasks on their own computer, with a computer mouse, and in an environment of their choosing.
@@
 #| fig-res: 300
-#| fig-width: 12
-#| fig-height: 6
+#| fig-width: 9
+#| fig-height: 11
 #| fig-env: figure
 
-(eyefitting_example + ggtitle("A. Example")) + 
-  (eyefitting_lmer_plot + ggtitle("B. LMER Results")) + 
-  (eyefitting_gamm_plot + ggtitle("C. GAMM Results")) +
-  plot_layout(ncol =  3)
+((eyefitting_example + 
+    ggtitle("A. Example") + 
+    theme(plot.title = element_text(hjust = 1/3), legend.position = "bottom"))
+ ) /
+ (((eyefitting_lmer_plot + 
+      ggtitle("B. LMER Results")  +
+     theme(legend.position = "bottom")
+    ) + 
+  (eyefitting_gamm_plot + 
+     ggtitle("C. GAMM Results")  +
+     theme(legend.position = "bottom")
+   )
+  )
+  # plot_layout(guides = 'collect')
+  ) +
+  plot_layout(widths = c(1, 2))
 ```
@@
 
+# Discussion
+
+The 'You Draw It' graphical task presented in this work is an example of an interactive graphic. 
+While any communication between user and device can be considered an interaction, interactive graphics should be defined as a user-driven direct manipulation of plots and plotting elements with an immediate reaction (Unwin 1999, Becker et al.
+1987, Eick & Wills 1995).
+Here we discuss the use of 'You Draw It' as a tool for interactive reading, interactive testing, and interactive teaching as a way to increase user experience and user engagement with data.
+The New York Times first introduced the 'You Draw It' feature as a tool for interactive reading.
+The inclusion of interactive graphics in news and media have been used to inform readers in efforts to reduce misconceptions and aid in storytelling [@geidner2015role; @george2011interactive; @george2013storytelling].
+In our work, we demonstrated the use of the 'You Draw It' method as a tool for interactive testing of graphics.
+We collect data of the user drawn lines and compare to statistical results in order to understand perceptual biases.
+While this work has many possible directions for future work, there is a strong application to extend the 'You Draw It' tool for interactive teaching.
+The line of best fit is introduced in secondary schools as a student's first encounter with the concept of statistical association [@casey2016students].
+@casey2015examining and @casey2016finding conducted task based interviews to examine student's conceptions of the line of best fit prior to introducing the topic.
+The implications of using interactive teaching tools for the line of best fit can help students make connections between the shift in slope and the location of the best fit line [@nagle2017slope].
+@casey2015teachers conducted similar task based interviews wish pre-service mathematics teachers in order to understand the implications regarding teacher preparation for the teaching of this topic. 
+This demonstrates the use of the 'You Draw It' interactive tool for training purposes both within a classroom setting and for those in the workforce.
+
 # Conclusion

### 2022-11-12 2022-sdss-you-draw-it-manuscript.tex
added discussion section. needs to be reviewed/edited.

@@
 \texttt{r2d3}}
-% \author{}
+\author{}
 \date{}
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, breakable, enhanced, sharp corners, borderline west={3pt}{0pt}{shadecolor}, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
 
@@
 
-\caption{\label{fig-gas-prices}Weekly average gas prices in the United
-States, 2019-June 2022. Additional mental operations a viewer might
-perform while looking at the plot are annotated in grey (US Energy
-Information Administration 2022).}
+\caption{\label{fig-gas-prices}Demonstration of a viewer's engagement
+with a scatterplot showing the weekly average gas prices in the United
+States, from 2019-June 2022. The veiwer might perform several cognitive
+operations such as assessing points to determine outliers, fitting a
+rough mental smooth/trend line (overlaying solid gray line), and using
+additional contextual information from long term memory to explain the
+variation (shaded rectangular gray regions annotated in black text) (US
+Energy Information Administration 2022).}
 
@@
 
-{\centering \includegraphics{images/ydi-stimuli.png}
+{\centering \includegraphics[width=0.7\textwidth,height=\textheight]{./images/fig-you-draw-it-task-plot-1.pdf}
 
@@
 \caption{\label{fig-you-draw-it-task-plot}You Draw It task plot as shown
-to user.\\
-\textbf{left:} Initial state, with instructions \emph{Use your mouse to
-fill in the trend in the yellow box region.}\\
-\textbf{middle:} User view during task completion.\\
-\textbf{right:} Finished state}
+to user. \textbf{top:} Initial state, with instructions,
+\textit{Use your mouse to fill in the trend in the yellow box region.}
+\textbf{middle:} User view during task completion. \textbf{bottom:}
+Finished state.}
 
@@
 
-{\centering \includegraphics{images/code-sketch-2.png}
+{\centering \includegraphics{images/code-sketch.png}
 
@@
 
-Participants were recruited via Prolific in March 2022; while this study
-does utilize a convenience sample, as this is primarily a perceptual
-task, previous results have found few differences between expert and
-non-expert participants in this context (VanderPlas and Hofmann 2015).
-The study was conducted and distributed via a Shiny application and
-participants completed the tasks on their own computer, with a computer
-mouse, and in an environment of their choosing. The data from this study
-were collected to validate this method of graphical testing, with the
-hopes of providing a new tool to assess graphical perception
-interactively.
+Participants were recruited via Prolific, a crowd sourcing platform for
+researchers to recruit participants for their online studies, in March
+2022; while this study does utilize a convenience sample, as this is
+primarily a perceptual task, previous results have found few differences
+between expert and non-expert participants in this context (VanderPlas
+and Hofmann 2015). The study was conducted and distributed via a Shiny
+application and participants completed the tasks on their own computer,
+with a computer mouse, and in an environment of their choosing. The data
+from this study were collected to validate this method of graphical
+testing, with the hopes of providing a new tool to assess graphical
+perception interactively.
 
@@
 
+\hypertarget{discussion}{%
+\section{Discussion}\label{discussion}}
+
+The `You Draw It' graphical task presented in this work is an example of
+an interactive graphic. While any communication between user and device
+can be considered an interaction, interactive graphics should be defined
+as a user-driven direct manipulation of plots and plotting elements with
+an immediate reaction (Unwin 1999, Becker et al.~1987, Eick \& Wills
+1995). Here we discuss the use of `You Draw It' as a tool for
+interactive reading, interactive testing, and interactive teaching as a
+way to increase user experience and user engagement with data. The New
+York Times first introduced the `You Draw It' feature as a tool for
+interactive reading. The inclusion of interactive graphics in news and
+media have been used to inform readers in efforts to reduce
+misconceptions and aid in storytelling (Geidner et al. 2015;
+George-Palilonis and Spillman 2011, 2013). In our work, we demonstrated
+the use of the `You Draw It' method as a tool for interactive testing of
+graphics. We collect data of the user drawn lines and compare to
+statistical results in order to understand perceptual biases. While this
+work has many possible directions for future work, there is a strong
+application to extend the `You Draw It' tool for interactive teaching.
+The line of best fit is introduced in secondary schools as a student's
+first encounter with the concept of statistical association (Casey and
+Nagle 2016). Casey (2015) and Casey (2016) conducted task based
+interviews to examine student's conceptions of the line of best fit
+prior to introducing the topic. The implications of using interactive
+teaching tools for the line of best fit can help students make
+connections between the shift in slope and the location of the best fit
+line (Nagle, Casey, and Moore-Russo 2017). Casey and Wasserman (2015)
+conducted similar task based interviews wish pre-service mathematics
+teachers in order to understand the implications regarding teacher
+preparation for the teaching of this topic. This demonstrates the use of
+the `You Draw It' interactive tool for training purposes both within a
+classroom setting and for those in the workforce.
+
 \hypertarget{conclusion}{%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-casey2015examining}{}}%
+Casey, Stephanie A. 2015. {``Examining Student Conceptions of
+Covariation: A Focus on the Line of Best Fit.''} \emph{Journal of
+Statistics Education} 23 (1).
+
+\leavevmode\vadjust pre{\hypertarget{ref-casey2016finding}{}}%
+---------. 2016. {``Finding What Fits.''} \emph{Mathematics Teaching in
+the Middle School} 21 (8): 484--91.
+
+\leavevmode\vadjust pre{\hypertarget{ref-casey2016students}{}}%
+Casey, Stephanie A, and Courtney Nagle. 2016. {``Students' Use of Slope
+Conceptualizations When Reasoning about the Line of Best Fit.''}
+\emph{Educational Studies in Mathematics} 92 (2): 163--77.
+
+\leavevmode\vadjust pre{\hypertarget{ref-casey2015teachers}{}}%
+Casey, Stephanie A, and Nicholas H Wasserman. 2015.
+{``TEACHERS'KNOWLEDGE ABOUT INFORMAL LINE OF BEST FIT.''}
+\emph{Statistics Education Research Journal} 14 (1): 8--35.
+
 \leavevmode\vadjust pre{\hypertarget{ref-ciccione2021can}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-geidner2015role}{}}%
+Geidner, Nick, Ivanka Pjesivac, Iveta Imre, Ioana Coman, and Dzmitry
+Yuran. 2015. {``The Role of Interactive Graphics in Reducing
+Misperceptions in the Electorate.''} \emph{Visual Communication
+Quarterly} 22 (3): 133--45.
+
+\leavevmode\vadjust pre{\hypertarget{ref-george2011interactive}{}}%
+George-Palilonis, Jennifer, and Mary Spillman. 2011. {``Interactive
+Graphics Development: A Framework for Studying Innovative Visual Story
+Forms.''} \emph{Visual Communication Quarterly} 18 (3): 167--77.
+
+\leavevmode\vadjust pre{\hypertarget{ref-george2013storytelling}{}}%
+---------. 2013. {``Storytelling with Interactive Graphics: An Analysis
+of Editors' Attitudes and Practices.''} \emph{Visual Communication
+Quarterly} 20 (1): 20--27.
+
 \leavevmode\vadjust pre{\hypertarget{ref-mosteller1981eye}{}}%
@@
 35 (3): 150--52.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nagle2017slope}{}}%
+Nagle, Courtney, Stephanie Casey, and Deborah Moore-Russo. 2017.
+{``Slope and Line of Best Fit: A Transfer of Knowledge Case Study.''}
+\emph{School Science and Mathematics} 117 (1-2): 13--26.
 

### 2022-11-14 2022-sdss-you-draw-it-manuscript.qmd
edited discussion

@@
 
-The 'You Draw It' graphical task presented in this work is an example of an interactive graphic. 
-While any communication between user and device can be considered an interaction, interactive graphics should be defined as a user-driven direct manipulation of plots and plotting elements with an immediate reaction (Unwin 1999, Becker et al.
-1987, Eick & Wills 1995).
+The 'You Draw It' graphical task presented in this work is an example of an interactive graphic; an interactive graphic is defined as a user-driven direct manipulation of the plot followed by an immediate reaction [@unwin1999requirements; @eick1995high; @becker1987dynamic].
 Here we discuss the use of 'You Draw It' as a tool for interactive reading, interactive testing, and interactive teaching as a way to increase user experience and user engagement with data.
@@
 In our work, we demonstrated the use of the 'You Draw It' method as a tool for interactive testing of graphics.
-We collect data of the user drawn lines and compare to statistical results in order to understand perceptual biases.
+We collected and compared data from user drawn lines to statistical results in order to understand perceptual biases.
 While this work has many possible directions for future work, there is a strong application to extend the 'You Draw It' tool for interactive teaching.
-The line of best fit is introduced in secondary schools as a student's first encounter with the concept of statistical association [@casey2016students].
+The line of best fit is introduced in secondary schools as a student's first encounter with the concept of statistical association between multiple variables [@casey2016students].
 @casey2015examining and @casey2016finding conducted task based interviews to examine student's conceptions of the line of best fit prior to introducing the topic.
 The implications of using interactive teaching tools for the line of best fit can help students make connections between the shift in slope and the location of the best fit line [@nagle2017slope].
-@casey2015teachers conducted similar task based interviews wish pre-service mathematics teachers in order to understand the implications regarding teacher preparation for the teaching of this topic. 
+@casey2015teachers conducted similar task based interviews wish pre-service mathematics teachers in order to understand the effects regarding teacher preparation for the teaching of this topic. 
 This demonstrates the use of the 'You Draw It' interactive tool for training purposes both within a classroom setting and for those in the workforce.

### 2022-11-14 2022-sdss-you-draw-it-manuscript.tex
edited discussion

@@
 \texttt{r2d3}}
-\author{}
-\date{}
+% \author{}
+% \date{}
 
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[boxrule=0pt, sharp corners, borderline west={3pt}{0pt}{shadecolor}, breakable, enhanced, frame hidden, interior hidden]}{\end{tcolorbox}}\fi
 
@@
 Initial studies in the 20th century explored the use of fitting lines by
-eye through a set of points (Finney 1951; Mosteller et al. 1981; Unwin
-and Wills 1988). Common methods of fitting trends by eye involved
+eye through a set of points (Finney 1951; Mosteller et al. 1981; A. R.
+Unwin and Wills 1988). Common methods of fitting trends by eye involved
 maneuvering a string, black thread, or ruler until the fit was suitable,
@@
 The `You Draw It' graphical task presented in this work is an example of
-an interactive graphic. While any communication between user and device
-can be considered an interaction, interactive graphics should be defined
-as a user-driven direct manipulation of plots and plotting elements with
-an immediate reaction (Unwin 1999, Becker et al.~1987, Eick \& Wills
-1995). Here we discuss the use of `You Draw It' as a tool for
+an interactive graphic; an interactive graphic is defined as a
+user-driven direct manipulation of the plot followed by an immediate
+reaction (A. Unwin 1999; Eick and Wills 1995; Becker, Cleveland, and
+Wilks 1987). Here we discuss the use of `You Draw It' as a tool for
 interactive reading, interactive testing, and interactive teaching as a
@@
 the use of the `You Draw It' method as a tool for interactive testing of
-graphics. We collect data of the user drawn lines and compare to
+graphics. We collected and compared data from user drawn lines to
 statistical results in order to understand perceptual biases. While this
@@
 The line of best fit is introduced in secondary schools as a student's
-first encounter with the concept of statistical association (Casey and
-Nagle 2016). Casey (2015) and Casey (2016) conducted task based
-interviews to examine student's conceptions of the line of best fit
-prior to introducing the topic. The implications of using interactive
-teaching tools for the line of best fit can help students make
-connections between the shift in slope and the location of the best fit
-line (Nagle, Casey, and Moore-Russo 2017). Casey and Wasserman (2015)
-conducted similar task based interviews wish pre-service mathematics
-teachers in order to understand the implications regarding teacher
-preparation for the teaching of this topic. This demonstrates the use of
-the `You Draw It' interactive tool for training purposes both within a
-classroom setting and for those in the workforce.
+first encounter with the concept of statistical association between
+multiple variables (Casey and Nagle 2016). Casey (2015) and Casey (2016)
+conducted task based interviews to examine student's conceptions of the
+line of best fit prior to introducing the topic. The implications of
+using interactive teaching tools for the line of best fit can help
+students make connections between the shift in slope and the location of
+the best fit line (Nagle, Casey, and Moore-Russo 2017). Casey and
+Wasserman (2015) conducted similar task based interviews wish
+pre-service mathematics teachers in order to understand the effects
+regarding teacher preparation for the teaching of this topic. This
+demonstrates the use of the `You Draw It' interactive tool for training
+purposes both within a classroom setting and for those in the workforce.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-becker1987dynamic}{}}%
+Becker, Richard A, William S Cleveland, and Allan R Wilks. 1987.
+{``Dynamic Graphics for Data Analysis.''} \emph{Statistical Science} 2
+(4): 355--83.
+
 \leavevmode\vadjust pre{\hypertarget{ref-bostock2011d3}{}}%
@@
 Deming, William Edwards. 1943. {``Statistical Adjustment of Data.''}
+
+\leavevmode\vadjust pre{\hypertarget{ref-eick1995high}{}}%
+Eick, Stephen G, and Graham J Wills. 1995. {``High Interaction
+Graphics.''} \emph{European Journal of Operational Research} 81 (3):
+445--59.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-unwin1999requirements}{}}%
+Unwin, Antony. 1999. {``Requirements for Interactive Graphics Software
+for Exploratory Data Analysis.''} \emph{Computational Statistics} 14
+(1): 7--22.
+
 \leavevmode\vadjust pre{\hypertarget{ref-unwin1988eyeballing}{}}%

### 2022-11-14 revisions/response-to-reviewers.qmd
revisions 1 done

@@
 
-All responses to reviewers' comments are written in \response{blue}.
+All responses to reviewers' comments are written in \response{blue}. A `diffs1.pdf` file has also been included to show changes.
 
@@
 We saved this at the highest resoultion the tool allowed as a png file.}
+
+<!-- latexdiff C:/Users/erobin17/Documents/publications/2022-sdss-you-draw-it-manuscript/original-2022-sdss-you-draw-it-manuscript.tex C:/Users/erobin17/Documents/publications/2022-sdss-you-draw-it-manuscript/2022-sdss-you-draw-it-manuscript.tex > C:/Users/erobin17/Documents/publications/2022-sdss-you-draw-it-manuscript/diffs1.tex -->

### 2022-11-14 2022-sdss-you-draw-it-manuscript.qmd
revisions 1 done

@@
 
-The 'You Draw It' graphical task presented in this work is an example of an interactive graphic; an interactive graphic is defined as a user-driven direct manipulation of the plot followed by an immediate reaction [@unwin1999requirements; @eick1995high; @becker1987dynamic].
+The 'You Draw It' graphical task presented in this work is an example of an interactive graphic.
+<!-- an interactive graphic is defined as a user-driven direct manipulation of the plot followed by an immediate reaction [@unwin1999requirements; @eick1995high; @becker1987dynamic]. -->
 Here we discuss the use of 'You Draw It' as a tool for interactive reading, interactive testing, and interactive teaching as a way to increase user experience and user engagement with data.
@@
 The implications of using interactive teaching tools for the line of best fit can help students make connections between the shift in slope and the location of the best fit line [@nagle2017slope].
-@casey2015teachers conducted similar task based interviews wish pre-service mathematics teachers in order to understand the effects regarding teacher preparation for the teaching of this topic. 
+@casey2015teachers conducted similar task based interviews with pre-service mathematics teachers in order to understand the effects regarding teacher preparation for the teaching of this topic. 
 This demonstrates the use of the 'You Draw It' interactive tool for training purposes both within a classroom setting and for those in the workforce.

### 2022-11-14 2022-sdss-you-draw-it-manuscript.tex
submissions

--- 2022-11-12 2022-sdss-you-draw-it-manuscript.tex
@@
 \end{abstract}
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[breakable, enhanced, interior hidden, sharp corners, borderline west={3pt}{0pt}{shadecolor}, frame hidden, boxrule=0pt]}{\end{tcolorbox}}\fi
+\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[interior hidden, boxrule=0pt, frame hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, enhanced, breakable]}{\end{tcolorbox}}\fi
 
@@
 The `You Draw It' graphical task presented in this work is an example of
-an interactive graphic. While any communication between user and device
-can be considered an interaction, interactive graphics should be defined
-as a user-driven direct manipulation of plots and plotting elements with
-an immediate reaction (Unwin 1999, Becker et al.~1987, Eick \& Wills
-1995). Here we discuss the use of `You Draw It' as a tool for
-interactive reading, interactive testing, and interactive teaching as a
-way to increase user experience and user engagement with data. The New
-York Times first introduced the `You Draw It' feature as a tool for
-interactive reading. The inclusion of interactive graphics in news and
-media have been used to inform readers in efforts to reduce
+an interactive graphic. Here we discuss the use of `You Draw It' as a
+tool for interactive reading, interactive testing, and interactive
+teaching as a way to increase user experience and user engagement with
+data. The New York Times first introduced the `You Draw It' feature as a
+tool for interactive reading. The inclusion of interactive graphics in
+news and media have been used to inform readers in efforts to reduce
 misconceptions and aid in storytelling (Geidner et al. 2015;
@@
 The line of best fit is introduced in secondary schools as a student's
-first encounter with the concept of statistical association (Casey and
-Nagle 2016). Casey (2015) and Casey (2016) conducted task based
-interviews to examine student's conceptions of the line of best fit
-prior to introducing the topic. The implications of using interactive
-teaching tools for the line of best fit can help students make
-connections between the shift in slope and the location of the best fit
-line (Nagle, Casey, and Moore-Russo 2017). Casey and Wasserman (2015)
-conducted similar task based interviews wish pre-service mathematics
-teachers in order to understand the implications regarding teacher
-preparation for the teaching of this topic. This demonstrates the use of
-the `You Draw It' interactive tool for training purposes both within a
-classroom setting and for those in the workforce.
+first encounter with the concept of statistical association between
+multiple variables (Casey and Nagle 2016). Casey (2015) and Casey (2016)
+conducted task based interviews to examine student's conceptions of the
+line of best fit prior to introducing the topic. The implications of
+using interactive teaching tools for the line of best fit can help
+students make connections between the shift in slope and the location of
+the best fit line (Nagle, Casey, and Moore-Russo 2017). Casey and
+Wasserman (2015) conducted similar task based interviews with
+pre-service mathematics teachers in order to understand the effects
+regarding teacher preparation for the teaching of this topic. This
+demonstrates the use of the `You Draw It' interactive tool for training
+purposes both within a classroom setting and for those in the workforce.
 
